---
format: gfm
execute:
  echo: true
  eval: false
---

# irishbuoys

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The `irishbuoys` package provides tools to download, process, and analyze data from the Irish Weather Buoy Network. It includes functions for accessing real-time and historical data via the Marine Institute's ERDDAP server, storing data in DuckDB for efficient querying, and building predictive models for wave height and weather conditions.

## Installation

### Standard R Installation

You can install the development version of irishbuoys from GitHub:

```r
# install.packages("remotes")
remotes::install_github("johngavin/irishbuoys")
```

### Nix Environment Installation (Recommended)

For a reproducible development environment using Nix:

```bash
# Clone the repository
git clone https://github.com/johngavin/irishbuoys.git
cd irishbuoys

# Enter Nix shell (first time may take a while, subsequent entries are fast)
./default.sh
```

The `default.sh` script automatically:
- Generates `default.nix` from DESCRIPTION if needed (via `default.R`)
- Creates a GC root to prevent garbage collection
- Enters pure mode for reproducibility

### Pure Mode

This project enforces Nix `--pure` mode to guarantee:

- **Reproducibility**: Only Nix-provided tools are available
- **Security**: No accidental use of system tools with different versions
- **Consistency**: Same environment locally and in CI

**Entering the environment:**

```bash
# Recommended: Use default.sh (pure mode enforced automatically)
./default.sh
# Output shows: 🔒 SECURITY: Running in --pure mode

# Or manually with pure flag:
nix-shell --pure default.nix
```

**Verifying pure mode:**

```bash
# Check IN_NIX_SHELL (should be "pure", not "impure")
echo $IN_NIX_SHELL
# Expected: pure

# Check tools are from Nix store
which R
# Expected: /nix/store/...

which git
# Expected: /nix/store/...

# Verify R version
R --version
```

**Passing additional environment variables:**

```bash
# If you need to pass secrets/tokens through pure mode:
nix-shell --pure --keep GITHUB_TOKEN --keep MY_API_KEY default.nix
```

### Using with rix

To integrate this package into your own Nix environment:

```r
library(rix)

rix(
  r_ver = "4.5.2",  # Must match project's R version
  r_pkgs = c(
    # Core dependencies (from Imports)
    "arrow", "cli", "dbplyr", "DBI", "dplyr", "duckdb",
    "glue", "httr2", "jsonlite", "lubridate", "pointblank",
    "purrr", "rlang", "tibble",
    # Visualization (from Suggests)
    "plotly", "ggplot2", "dygraphs", "DT"
  ),
  git_pkgs = list(
    list(
      package_name = "irishbuoys",
      repo_url = "https://github.com/johngavin/irishbuoys",
      commit = "main"  # Use specific SHA for reproducibility
    )
  ),
  ide = "other",
  project_path = "."
)
```

### Cachix Binary Cache (Faster Builds)
::: {.panel-tabset}

#### For Other Machines

Use the pre-built R packages from `rstats-on-nix` Cachix cache for much faster builds:

```bash
# Install cachix (one-time setup)
nix-shell -p cachix --run "cachix use rstats-on-nix"

# Now nix-shell will download pre-built packages instead of compiling
cd irishbuoys
./default.sh  # Much faster with cache!
```

#### Two-Tier Caching Strategy

This project uses a two-tier Cachix strategy:

| Priority | Cache | Contains |
|----------|-------|----------|
| 1st | `rstats-on-nix` | All standard R packages (public, pre-built) |
| 2nd | `johngavin` | Project-specific custom packages only |

**Important**: Standard R packages (dplyr, targets, etc.) are ALL available from `rstats-on-nix`. The `johngavin` cache is only for custom packages not in rstats-on-nix.

For irishbuoys:
- All dependencies come from `rstats-on-nix`
- The irishbuoys package itself is loaded via `pkgload::load_all()` (development mode)
- Nothing needs to be pushed to `johngavin` cache

#### GitHub Actions CI

CI workflows automatically use both caches:

```yaml
# In .github/workflows/r-cmd-check.yaml
- uses: cachix/cachix-action@v15
  with:
    name: rstats-on-nix  # Public cache FIRST

- uses: cachix/cachix-action@v15
  with:
    name: johngavin      # Project cache SECOND
    authToken: '${{ secrets.CACHIX_AUTH_TOKEN }}'
    skipPush: true       # Don't push during checks
```

:::

## Documentation

Comprehensive vignettes are available on the [package website](https://johngavin.github.io/irishbuoys/):

### [Static Dashboard](https://johngavin.github.io/irishbuoys/articles/dashboard_static.html)
Interactive visualization of Irish Weather Buoy Network data.
- **Real-time station data**: Max Wave, Signif Wave, and Wind Speed by station with tabbed navigation
- **Time series analysis**: Hourly measurements with interactive plotly charts
- **Extreme event detection**: Rogue wave identification where Max Wave > 2× Signif Wave

### [Wave Analysis](https://johngavin.github.io/irishbuoys/articles/wave_analysis.html)
Scientific analysis of wave patterns and extreme value statistics.
- **Key metrics dashboard**: Station statistics, date coverage, and data quality summary
- **Extreme value modeling**: GEV/GPD fits for return period estimation
- **Seasonal patterns**: STL decomposition and annual trend analysis

### [Shinylive Dashboard](https://johngavin.github.io/irishbuoys/articles/dashboard_shinylive.html)
Client-side interactive dashboard powered by WebR/Shinylive.
- **Browser-based R**: No server required - runs entirely in your browser
- **Dynamic filtering**: Station and date range selection with live updates
- **Offline capable**: Works without internet after initial load

### [Debug](https://johngavin.github.io/irishbuoys/articles/debug.html)
Pipeline diagnostics and data validation reports.
- **pointblank validation**: HTML reports showing data quality checks with pass/fail status
- **Pipeline status**: targets progress tracking with outdated target detection
- **Session info**: R environment details for reproducibility debugging

### Data Quality (pointblank)

Interactive validation reports generated by the [pointblank](https://rstudio.github.io/pointblank/) package:

- **[Analysis Data Validation](https://johngavin.github.io/irishbuoys/articles/validation_analysis_data.html)**: Comprehensive checks on buoy data including column existence, physical bounds (wave height 0-30m, wind speed 0-100 m/s), and station ID validation
- **[Rogue Wave Validation](https://johngavin.github.io/irishbuoys/articles/validation_rogue_events.html)**: Event-level validation ensuring detected rogue waves meet the Hmax > 2 x Hs criterion
- **Interactive reports**: Each report shows pass/fail status with detailed interrogation results

## Quick Start

### Download Recent Data

```r
library(irishbuoys)

# Download last 7 days of data
# Show structure of first 3 rows for efficiency
data <- download_buoy_data(
  start_date = Sys.Date() - 7,
  end_date = Sys.Date()
)
str(head(data, 3))

# Get data for specific station
m3_data <- download_buoy_data(
  stations = "M3",
  start_date = "2024-01-01"
)
dim(m3_data)

# Get earliest available data (buoy network started 2001-02-05)
# Show structure of first 3 rows for efficiency
waves <- download_buoy_data(
  start_date = "2001-02-05",
  end_date = "2001-02-06"
)
str(head(waves, 3))
```

### Initialize and Query Database

```r
# Initialize database with historical data (chunk_days=365 for faster downloads)
initialize_database(
  start_date = "2024-01-01",  # Default: recent data for quick start
  end_date = Sys.Date(),
  chunk_days = 365  # Download in 1-year chunks for efficiency
)

# Check database statistics immediately after initialization
stats <- get_database_stats()

# Connect to database
con <- connect_duckdb()

# Check QC flag distribution by station FIRST
# qc_flag: 0=unknown, 1=good, 9=missing
qc_tally <- tbl(con, "buoy_data") |>
  group_by(station_id, qc_flag) |>
  summarise(n = n(), .groups = "drop") |>
  collect() |>
  tidyr::pivot_wider(names_from = qc_flag, values_from = n, names_prefix = "qc_")
print(qc_tally)

# Query wave data (qc_filter=FALSE returns all data; TRUE filters for qc_flag==1)
wave_data <- query_buoy_data(
  con,
  stations = c("M3", "M4"),
  variables = c("time", "station_id", "wave_height", "wave_period"),
  start_date = "2024-01-01",
  qc_filter = FALSE  # Set TRUE to filter for qc_flag==1 only
)

# Custom SQL query: Find top 10 most extreme rogue waves
# Ordered by hmax (highest individual wave) because "extreme" = largest waves
extreme_waves <- query_buoy_data(
  con,
  sql_query = "
    SELECT station_id, time, wave_height, hmax
    FROM buoy_data
    WHERE hmax > 2 * wave_height
      AND wave_height > 0
      AND qc_flag = 1
    ORDER BY hmax DESC
    LIMIT 10
  "
)
```

### Tidyverse Alternative (duckplyr)

The same query using `dplyr` verbs with `duckplyr` backend:
```r
# Tidyverse alternative using duckplyr
# Same query as SQL above, ordered by hmax (highest waves first)
library(dplyr)
library(duckplyr)

extreme_waves_tidy <- tbl(con, "buoy_data") |>
  filter(
    hmax > 2 * wave_height,
    wave_height > 0,
    qc_flag == 1
  ) |>
  select(station_id, time, wave_height, hmax) |>
  arrange(desc(hmax)) |>
  head(10) |>
  collect()
```

**Why duckplyr?**
- Familiar tidyverse syntax
- Lazy evaluation - query runs only on `collect()`
- Automatic SQL translation for performance
- Works with any DBI connection

```r
# Don't forget to disconnect
DBI::dbDisconnect(con)
```

### Incremental Updates

```r
# Perform incremental update (for scheduled jobs)
result <- incremental_update()
print(result$summary)
```

### Data Dictionary

```r
# Get complete data dictionary (returns tibble)
dict <- get_data_dictionary()
print(dict)

# Get detailed documentation for specific variable
(wave_docs <- get_variable_docs("WaveHeight"))

# Merge dictionary with database column info
# Useful for creating documentation or understanding data
library(dplyr)
db_cols <- tibble(
  variable = c("wave_height", "hmax", "wind_speed", "gust"),
  db_column = c("wave_height", "hmax", "wind_speed", "gust")
)
dict |>
  filter(tolower(variable) %in% db_cols$variable |
         variable %in% c("WaveHeight", "Hmax", "WindSpeed", "Gust")) |>
  select(variable, units, description)
```

## Data Source

Data is sourced from the [Marine Institute's ERDDAP server](https://erddap.marine.ie/erddap/tabledap/IWBNetwork.html), which provides real-time and historical measurements from the Irish Weather Buoy Network.

### Available Stations

- **M2**: Southwest of Ireland
- **M3**: Southwest of Ireland
- **M4**: Southeast of Ireland
- **M5**: West of Ireland
- **M6**: Northwest of Ireland
- **M1**: Historical data (decommissioned)
- **FS1**: Historical data

### Measured Parameters

- **Meteorological**: Air temperature, pressure, humidity, wind speed/direction
- **Oceanographic**: Wave height/period/direction, sea temperature, salinity
- **Quality**: QC flags for data validation

## Project Structure

```{r}
#| echo: false
#| eval: true
fs::dir_tree(recurse = 2, regexp = "^(_targets|docs)$|^(_targets|docs)/", invert = TRUE)
```

*Note: `_targets/` (pipeline cache) and `docs/` (generated site) excluded for clarity.*

## Key Features

1. **Efficient Data Storage**: Uses DuckDB for fast querying of large datasets
2. **Incremental Updates**: Smart updating to only download new data
3. **Quality Control**: Built-in filtering for data quality
4. **Rogue Wave Detection**: Identify extreme wave events (Hmax > 2 × Hs)
5. **Comprehensive Documentation**: Full data dictionary with scientific definitions

## Use Cases

- Marine safety and operations planning
- Climate and oceanographic research
- Extreme event analysis
- Wave energy resource assessment
- Weather forecast validation

## Contributing

Contributions are welcome! Please feel free to submit a Pull Request.

## License

This package is licensed under the MIT License. See LICENSE file for details.

## Acknowledgments

Data provided by the Marine Institute Ireland in collaboration with Met Éireann and the UK Met Office.

## Sources

- [Marine Institute ERDDAP Server](https://erddap.marine.ie/erddap/tabledap/IWBNetwork.html)
- [Irish Weather Buoy Network on data.gov.ie](https://data.gov.ie/dataset/weather-buoy-network)

---

```{r}
#| echo: false
#| eval: true
cat("*Last updated:", format(Sys.time(), "%Y-%m-%d %H:%M UTC"), "*")
```