---
title: "Static Data API"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# All data loaded from pre-computed targets via tar_load().
# MANDATORY: No inline computation. Every chunk is ONE expression.
library(targets)

store_path <- if (dir.exists("_targets")) "_targets" else "../_targets"
targets::tar_config_set(store = store_path)

safe_tar_read <- function(name) {
  tryCatch(
    targets::tar_read_raw(name),
    error = function(e) {
      htmltools::div(
        class = "alert alert-warning",
        htmltools::h4("Data Not Available"),
        htmltools::p(
          "The ", htmltools::code(name), " target has not been built."
        ),
        htmltools::p("Run ", htmltools::code("targets::tar_make()"), " to generate.")
      )
    }
  )
}

tryCatch({
  tar_load(api_index)
  tar_load(api_vignette_endpoints_dt)
  tar_load(api_vignette_example_response)
  tar_load(save_api_files)
  tar_load(api_seasonal)
  tar_load(api_correlations)
  tar_load(code_api_curl_latest)
  tar_load(code_api_r_jsonlite)
  tar_load(code_api_python_requests)
  targets_loaded <- TRUE
}, error = function(e) {
  targets_loaded <<- FALSE
})
```

## Overview {#overview}

The irishbuoys package provides a **static JSON API** served via GitHub Pages.
All endpoints are pre-computed by the targets pipeline and updated weekly
(Sundays 2 AM UTC).

**Base URL:** `https://johngavin.github.io/irishbuoys/api/v1/`

No authentication required. All responses are JSON. Data is updated weekly
from the Marine Institute's ERDDAP server.

## Available Endpoints {#endpoints}

```{r endpoints-table}
safe_tar_read("api_vignette_endpoints_dt")
```

## Example Response {#example-response}

A sample response from `latest.json` (one observation per station):

```{r example-json}
#| results: asis
safe_tar_read("api_vignette_example_response")
```

## Seasonal Endpoint {#seasonal-endpoint}

The `seasonal.json` endpoint provides monthly and seasonal (DJF/MAM/JJA/SON)
mean statistics for wave height and wind speed, plus annual trend analysis
with linear trend per decade and significance (p-value).

**Fields:** `wave.monthly`, `wave.seasonal`, `wind.monthly`, `wind.seasonal`,
`annual_trends.wave`, `annual_trends.wind`

```{r seasonal-example}
#| results: asis
seasonal <- safe_tar_read("api_seasonal")
if (is.list(seasonal) && "wave" %in% names(seasonal)) {
  cat("```json\n")
  cat(jsonlite::toJSON(
    list(wave = list(monthly = head(seasonal$wave$monthly, 3))),
    pretty = TRUE, auto_unbox = TRUE
  ))
  cat("\n```\n")
}
```

## Correlations Endpoint {#correlations-endpoint}

The `correlations.json` endpoint provides overall wind-wave and wave-Hmax
correlations plus inter-station pair correlations (cross-correlation analysis)
for wave height, wind speed, and maximum wave height.

**Fields:** `overall.wind_wave`, `overall.wave_hmax`, `station_pairs.wave_height`,
`station_pairs.wind_speed`, `station_pairs.hmax`

```{r correlations-example}
#| results: asis
corr <- safe_tar_read("api_correlations")
if (is.list(corr) && "overall" %in% names(corr)) {
  cat("```json\n")
  cat(jsonlite::toJSON(
    list(overall = corr$overall),
    pretty = TRUE, auto_unbox = TRUE
  ))
  cat("\n```\n")
}
```

## Usage Examples {#usage-examples}

### curl {#curl-example}

```{r curl-code}
#| results: asis
code <- safe_tar_read("code_api_curl_latest")
if (is.character(code)) {
  cat("```bash\n", paste(code, collapse = "\n"), "\n```\n")
}
```

### R (jsonlite) {#r-example}

```{r r-code}
#| results: asis
code <- safe_tar_read("code_api_r_jsonlite")
if (is.character(code)) {
  cat("```r\n", paste(code, collapse = "\n"), "\n```\n")
}
```

### Python (requests) {#python-example}

```{r python-code}
#| results: asis
code <- safe_tar_read("code_api_python_requests")
if (is.character(code)) {
  cat("```python\n", paste(code, collapse = "\n"), "\n```\n")
}
```

## Bulk Data Access {#bulk-data}

For bulk analysis, download the full dataset as Parquet:

```bash
curl -O https://johngavin.github.io/irishbuoys/vignettes/data/buoy_data.parquet
```

Read in R with `arrow::read_parquet()` or in Python with `pandas.read_parquet()`.

## API File Summary {#file-summary}

```{r file-sizes}
safe_tar_read("save_api_files")
```

## Update Schedule {#update-schedule}

- **Frequency:** Weekly (Sundays 2 AM UTC)
- **Source:** Marine Institute ERDDAP server
- **Pipeline:** `targets::tar_make()` with incremental DuckDB updates
- **CI:** [weekly-update.yml](https://github.com/JohnGavin/irishbuoys/actions/workflows/weekly-update.yml)

Data freshness depends on the ERDDAP server. The pipeline downloads the
latest 72 hours of observations each run to catch any late-arriving data.

---

::: {.callout-note appearance="minimal"}
**Disclaimer:** `irishbuoys` is an independent open-source project ([website](https://johngavin.github.io/irishbuoys/), [GitHub](https://github.com/JohnGavin/irishbuoys)). It is not affiliated with, endorsed by, or part of the [Marine Institute](https://www.marine.ie/) or the Irish Weather Buoy Network. For official marine weather warnings, consult [Met Eireann](https://www.met.ie/warnings).
:::
