---
title: "Static Data API"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# All data loaded from pre-computed targets via tar_load().
# MANDATORY: No inline computation. Every chunk is ONE expression.
library(targets)

store_path <- if (dir.exists("_targets")) "_targets" else "../_targets"
targets::tar_config_set(store = store_path)

safe_tar_read <- function(name) {
  tryCatch(
    targets::tar_read_raw(name),
    error = function(e) {
      htmltools::div(
        class = "alert alert-warning",
        htmltools::h4("Data Not Available"),
        htmltools::p(
          "The ", htmltools::code(name), " target has not been built."
        ),
        htmltools::p("Run ", htmltools::code("targets::tar_make()"), " to generate.")
      )
    }
  )
}

tryCatch({
  tar_load(api_index)
  tar_load(api_vignette_endpoints_dt)
  tar_load(api_vignette_example_response)
  tar_load(api_vignette_stations_dt)
  tar_load(api_vignette_stats_dt)
  tar_load(api_vignette_return_levels_dt)
  tar_load(api_vignette_rogue_waves_dt)
  tar_load(api_vignette_data_dict_dt)
  tar_load(save_api_files)
  tar_load(api_seasonal)
  tar_load(api_correlations)
  tar_load(api_stats)
  tar_load(api_return_levels)
  tar_load(api_rogue_waves)
  tar_load(api_data_dictionary)
  tar_load(code_api_curl_latest)
  tar_load(code_api_r_jsonlite)
  tar_load(code_api_python_requests)
  targets_loaded <- TRUE
}, error = function(e) {
  targets_loaded <<- FALSE
})
```

## Overview {#overview}

The irishbuoys package provides a **static JSON API** served via GitHub Pages.
All endpoints are pre-computed by the targets pipeline and updated weekly
(Sundays 2 AM UTC).

**Base URL:** `https://johngavin.github.io/irishbuoys/api/v1/`

No authentication required. All responses are JSON. Data is updated weekly
from the Marine Institute's ERDDAP server.

## Available Endpoints {#endpoints}

```{r endpoints-table}
safe_tar_read("api_vignette_endpoints_dt")
```

## Stations {#stations}

The `stations.json` endpoint provides metadata for each buoy in the Irish
Weather Buoy Network, including station ID, name, latitude, and longitude.

```{r stations-table}
safe_tar_read("api_vignette_stations_dt")
```

## Latest Observations {#latest}

The `latest.json` endpoint returns the most recent observation per station.
A sample response (one station):

```{r example-json}
#| results: asis
safe_tar_read("api_vignette_example_response")
```

## Summary Statistics {#stats}

The `stats.json` endpoint provides per-station summary statistics including
record counts, mean/max wave height, and mean/max wind speed.

```{r stats-table}
safe_tar_read("api_vignette_stats_dt")
```

```{r stats-overall}
#| results: asis
stats <- safe_tar_read("api_stats")
if (is.list(stats) && "overall" %in% names(stats)) {
  cat("**Overall summary:**\n\n")
  cat(sprintf("- Total records: %s\n", format(stats$overall$total_records, big.mark = ",")))
  cat(sprintf("- Date range: %s to %s\n", stats$overall$date_range[1], stats$overall$date_range[2]))
  cat(sprintf("- Stations: %s\n", stats$overall$stations))
  cat(sprintf("- Wind-wave correlation: %s\n", stats$overall$wind_wave_correlation))
  cat(sprintf("- Wave-Hmax correlation: %s\n", stats$overall$wave_hmax_correlation))
}
```

## Seasonal Patterns {#seasonal}

The `seasonal.json` endpoint provides monthly and seasonal (DJF/MAM/JJA/SON)
mean statistics for wave height and wind speed, plus annual trend analysis
with linear trend per decade and significance (p-value).

**Fields:** `wave.monthly`, `wave.seasonal`, `wind.monthly`, `wind.seasonal`,
`annual_trends.wave`, `annual_trends.wind`

### Wave data (sample)

```{r seasonal-wave}
#| results: asis
seasonal <- safe_tar_read("api_seasonal")
if (is.list(seasonal) && "wave" %in% names(seasonal)) {
  cat("```json\n")
  cat(jsonlite::toJSON(
    list(wave = list(monthly = head(seasonal$wave$monthly, 3))),
    pretty = TRUE, auto_unbox = TRUE
  ))
  cat("\n```\n")
}
```

### Wind data (sample)

```{r seasonal-wind}
#| results: asis
seasonal <- safe_tar_read("api_seasonal")
if (is.list(seasonal) && "wind" %in% names(seasonal)) {
  cat("```json\n")
  cat(jsonlite::toJSON(
    list(wind = list(monthly = head(seasonal$wind$monthly, 3))),
    pretty = TRUE, auto_unbox = TRUE
  ))
  cat("\n```\n")
}
```

### Annual trends

```{r seasonal-trends}
#| results: asis
seasonal <- safe_tar_read("api_seasonal")
if (is.list(seasonal) && "annual_trends" %in% names(seasonal)) {
  cat("```json\n")
  cat(jsonlite::toJSON(
    list(annual_trends = seasonal$annual_trends),
    pretty = TRUE, auto_unbox = TRUE
  ))
  cat("\n```\n")
}
```

## Correlations {#correlations}

The `correlations.json` endpoint provides overall wind-wave and wave-Hmax
correlations plus inter-station pair correlations (cross-correlation analysis)
for wave height, wind speed, and maximum wave height.

**Fields:** `overall.wind_wave`, `overall.wave_hmax`, `station_pairs.wave_height`,
`station_pairs.wind_speed`, `station_pairs.hmax`

### Overall correlations

```{r correlations-overall}
#| results: asis
corr <- safe_tar_read("api_correlations")
if (is.list(corr) && "overall" %in% names(corr)) {
  cat("```json\n")
  cat(jsonlite::toJSON(
    list(overall = corr$overall),
    pretty = TRUE, auto_unbox = TRUE
  ))
  cat("\n```\n")
}
```

### Station pair sample (wave height)

```{r correlations-pairs}
#| results: asis
corr <- safe_tar_read("api_correlations")
if (is.list(corr) && "station_pairs" %in% names(corr)) {
  pairs <- corr$station_pairs$wave_height
  if (is.data.frame(pairs)) {
    sample_pairs <- utils::head(pairs, 5)
  } else if (is.list(pairs)) {
    sample_pairs <- utils::head(pairs, 5)
  } else {
    sample_pairs <- pairs
  }
  cat("```json\n")
  cat(jsonlite::toJSON(
    list(station_pairs = list(wave_height = sample_pairs)),
    pretty = TRUE, auto_unbox = TRUE
  ))
  cat("\n```\n")
}
```

## Return Levels {#return-levels}

The `return-levels.json` endpoint provides Generalised Pareto Distribution (GPD)
return level estimates for extreme wave heights. The 95th percentile threshold
is used, with return periods of 1, 5, and 10 years.

**Method:** GPD | **Threshold:** 95th percentile | **Return periods:** 1, 5, 10 years

```{r return-levels-table}
safe_tar_read("api_vignette_return_levels_dt")
```

## Rogue Waves {#rogue-waves}

The `rogue-waves.json` endpoint lists detected rogue wave events where the
maximum wave height (Hmax) exceeds twice the significant wave height (Hs).
Events are sorted by the Hmax/Hs ratio (most extreme first).

```{r rogue-waves-table}
safe_tar_read("api_vignette_rogue_waves_dt")
```

```{r rogue-waves-count}
#| results: asis
rw <- safe_tar_read("api_rogue_waves")
if (is.list(rw) && "n_events" %in% names(rw)) {
  cat(sprintf("\n**Total rogue wave events detected:** %s\n", format(rw$n_events, big.mark = ",")))
}
```

## Data Dictionary {#data-dictionary}

The `data-dictionary.json` endpoint defines all variables in the dataset,
including names, descriptions, units, and data types.

```{r data-dict-table}
safe_tar_read("api_vignette_data_dict_dt")
```

## Data Sources {#data-sources}

All data originates from the Marine Institute's ERDDAP server:

- **ERDDAP base URL:** <https://erddap.marine.ie/erddap/tabledap/IWBNetwork>
- **Dataset ID:** `IWBNetwork`
- **Direct data query (HTML):** <https://erddap.marine.ie/erddap/tabledap/IWBNetwork.html>
- **Metadata (info page):** <https://erddap.marine.ie/erddap/info/IWBNetwork/index.html>

Example ERDDAP queries:

```bash
# Latest 100 rows as CSV
curl "https://erddap.marine.ie/erddap/tabledap/IWBNetwork.csv?&orderByLimit(%22100%22)"

# Station metadata as JSON
curl "https://erddap.marine.ie/erddap/tabledap/IWBNetwork.json?station_id,longitude,latitude&distinct()"
```

## Usage Examples {#usage-examples}

### curl {#curl-example}

```{r curl-code}
#| results: asis
code <- safe_tar_read("code_api_curl_latest")
if (is.character(code)) {
  cat("```bash\n", paste(code, collapse = "\n"), "\n```\n")
}
```

### R (jsonlite) {#r-example}

```{r r-code}
#| results: asis
code <- safe_tar_read("code_api_r_jsonlite")
if (is.character(code)) {
  cat("```r\n", paste(code, collapse = "\n"), "\n```\n")
}
```

### Fetching stats and seasonal data (R)

```r
library(jsonlite)

# Per-station statistics
stats <- fromJSON("https://johngavin.github.io/irishbuoys/api/v1/stats.json")
stats$overall

# Seasonal patterns
seasonal <- fromJSON("https://johngavin.github.io/irishbuoys/api/v1/seasonal.json")
head(seasonal$wave$monthly)

# Return levels
rl <- fromJSON("https://johngavin.github.io/irishbuoys/api/v1/return-levels.json")
str(rl$by_station, max.level = 2)
```

### Python (requests) {#python-example}

```{r python-code}
#| results: asis
code <- safe_tar_read("code_api_python_requests")
if (is.character(code)) {
  cat("```python\n", paste(code, collapse = "\n"), "\n```\n")
}
```

### Fetching stats and seasonal data (Python)

```python
import requests
import pandas as pd

base = "https://johngavin.github.io/irishbuoys/api/v1/"

# Per-station statistics
stats = requests.get(base + "stats.json").json()
print(stats["overall"])

# Seasonal wave patterns
seasonal = requests.get(base + "seasonal.json").json()
df_monthly = pd.DataFrame(seasonal["wave"]["monthly"])
print(df_monthly.head())
```

## Bulk Data Access {#bulk-data}

For bulk analysis, download the full dataset as Parquet:

```bash
curl -O https://johngavin.github.io/irishbuoys/vignettes/data/buoy_data.parquet
```

Read in R with `arrow::read_parquet()` or in Python with `pandas.read_parquet()`.

## API File Summary {#file-summary}

```{r file-sizes}
safe_tar_read("save_api_files")
```

## Live API {#live-api}

In addition to the static JSON files, irishbuoys provides a live REST API
powered by [plumber](https://www.rplumber.io/). The API serves the same
pre-computed JSON data via HTTP endpoints.

### Starting the API

```r
library(irishbuoys)
run_api(port = 8080)
# API at http://localhost:8080
# Swagger docs at http://localhost:8080/__docs__/
```

### Endpoints

| Endpoint | Description |
|----------|-------------|
| `GET /stations` | Station metadata and locations |
| `GET /rogue-waves` | Detected rogue wave events |
| `GET /return-levels` | Extreme value return level estimates |
| `GET /seasonal` | Monthly and seasonal statistics |

### Programmatic Access

```r
# Create router without starting server (useful for testing)
pr <- create_api_router()
```

The live API requires the `plumber` package (`install.packages("plumber")`).

## Update Schedule {#update-schedule}

- **Frequency:** Weekly (Sundays 2 AM UTC)
- **Source:** Marine Institute ERDDAP server
- **Pipeline:** `targets::tar_make()` with incremental DuckDB updates
- **CI:** [weekly-update.yml](https://github.com/JohnGavin/irishbuoys/actions/workflows/weekly-update.yml)

Data freshness depends on the ERDDAP server. The pipeline downloads the
latest 72 hours of observations each run to catch any late-arriving data.

---

::: {.callout-note appearance="minimal"}
**Disclaimer:** `irishbuoys` is an independent open-source project ([website](https://johngavin.github.io/irishbuoys/), [GitHub](https://github.com/JohnGavin/irishbuoys)). It is not affiliated with, endorsed by, or part of the [Marine Institute](https://www.marine.ie/) or the Irish Weather Buoy Network. For official marine weather warnings, consult [Met Eireann](https://www.met.ie/warnings).
:::
