---
title: "Irish Buoys Explorer (2019-2026)"
format:
  dashboard:
    orientation: rows
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(dygraphs)
library(xts)
library(tidyr)

# Load pre-computed data from targets pipeline
data_dir <- "../inst/extdata"
if (!dir.exists(data_dir)) {
  data_dir <- system.file("extdata", package = "irishbuoys")
}

buoy_data <- readRDS(file.path(data_dir, "dashboard_buoy_data.rds"))
dashboard_stats <- readRDS(file.path(data_dir, "dashboard_stats.rds"))
dashboard_ts <- readRDS(file.path(data_dir, "dashboard_timeseries.rds"))

# Get package version
pkg_version <- tryCatch(
  as.character(packageVersion("irishbuoys")),
  error = function(e) {
    desc_file <- "../DESCRIPTION"
    if (file.exists(desc_file)) {
      desc <- read.dcf(desc_file)
      desc[1, "Version"]
    } else "dev"
  }
)

# Get unique stations
stations <- sort(unique(buoy_data$station_id))

# Use HOURLY data directly - preserve full detail
# Only sample if dataset is extremely large (>200k rows) to keep HTML manageable
# Reason: User wants to see actual hourly measurements, not daily aggregates
set.seed(42)
if (nrow(buoy_data) > 200000) {
  # Sample for time series plots only if absolutely necessary
  buoy_hourly <- buoy_data |> slice_sample(n = 200000)
  n_hourly_display <- format(200000, big.mark = ",")
  sampling_note <- paste0(" (sampled to ", n_hourly_display, " observations for performance)")
} else {
  buoy_hourly <- buoy_data
  n_hourly_display <- format(nrow(buoy_data), big.mark = ",")
  sampling_note <- ""
}

# Calculate summary stats
valid_ww <- complete.cases(buoy_data$wind_speed, buoy_data$wave_height)
valid_wh <- complete.cases(buoy_data$wave_height, buoy_data$hmax)
cor_wind_wave <- round(cor(buoy_data$wind_speed[valid_ww], buoy_data$wave_height[valid_ww]), 3)
cor_wave_hmax <- round(cor(buoy_data$wave_height[valid_wh], buoy_data$hmax[valid_wh]), 3)
days_span <- round(as.numeric(difftime(max(buoy_data$time), min(buoy_data$time), units = "days")))
generated_time <- format(Sys.time(), "%Y-%m-%d %H:%M UTC")
date_caption <- paste0(format(min(buoy_data$time), "%Y-%m-%d"), " to ", format(max(buoy_data$time), "%Y-%m-%d"))
n_records <- format(nrow(buoy_data), big.mark = ",")

# Helper function to create station-specific data
get_station_data <- function(data, station) {
  if (station != "All Buoys") {
    data <- data |> filter(station_id == station)
  }
  data
}

# Helper function to create dygraph time series - full width
create_dygraph <- function(data, variable = "wave_height", title = "", by_station = TRUE) {
  if (by_station && length(unique(data$station_id)) > 1) {
    wide_data <- data |>
      select(time, station_id, all_of(variable)) |>
      tidyr::pivot_wider(names_from = station_id, values_from = all_of(variable))
    ts_data <- xts(wide_data[, -1], order.by = wide_data$time)
  } else {
    ts_data <- xts(data[[variable]], order.by = data$time)
    colnames(ts_data) <- variable
  }

  dygraph(ts_data, main = title, width = "100%") |>
    dyOptions(
      labelsUTC = TRUE,
      fillGraph = TRUE,
      fillAlpha = 0.1,
      drawGrid = TRUE,
      colors = RColorBrewer::brewer.pal(max(3, ncol(ts_data)), "Set1")[1:ncol(ts_data)]
    ) |>
    dyRangeSelector(height = 40) |>
    dyCrosshair(direction = "vertical") |>
    dyHighlight(highlightCircleSize = 5, highlightSeriesBackgroundAlpha = 0.2, hideOnMouseOut = FALSE) |>
    dyLegend(show = "always", width = 400)
}

# Helper function for DT tables with optimized timestamp display and frozen headers
create_dt <- function(data, caption = "", time_width = "140px") {
  time_cols <- which(names(data) %in% c("time", "Time", "timestamp", "Timestamp", "datetime", "Time (UTC)", "First", "Last"))

  dt <- datatable(
    data,
    caption = caption,
    filter = 'top',
    extensions = c('Buttons', 'FixedHeader'),
    options = list(
      pageLength = 15,
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel'),
      order = list(list(0, 'desc')),
      scrollX = TRUE,
      fixedHeader = TRUE,
      columnDefs = if (length(time_cols) > 0) {
        list(list(width = time_width, targets = time_cols - 1))
      } else NULL
    ),
    class = 'cell-border stripe hover compact'
  )

  if (length(time_cols) > 0) {
    dt <- dt %>%
      DT::formatStyle(columns = time_cols, fontSize = '10px', whiteSpace = 'nowrap')
  }

  dt
}

# Station information with coordinates and depths
station_info <- data.frame(
  Station = c("M2", "M3", "M4", "M5", "M6"),
  Location = c("Southwest Ireland", "Southwest Ireland", "Southeast Ireland", "West Ireland", "Northwest Ireland"),
  Latitude = c(51.22, 51.22, 51.69, 51.69, 53.07),
  Longitude = c(-9.99, -10.55, -6.70, -10.00, -15.93),
  `Water Depth (m)` = c(155, 155, 62, 70, 3000),
  `Distance from Shore (km)` = c(50, 65, 30, 55, 320),
  check.names = FALSE
)

# Pre-compute max values for captions
max_hmax <- round(max(buoy_data$hmax, na.rm = TRUE), 1)
max_signif <- round(max(buoy_data$wave_height, na.rm = TRUE), 1)
max_station_hmax <- buoy_data |> filter(hmax == max(hmax, na.rm = TRUE)) |> pull(station_id) |> unique() |> first()
```

# Max Wave (m)

## Row

### Column {.tabset}

#### All Stations

```{r}
#| title: "Max Wave (m): Tallest Individual Wave per Hour - All 5 Stations"
create_dygraph(buoy_hourly, "hmax", paste0("Max Wave (m) - Hourly Data (", date_caption, ")"))
```

**Caption:** Time series of [Max Wave](#reference) (m) showing the tallest individual wave recorded during each hourly 17.5-min measurement window. **Hourly data (`r n_hourly_display` observations`r sampling_note`)** - no daily averaging to preserve full detail. The highest Max Wave (`r max_hmax` m) was recorded at `r max_station_hmax`. M6 (deep Atlantic, 3000m depth) typically shows larger waves due to open ocean exposure.

#### M2

```{r}
#| title: "Max Wave (m) at M2 - Southwest Ireland (155m depth, 50km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M2"), "hmax", "Max Wave (m) - M2 Hourly", by_station = FALSE)
```

#### M3

```{r}
#| title: "Max Wave (m) at M3 - Southwest Ireland (155m depth, 65km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M3"), "hmax", "Max Wave (m) - M3 Hourly", by_station = FALSE)
```

#### M4

```{r}
#| title: "Max Wave (m) at M4 - Southeast Ireland (62m depth, 30km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M4"), "hmax", "Max Wave (m) - M4 Hourly", by_station = FALSE)
```

#### M5

```{r}
#| title: "Max Wave (m) at M5 - West Ireland (70m depth, 55km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M5"), "hmax", "Max Wave (m) - M5 Hourly", by_station = FALSE)
```

#### M6

```{r}
#| title: "Max Wave (m) at M6 - Northwest Atlantic (3000m depth, 320km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M6"), "hmax", "Max Wave (m) - M6 Hourly", by_station = FALSE)
```

## Row {height=35%}

### Column {.tabset}

#### Statistics

```{r}
hmax_stats <- buoy_data |>
  group_by(`Station` = station_id) |>
  summarise(
    `First` = format(min(time, na.rm = TRUE), "%Y-%m-%d"),
    `Last` = format(max(time, na.rm = TRUE), "%Y-%m-%d"),
    `Max Wave (m)` = round(max(hmax, na.rm = TRUE), 2),
    `Avg Max Wave (m)` = round(mean(hmax, na.rm = TRUE), 2),
    `Max Signif (m)` = round(max(wave_height, na.rm = TRUE), 2),
    `Avg Ratio` = round(mean(hmax / wave_height, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(desc(`Max Wave (m)`))

create_dt(hmax_stats, paste0(
  "Max Wave Statistics by Station (", date_caption, "). ",
  "Max Wave = tallest individual wave in each 17.5-min hourly window. ",
  "Avg Max Wave = mean of all hourly Max Wave values (typical tallest wave per hour). ",
  "Avg Ratio = mean(Max Wave / Signif Wave); typical 1.5-1.9, >2.0 = rogue conditions. ",
  "Conclusion: M6 records highest waves due to deep Atlantic exposure. See [Reference](#reference)."
))
```

#### Data Table

```{r}
hmax_data <- buoy_data |>
  select(time, station_id, hmax, wave_height) |>
  mutate(ratio = round(hmax / wave_height, 2)) |>
  arrange(desc(hmax)) |>
  head(5000) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    `Max Wave (m)` = round(hmax, 2),
    `Signif Wave (m)` = round(wave_height, 2),
    `Ratio` = ratio
  ) |>
  select(`Time (UTC)`, Station = station_id, `Max Wave (m)`, `Signif Wave (m)`, `Ratio`)

create_dt(hmax_data, paste0(
  "Max Wave Data ordered by tallest waves (top 5000 of ", n_records, " hourly records). ",
  "Max Wave = single tallest wave. Signif Wave = mean of top 1/3. ",
  "Ratio > 2.0 = potential rogue wave. Conclusion: Extreme waves cluster in winter storms."
))
```

# Signif Wave (m)

## Row

### Column {.tabset}

#### All Stations

```{r}
#| title: "Signif Wave (m): Mean of Top 1/3 Waves per Hour - All 5 Stations"
create_dygraph(buoy_hourly, "wave_height", paste0("Signif Wave (m) - Hourly Data (", date_caption, ")"))
```

**Caption:** Time series of [Signif Wave](#reference) (m) showing the mean of the highest 1/3 of waves measured during each hourly 17.5-min window. **Hourly data (`r n_hourly_display` observations`r sampling_note`)** - no daily averaging to preserve full detail. This metric smooths out individual wave variability while capturing sea state severity. Max Signif Wave: `r max_signif` m. Clear seasonal pattern with higher values in winter (Oct-Mar).

#### M2

```{r}
#| title: "Signif Wave (m) at M2 - Southwest Ireland (155m depth)"
create_dygraph(get_station_data(buoy_hourly, "M2"), "wave_height", "Signif Wave (m) - M2 Hourly", by_station = FALSE)
```

#### M3

```{r}
#| title: "Signif Wave (m) at M3 - Southwest Ireland (155m depth)"
create_dygraph(get_station_data(buoy_hourly, "M3"), "wave_height", "Signif Wave (m) - M3 Hourly", by_station = FALSE)
```

#### M4

```{r}
#| title: "Signif Wave (m) at M4 - Southeast Ireland (62m depth)"
create_dygraph(get_station_data(buoy_hourly, "M4"), "wave_height", "Signif Wave (m) - M4 Hourly", by_station = FALSE)
```

#### M5

```{r}
#| title: "Signif Wave (m) at M5 - West Ireland (70m depth)"
create_dygraph(get_station_data(buoy_hourly, "M5"), "wave_height", "Signif Wave (m) - M5 Hourly", by_station = FALSE)
```

#### M6

```{r}
#| title: "Signif Wave (m) at M6 - Northwest Atlantic (3000m depth)"
create_dygraph(get_station_data(buoy_hourly, "M6"), "wave_height", "Signif Wave (m) - M6 Hourly", by_station = FALSE)
```

## Row {height=35%}

### Column {.tabset}

#### Statistics

```{r}
wave_stats <- buoy_data |>
  group_by(`Station` = station_id) |>
  summarise(
    `Records (hourly)` = n(),
    `Max Wave (m)` = round(max(hmax, na.rm = TRUE), 2),
    `Max Signif (m)` = round(max(wave_height, na.rm = TRUE), 2),
    `Avg Signif (m)` = round(mean(wave_height, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(desc(`Max Signif (m)`))

create_dt(wave_stats, paste0(
  "Signif Wave Statistics by Station (", date_caption, "). ",
  "Signif Wave = mean of highest 1/3 of ~300 waves per 17.5-min window (hourly). ",
  "Max Wave = tallest single wave. ",
  "Conclusion: M6 has highest average Signif Wave due to Atlantic swell exposure. See [Reference](#reference)."
))
```

#### Data Table

```{r}
wave_data <- buoy_data |>
  select(time, station_id, wave_height, hmax, wave_period) |>
  arrange(desc(wave_height)) |>
  head(5000) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    `Signif Wave (m)` = round(wave_height, 2),
    `Max Wave (m)` = round(hmax, 2),
    `Wave Period (s)` = round(wave_period, 1)
  ) |>
  select(`Time (UTC)`, Station = station_id, `Signif Wave (m)`, `Max Wave (m)`, `Wave Period (s)`)

create_dt(wave_data, paste0(
  "Signif Wave Data ordered by largest values (top 5000 of ", n_records, " hourly records). ",
  "Signif Wave = mean of top 1/3. Wave Period = average time between wave crests (s). ",
  "Conclusion: Highest Signif Waves typically have longer periods (>10s), indicating Atlantic swell."
))
```

# Wind Speed (kn)

## Row

### Column {.tabset}

#### All Stations

```{r}
#| title: "Wind Speed (knots): 10-min Average per Hour - All 5 Stations"
create_dygraph(buoy_hourly, "wind_speed", paste0("Wind Speed (knots) - Hourly Data (", date_caption, ")"))
```

**Caption:** Time series of [Wind Speed](#reference) (knots) showing the 10-minute average wind speed recorded each hour. **Hourly data (`r n_hourly_display` observations`r sampling_note`)** - no daily averaging to preserve full detail. Wind measurements are taken at ~3m above sea level. Gust speeds (not shown here) can be 40-60% higher. Strong correlation with [Signif Wave](#reference) (r = `r cor_wind_wave`).

#### M2

```{r}
#| title: "Wind Speed (kn) at M2 - Southwest Ireland"
create_dygraph(get_station_data(buoy_hourly, "M2"), "wind_speed", "Wind Speed (kn) - M2 Hourly", by_station = FALSE)
```

#### M3

```{r}
#| title: "Wind Speed (kn) at M3 - Southwest Ireland"
create_dygraph(get_station_data(buoy_hourly, "M3"), "wind_speed", "Wind Speed (kn) - M3 Hourly", by_station = FALSE)
```

#### M4

```{r}
#| title: "Wind Speed (kn) at M4 - Southeast Ireland"
create_dygraph(get_station_data(buoy_hourly, "M4"), "wind_speed", "Wind Speed (kn) - M4 Hourly", by_station = FALSE)
```

#### M5

```{r}
#| title: "Wind Speed (kn) at M5 - West Ireland"
create_dygraph(get_station_data(buoy_hourly, "M5"), "wind_speed", "Wind Speed (kn) - M5 Hourly", by_station = FALSE)
```

#### M6

```{r}
#| title: "Wind Speed (kn) at M6 - Northwest Atlantic"
create_dygraph(get_station_data(buoy_hourly, "M6"), "wind_speed", "Wind Speed (kn) - M6 Hourly", by_station = FALSE)
```

## Row {height=35%}

### Column {.tabset}

#### Statistics

```{r}
wind_stats <- buoy_data |>
  group_by(`Station` = station_id) |>
  summarise(
    `Records (hourly)` = n(),
    `Avg Wind (kn)` = round(mean(wind_speed, na.rm = TRUE), 1),
    `Max Wind (kn)` = round(max(wind_speed, na.rm = TRUE), 1),
    `Max Gust (kn)` = round(max(gust, na.rm = TRUE), 1),
    `Gust Factor` = round(max(gust, na.rm = TRUE) / max(wind_speed, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(desc(`Max Gust (kn)`))

create_dt(wind_stats, paste0(
  "Wind Statistics by Station (", date_caption, "). ",
  "Wind Speed = 10-min average (hourly). Gust = peak 3-second wind (hourly). ",
  "Gust Factor = Gust/Wind ratio. ",
  "Conclusion: M6 experiences strongest sustained winds due to open Atlantic exposure. See [Reference](#reference)."
))
```

#### Data Table

```{r}
wind_data <- buoy_data |>
  select(time, station_id, wind_speed, gust, atmospheric_pressure) |>
  arrange(desc(gust)) |>
  head(5000) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    `Wind Speed (kn)` = round(wind_speed, 1),
    `Gust (kn)` = round(gust, 1),
    `Pressure (hPa)` = round(atmospheric_pressure, 1)
  ) |>
  select(`Time (UTC)`, Station = station_id, `Wind Speed (kn)`, `Gust (kn)`, `Pressure (hPa)`)

create_dt(wind_data, paste0(
  "Wind Data ordered by strongest gusts (top 5000 of ", n_records, " hourly records). ",
  "Gust = peak 3-second wind. Low pressure often accompanies high winds. ",
  "Conclusion: Strongest gusts cluster in winter storm events with pressure <990 hPa."
))
```

# Wind vs Wave

## Row

### Column {width=100%}

```{r}
#| fig-height: 18
# Compute regression slopes to order panels
reg_stats <- buoy_data |>
  filter(!is.na(wind_speed), !is.na(wave_height)) |>
  group_by(Station = station_id) |>
  summarise(
    Records = n(),
    Correlation = round(cor(wind_speed, wave_height), 2),
    Intercept = round(coef(lm(wave_height ~ wind_speed))[1], 2),
    Slope = round(coef(lm(wave_height ~ wind_speed))[2], 2),
    .groups = "drop"
  ) |>
  arrange(desc(Slope))

# Order stations by slope (largest first)
station_order <- reg_stats$Station
panel_labels <- paste0(station_order, " (slope=", sprintf("%.2f", reg_stats$Slope), ")")

# Sample data for scatter plot to keep HTML size manageable
set.seed(42)
scatter_sample <- buoy_data |>
  filter(!is.na(wind_speed), !is.na(wave_height))
if (nrow(scatter_sample) > 15000) {
  scatter_sample <- scatter_sample |> slice_sample(n = 15000)
}
plot_data <- scatter_sample |>
  mutate(
    station_label = factor(
      paste0(station_id, " (slope=", sprintf("%.2f", reg_stats$Slope[match(station_id, reg_stats$Station)]), ")"),
      levels = panel_labels
    )
  )

p <- ggplot(plot_data, aes(x = wind_speed, y = wave_height)) +
  geom_point(aes(
    text = paste0("Station: ", station_id, "<br>",
                  "Wind Speed: ", round(wind_speed, 1), " knots<br>",
                  "Signif Wave: ", round(wave_height, 2), " m<br>",
                  "Time: ", format(time, "%Y-%m-%d %H:%M"))
  ), color = "#1f77b4", alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linewidth = 1) +
  facet_wrap(~station_label, ncol = 2) +
  labs(
    x = "Wind Speed (knots) - 10-min average",
    y = "Signif Wave (m) - mean of top 1/3 waves"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.background = element_rect(fill = "grey95", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_line(color = "white")
  )

ggplotly(p, tooltip = "text") |>
  layout(
    title = list(
      text = paste0("Wind Speed vs Signif Wave by Station (", date_caption, ")"),
      font = list(size = 14),
      y = 0.98
    ),
    margin = list(t = 80),
    hoverlabel = list(bgcolor = "white", font = list(color = "black", size = 12))
  )
```

**Caption:** Scatter plots showing relationship between Wind Speed (knots) and Signif Wave (m) for each station using hourly observations. Panels ordered by regression slope (steepest first). The slope indicates how much wave height increases per unit wind speed. M6 (deep Atlantic) shows highest slope due to unlimited fetch for wave development. Correlation: r = `r cor_wind_wave`. See [Reference](#reference) for definitions.

## Row {height=25%}

### Column {.tabset}

#### Regression Statistics

```{r}
create_dt(reg_stats, paste0(
  "Linear Regression by Station ordered by slope (", date_caption, "). ",
  "Slope = Signif Wave increase (m) per knot of wind. ",
  "Higher correlation = wind explains more wave variability. ",
  "Conclusion: Deep-water stations (M6) show strongest wind-wave coupling."
))
```

#### Data Sample

```{r}
scatter_data <- buoy_data |>
  select(time, station_id, wind_speed, wave_height, hmax) |>
  filter(!is.na(wind_speed), !is.na(wave_height)) |>
  arrange(desc(wave_height)) |>
  head(5000) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    `Wind Speed (kn)` = round(wind_speed, 1),
    `Signif Wave (m)` = round(wave_height, 2),
    `Max Wave (m)` = round(hmax, 2)
  ) |>
  select(`Time (UTC)`, Station = station_id, `Wind Speed (kn)`, `Signif Wave (m)`, `Max Wave (m)`)

create_dt(scatter_data, paste0(
  "Wind-Wave Data ordered by Signif Wave (top 5000 of ", n_records, " hourly records). ",
  "Shows hourly measurements used in regression analysis. ",
  "Conclusion: High Signif Waves generally coincide with high winds, but swell can arrive without local wind."
))
```

# Data

## Row

### Column {.tabset}

#### Most Recent 1000 Hours

```{r}
recent_data <- buoy_data |>
  arrange(desc(time)) |>
  head(1000) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    Station = station_id,
    `Max Wave (m)` = round(hmax, 2),
    `Signif Wave (m)` = round(wave_height, 2),
    `Wind (kn)` = round(wind_speed, 1),
    `Gust (kn)` = round(gust, 1),
    `Air Temp (C)` = round(air_temperature, 1),
    `Sea Temp (C)` = round(sea_temperature, 1),
    `Pressure (hPa)` = round(atmospheric_pressure, 1)
  ) |>
  select(`Time (UTC)`, Station, `Max Wave (m)`, `Signif Wave (m)`, `Wind (kn)`, `Gust (kn)`, `Air Temp (C)`, `Sea Temp (C)`, `Pressure (hPa)`)

datatable(
  recent_data,
  caption = htmltools::HTML(paste0(
    "<b>Most recent 1000 hourly observations</b> (of ", n_records, " total). ",
    "<b>Max Wave</b> = tallest single wave in 17.5-min window. ",
    "<b>Signif Wave</b> = mean of top 1/3 waves. ",
    "<b>Wind</b> = 10-min average. <b>Gust</b> = peak 3-sec wind. ",
    "See <a href='#reference'>Reference</a> for definitions. ",
    "<b>Note:</b> Data starts ", format(min(buoy_data$time), "%Y"), " - for historical data back to 2002, ",
    "reinitialize database with <code>initialize_database(start_date = '2001-01-01')</code>."
  )),
  filter = 'top',
  extensions = c('Buttons', 'FixedHeader'),
  options = list(
    pageLength = 25,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel'),
    order = list(list(0, 'desc')),
    scrollX = TRUE,
    fixedHeader = TRUE,
    columnDefs = list(list(width = '100px', targets = 0))
  ),
  class = 'cell-border stripe hover compact'
) %>%
  DT::formatStyle(columns = 1, fontSize = '10px', whiteSpace = 'nowrap')
```

#### Rogue Waves (Ratio > 2.0)

```{r}
rogue_waves <- buoy_data |>
  filter(!is.na(hmax), !is.na(wave_height), wave_height > 0) |>
  mutate(ratio = hmax / wave_height) |>
  filter(ratio > 2.0) |>
  arrange(desc(ratio)) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    Station = station_id,
    `Max Wave (m)` = round(hmax, 2),
    `Signif Wave (m)` = round(wave_height, 2),
    `Ratio` = round(ratio, 2),
    `Wind (kn)` = round(wind_speed, 1),
    `Gust (kn)` = round(gust, 1)
  ) |>
  select(`Time (UTC)`, Station, `Max Wave (m)`, `Signif Wave (m)`, `Ratio`, `Wind (kn)`, `Gust (kn)`)

n_rogue <- nrow(rogue_waves)

datatable(
  rogue_waves,
  caption = htmltools::HTML(paste0(
    "<b>Rogue Wave Events</b> (", format(n_rogue, big.mark = ","), " events where Ratio > 2.0). ",
    "A rogue wave is defined as Max Wave > 2× Signif Wave. ",
    "These are potentially dangerous individual waves that exceed normal sea state. ",
    "Sorted by Ratio (most extreme first). ",
    "See <a href='#rogue-waves'>Rogue Waves</a> reference for interpretation."
  )),
  filter = 'top',
  extensions = c('Buttons', 'FixedHeader'),
  options = list(
    pageLength = 25,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel'),
    order = list(list(4, 'desc')),
    scrollX = TRUE,
    fixedHeader = TRUE,
    columnDefs = list(list(width = '100px', targets = 0))
  ),
  class = 'cell-border stripe hover compact'
) %>%
  DT::formatStyle(columns = 1, fontSize = '10px', whiteSpace = 'nowrap')
```

#### Extreme Gusts (Top 1%)

```{r}
gust_threshold <- quantile(buoy_data$gust, 0.99, na.rm = TRUE)
extreme_gusts <- buoy_data |>
  filter(!is.na(gust), gust >= gust_threshold) |>
  arrange(desc(gust)) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    Station = station_id,
    `Gust (kn)` = round(gust, 1),
    `Wind (kn)` = round(wind_speed, 1),
    `Gust Factor` = round(gust / wind_speed, 2),
    `Max Wave (m)` = round(hmax, 2),
    `Pressure (hPa)` = round(atmospheric_pressure, 1)
  ) |>
  select(`Time (UTC)`, Station, `Gust (kn)`, `Wind (kn)`, `Gust Factor`, `Max Wave (m)`, `Pressure (hPa)`)

n_gusts <- nrow(extreme_gusts)

datatable(
  extreme_gusts,
  caption = htmltools::HTML(paste0(
    "<b>Extreme Gust Events</b> (", format(n_gusts, big.mark = ","), " events in top 1%, ≥", round(gust_threshold, 1), " kn). ",
    "Gust Factor = Gust / Wind Speed; typical 1.3-1.6, >1.8 indicates gusty conditions. ",
    "Sorted by strongest gusts. Low pressure often accompanies extreme wind events."
  )),
  filter = 'top',
  extensions = c('Buttons', 'FixedHeader'),
  options = list(
    pageLength = 25,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel'),
    order = list(list(2, 'desc')),
    scrollX = TRUE,
    fixedHeader = TRUE,
    columnDefs = list(list(width = '100px', targets = 0))
  ),
  class = 'cell-border stripe hover compact'
) %>%
  DT::formatStyle(columns = 1, fontSize = '10px', whiteSpace = 'nowrap')
```

# Reference

## Row

### Column {.tabset}

#### Variable Definitions

##### {.tabset}

###### Column Names

**Wave Measurements:**

| Variable | Display Name | Units | Definition |
|----------|--------------|-------|------------|
| `hmax` | **Max Wave** | m | Tallest individual wave (crest-to-trough) during each [17.5-min hourly measurement window](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html) |
| `wave_height` | **Signif Wave** | m | Mean height of the highest 1/3 of waves (~35 of ~105 waves) during each 17.5-min hourly window. [NOAA definition](https://www.weather.gov/key/marine_sigwave) |
| `wave_period` | **Wave Period** | s | Average time between successive wave crests via [zero-crossing analysis](https://www.nortekgroup.com/knowledge-center/wiki/new-to-waves) |

**Key distinction:** Max Wave is a single wave; Signif Wave is an average of the top ~35 waves (highest 1/3). Both come from the same 17.5-min measurement window each hour.

**Wind Measurements:**

| Variable | Display Name | Units | Definition |
|----------|--------------|-------|------------|
| `wind_speed` | **Wind Speed** | knots | 10-minute average wind speed measured each hour at ~3m above sea level |
| `gust` | **Gust** | knots | Peak 3-second wind speed recorded during the hour |

**Atmospheric:**

| Variable | Display Name | Units | Definition |
|----------|--------------|-------|------------|
| `atmospheric_pressure` | **Pressure** | hPa | Barometric pressure at sea level. Low pressure (<1000 hPa) often indicates storm systems |
| `air_temperature` | **Air Temp** | C | Air temperature at buoy (~3m above sea) |
| `sea_temperature` | **Sea Temp** | C | Sea surface temperature |

###### Aggregated Stats

**Computed Statistics (aggregated over all hourly records):**

| Statistic | Source Column | Definition |
|-----------|---------------|------------|
| **Max Wave (m)** | `hmax` | Highest `hmax` across all hours (largest single wave ever recorded in dataset) |
| **Max Signif (m)** | `wave_height` | Highest `wave_height` across all hours (highest hourly "average of top 1/3" value) |
| **Avg Signif (m)** | `wave_height` | Mean of all hourly `wave_height` values (typical sea state) |
| **Avg Max Wave (m)** | `hmax` | Mean of all hourly `hmax` values (typical tallest wave per hour) |
| **Avg Ratio** | `hmax / wave_height` | Mean ratio of Max Wave to Signif Wave. Typical: 1.5-1.9. >2.0 = rogue conditions |

**Why different?** Max Wave (m) and Max Signif (m) come from DIFFERENT columns:
- Max Wave = largest `hmax` = one extreme wave
- Max Signif = largest `wave_height` = one extreme hourly average

**Example:** A station might record Max Wave = 15.2m (one extreme individual wave) and Max Signif = 10.8m (one extreme hourly average of top 1/3 waves). These come from different measurement windows.

###### Ratios

**Max Wave / Signif Wave Ratio Interpretation:**

A **rogue wave** is typically defined as a wave with height > 2× the Signif Wave.

| Ratio (Max Wave / Signif Wave) | Interpretation |
|--------------------------------|----------------|
| 1.5 - 1.9 | Normal sea state |
| > 2.0 | Potential rogue wave conditions |

Rogue waves are rare but dangerous. They can appear without warning in otherwise moderate seas. The ratio helps identify when individual waves significantly exceed the average sea state.

**Technical Terms (for reference only):**

| Term | Equivalent | Notes |
|------|------------|-------|
| Hs, H1/3, Significant Wave Height | Signif Wave | All mean the same thing |
| Hmax, Maximum Wave Height | Max Wave | Largest individual wave |
| 4σ | Signif Wave | Statistical definition: 4 × standard deviation of sea surface elevation |

These technical terms appear in scientific literature. This dashboard uses "Signif Wave" and "Max Wave" consistently.

#### Methodology

##### {.tabset}

###### Measurement Window

Each buoy samples the sea surface **once per hour**. At the top of each hour, a [**17.5-minute measurement period**](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html) begins. During this window, the buoy measures approximately 105 individual wave heights (assuming ~10s wave period).

**Sampling limitation:** The buoy does NOT measure continuously. Only the 17.5-min window is sampled; waves occurring in the remaining 42.5 minutes of each hour are not recorded. Source: [Marine Institute ERDDAP metadata](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html) states *"The sensor measures this for 17.5 minutes and then takes an average."*

From this 17.5-min sample:

1. Waves are sorted by height, tallest first
2. **Max Wave**: The single tallest wave = `hmax`
3. **Signif Wave**: Mean of the highest 1/3 (~35 waves) = `wave_height` ([NOAA definition](https://www.weather.gov/key/marine_sigwave))

Both Max Wave and Signif Wave come from the SAME 17.5-min window, reported at the same timestamp.

###### Wind Measurement

**Wind Speed vs Gust:**

Wind measurements are taken at approximately 3 meters above sea level. Two distinct metrics are reported:

| Measurement | Averaging Period | Definition |
|-------------|-----------------|------------|
| **Wind Speed** | 10-minute average | Mean wind speed over a 10-minute period, measured once per hour |
| **Gust** | 3-second peak | Maximum 3-second average wind speed during the hour |

**Gust Factor:** The ratio of Gust to Wind Speed indicates atmospheric turbulence and stability:

| Gust Factor | Interpretation |
|-------------|----------------|
| 1.3 - 1.6 | Normal conditions |
| > 1.8 | Gusty/unstable conditions |

High gust factors often occur with convective activity, frontal passages, or flow over rough terrain/seas. Low pressure systems (<1000 hPa) frequently accompany extreme wind events.

###### Station Info

```{r}
create_dt(station_info, paste0(
  "Irish Weather Buoy Network Stations. ",
  "Water depth affects wave behavior: shallow (M4) vs deep (M6) stations show different wave characteristics. ",
  "Distance offshore affects wind exposure and fetch length. ",
  "M6 is in deep Atlantic waters (3000m depth, 320km offshore), experiencing open ocean conditions. ",
  "Other stations are on the continental shelf (62-155m depth, 30-65km offshore)."
))
```

#### Data Sources

**Primary Data:**

- [Marine Institute Ireland](https://www.marine.ie/) - Irish Weather Buoy Network operator
- [ERDDAP Data Portal](https://erddap.marine.ie/erddap/tabledap/IWBNetwork.html) - Access raw buoy data
- [ERDDAP Metadata](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html) - Dataset documentation

**Reference Documentation:**

- [NOAA Significant Wave Height Definition](https://www.weather.gov/key/marine_sigwave)
- [Nortek Wave Measurement Guide](https://www.nortekgroup.com/knowledge-center/wiki/new-to-waves) - Zero-crossing analysis explanation

**Code & Analysis:**

- [irishbuoys R Package](https://github.com/JohnGavin/irishbuoys) - Data processing and dashboard code

---

*Dashboard generated `r generated_time` using irishbuoys v`r pkg_version`*

*Data coverage: `r date_caption` | Total records: `r n_records` hourly observations*
