---
title: "Irish Weather Buoy Explorer"
format:
  dashboard:
    orientation: rows
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(dygraphs)
library(xts)
library(tidyr)

# Load pre-computed data from targets pipeline
data_dir <- "../inst/extdata"
if (!dir.exists(data_dir)) {
  data_dir <- system.file("extdata", package = "irishbuoys")
}

buoy_data <- readRDS(file.path(data_dir, "dashboard_buoy_data.rds"))
dashboard_stats <- readRDS(file.path(data_dir, "dashboard_stats.rds"))
dashboard_ts <- readRDS(file.path(data_dir, "dashboard_timeseries.rds"))

# Get package version
pkg_version <- tryCatch(
  as.character(packageVersion("irishbuoys")),
  error = function(e) {
    desc_file <- "../DESCRIPTION"
    if (file.exists(desc_file)) {
      desc <- read.dcf(desc_file)
      desc[1, "Version"]
    } else "dev"
  }
)

# Get unique stations
stations <- sort(unique(buoy_data$station_id))

# Create daily averages for time series (reduces HTML size)
buoy_daily <- buoy_data |>
  mutate(date = as.Date(time)) |>
  group_by(station_id, date) |>
  summarise(
    time = as.POSIXct(paste0(unique(date), " 12:00:00"), tz = "UTC"),
    wave_height = mean(wave_height, na.rm = TRUE),
    hmax = max(hmax, na.rm = TRUE),
    wave_period = mean(wave_period, na.rm = TRUE),
    wind_speed = mean(wind_speed, na.rm = TRUE),
    gust = max(gust, na.rm = TRUE),
    atmospheric_pressure = mean(atmospheric_pressure, na.rm = TRUE),
    air_temperature = mean(air_temperature, na.rm = TRUE),
    sea_temperature = mean(sea_temperature, na.rm = TRUE),
    .groups = "drop"
  ) |>
  select(-date)

# Calculate summary stats
valid_ww <- complete.cases(buoy_data$wind_speed, buoy_data$wave_height)
valid_wh <- complete.cases(buoy_data$wave_height, buoy_data$hmax)
cor_wind_wave <- round(cor(buoy_data$wind_speed[valid_ww], buoy_data$wave_height[valid_ww]), 3)
cor_wave_hmax <- round(cor(buoy_data$wave_height[valid_wh], buoy_data$hmax[valid_wh]), 3)
days_span <- round(as.numeric(difftime(max(buoy_data$time), min(buoy_data$time), units = "days")))
generated_time <- format(Sys.time(), "%Y-%m-%d %H:%M UTC")
date_caption <- paste0(format(min(buoy_data$time), "%Y-%m-%d"), " to ", format(max(buoy_data$time), "%Y-%m-%d"))

# Helper function to create station-specific data
get_station_data <- function(data, station) {
  if (station != "All Buoys") {
    data <- data |> filter(station_id == station)
  }
  data
}

# Helper function to create dygraph time series - full width
create_dygraph <- function(data, variable = "wave_height", title = "", by_station = TRUE) {
  if (by_station && length(unique(data$station_id)) > 1) {
    wide_data <- data |>
      select(time, station_id, all_of(variable)) |>
      tidyr::pivot_wider(names_from = station_id, values_from = all_of(variable))
    ts_data <- xts(wide_data[, -1], order.by = wide_data$time)
  } else {
    ts_data <- xts(data[[variable]], order.by = data$time)
    colnames(ts_data) <- variable
  }

  dygraph(ts_data, main = title, width = "100%") |>
    dyOptions(
      labelsUTC = TRUE,
      fillGraph = TRUE,
      fillAlpha = 0.1,
      drawGrid = TRUE,
      colors = RColorBrewer::brewer.pal(max(3, ncol(ts_data)), "Set1")[1:ncol(ts_data)]
    ) |>
    dyRangeSelector(height = 40) |>
    dyCrosshair(direction = "vertical") |>
    dyHighlight(highlightCircleSize = 5, highlightSeriesBackgroundAlpha = 0.2, hideOnMouseOut = FALSE) |>
    dyLegend(show = "always", width = 400)
}

# Helper function for DT tables with optimized timestamp display
create_dt <- function(data, caption = "", time_width = "140px") {
  time_cols <- which(names(data) %in% c("time", "Time", "timestamp", "Timestamp", "datetime", "Time (UTC)"))

  dt <- datatable(
    data,
    caption = caption,
    filter = 'top',
    extensions = 'Buttons',
    options = list(
      pageLength = 15,
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel'),
      order = list(list(0, 'desc')),
      scrollX = TRUE,
      columnDefs = if (length(time_cols) > 0) {
        list(list(width = time_width, targets = time_cols - 1))
      } else NULL
    ),
    class = 'cell-border stripe hover compact'
  )

  if (length(time_cols) > 0) {
    dt <- dt %>%
      DT::formatStyle(columns = time_cols, fontSize = '10px', whiteSpace = 'nowrap')
  }

  dt
}

# Station information with coordinates and depths
station_info <- data.frame(
  Station = c("M2", "M3", "M4", "M5", "M6"),
  Location = c("Southwest Ireland", "Southwest Ireland", "Southeast Ireland", "West Ireland", "Northwest Ireland"),
  Latitude = c(51.22, 51.22, 51.69, 51.69, 53.07),
  Longitude = c(-9.99, -10.55, -6.70, -10.00, -15.93),
  `Water Depth (m)` = c(155, 155, 62, 70, 3000),
  `Distance from Shore (km)` = c(50, 65, 30, 55, 320),
  check.names = FALSE
)
```

# Overview

## Row

### Column {width=100%}

```{r}
#| title: "Wave Height Time Series - All Stations (Full Dataset)"
create_dygraph(buoy_daily, "wave_height", "Signif Wave (m) - Daily Means")
```

## Row {height=25%}

### Column

**Dataset Summary:** `r format(nrow(buoy_data), big.mark=",")` records from `r length(stations)` stations (`r paste(stations, collapse=", ")`) spanning `r days_span` days (`r format(min(buoy_data$time), "%Y-%m-%d")` to `r format(max(buoy_data$time), "%Y-%m-%d")`). Wind-wave correlation: `r cor_wind_wave`, Wave-Hmax correlation: `r cor_wave_hmax`.

| Metric | Value | Metric | Value |
|--------|-------|--------|-------|
| Total Records | `r format(nrow(buoy_data), big.mark=",")` | Wind-Wave Correlation | `r cor_wind_wave` |
| Active Stations | `r length(stations)` | Wave-Hmax Correlation | `r cor_wave_hmax` |
| Date Range | `r format(min(buoy_data$time), "%Y-%m-%d")` - `r format(max(buoy_data$time), "%Y-%m-%d")` | Max Wave | `r round(max(buoy_data$hmax, na.rm=TRUE), 1)` m |
| Time Span | `r days_span` days | Max Wind Gust | `r round(max(buoy_data$gust, na.rm=TRUE), 1)` knots |

*Dashboard generated `r generated_time` using irishbuoys v`r pkg_version`*

# Max Wave (m)

## Row

### Column {.tabset}

#### All Stations

```{r}
create_dygraph(buoy_daily, "hmax", "Max Wave (m) - All Stations (Daily Max)")
```

#### M2

```{r}
create_dygraph(get_station_data(buoy_daily, "M2"), "hmax", "Max Wave - M2 (Daily)", by_station = FALSE)
```

#### M3

```{r}
create_dygraph(get_station_data(buoy_daily, "M3"), "hmax", "Max Wave - M3 (Daily)", by_station = FALSE)
```

#### M4

```{r}
create_dygraph(get_station_data(buoy_daily, "M4"), "hmax", "Max Wave - M4 (Daily)", by_station = FALSE)
```

#### M5

```{r}
create_dygraph(get_station_data(buoy_daily, "M5"), "hmax", "Max Wave - M5 (Daily)", by_station = FALSE)
```

#### M6

```{r}
create_dygraph(get_station_data(buoy_daily, "M6"), "hmax", "Max Wave - M6 (Daily)", by_station = FALSE)
```

## Row {height=35%}

### Column {.tabset}

#### Statistics

```{r}
hmax_stats <- buoy_data |>
  group_by(station_id) |>
  summarise(
    Records = n(),
    `Max Wave (m)` = round(max(hmax, na.rm = TRUE), 2),
    `Mean Max Wave (m)` = round(mean(hmax, na.rm = TRUE), 2),
    `Max Signif (m)` = round(max(wave_height, na.rm = TRUE), 2),
    `Avg Ratio` = round(mean(hmax / wave_height, na.rm = TRUE), 2),
    .groups = "drop"
  )
create_dt(hmax_stats, paste0("Max Wave Statistics by Station (", date_caption, "). Max Wave: largest individual wave from 17.5-min measurement. Ratio > 2.0 indicates rogue wave. [See Reference](#reference)"))
```

#### Data Table

```{r}
hmax_data <- buoy_data |>
  select(time, station_id, hmax, wave_height) |>
  mutate(ratio = round(hmax / wave_height, 2)) |>
  arrange(desc(hmax)) |>
  head(5000) |>
  mutate(
    time = format(time, "%Y-%m-%d %H:%M"),
    hmax = round(hmax, 2),
    wave_height = round(wave_height, 2)
  ) |>
  rename(`Max Wave (m)` = hmax, `Signif (m)` = wave_height, `Ratio` = ratio)
create_dt(hmax_data, paste0("Max Wave Data ordered by highest waves (most recent 5000 of ", format(nrow(buoy_data), big.mark=","), " records)"))
```

# Signif Wave (m)

## Row

### Column {.tabset}

#### All Stations

```{r}
create_dygraph(buoy_daily, "wave_height", "Signif Wave (m) - All Stations (Daily Means)")
```

#### M2

```{r}
create_dygraph(get_station_data(buoy_daily, "M2"), "wave_height", "Signif Wave - M2 (Daily)", by_station = FALSE)
```

#### M3

```{r}
create_dygraph(get_station_data(buoy_daily, "M3"), "wave_height", "Signif Wave - M3 (Daily)", by_station = FALSE)
```

#### M4

```{r}
create_dygraph(get_station_data(buoy_daily, "M4"), "wave_height", "Signif Wave - M4 (Daily)", by_station = FALSE)
```

#### M5

```{r}
create_dygraph(get_station_data(buoy_daily, "M5"), "wave_height", "Signif Wave - M5 (Daily)", by_station = FALSE)
```

#### M6

```{r}
create_dygraph(get_station_data(buoy_daily, "M6"), "wave_height", "Signif Wave - M6 (Daily)", by_station = FALSE)
```

## Row {height=35%}

### Column {.tabset}

#### Statistics

```{r}
wave_stats <- buoy_data |>
  group_by(station_id) |>
  summarise(
    Records = n(),
    `Max Wave (m)` = round(max(hmax, na.rm = TRUE), 2),
    `Max Signif (m)` = round(max(wave_height, na.rm = TRUE), 2),
    `Avg Signif (m)` = round(mean(wave_height, na.rm = TRUE), 2),
    .groups = "drop"
  )
create_dt(wave_stats, "Wave Statistics by Station. Max Wave: largest individual wave from 17.5-min measurement. Signif Wave: mean of top 1/3 of waves. [See Reference](#reference)")
```

#### Data Table

```{r}
wave_data <- buoy_data |>
  select(time, station_id, wave_height, hmax, wave_period) |>
  arrange(desc(time)) |>
  head(5000) |>
  mutate(
    time = format(time, "%Y-%m-%d %H:%M"),
    wave_height = round(wave_height, 2),
    hmax = round(hmax, 2),
    wave_period = round(wave_period, 1)
  )
create_dt(wave_data, paste0("Wave Height Data (most recent 5000 of ", format(nrow(buoy_data), big.mark=","), " records)"))
```

# Wind Speed

## Row

### Column {.tabset}

#### All Stations

```{r}
create_dygraph(buoy_daily, "wind_speed", "Wind Speed - All Stations (Daily Means)")
```

#### M2

```{r}
create_dygraph(get_station_data(buoy_daily, "M2"), "wind_speed", "Wind Speed - M2 (Daily)", by_station = FALSE)
```

#### M3

```{r}
create_dygraph(get_station_data(buoy_daily, "M3"), "wind_speed", "Wind Speed - M3 (Daily)", by_station = FALSE)
```

#### M4

```{r}
create_dygraph(get_station_data(buoy_daily, "M4"), "wind_speed", "Wind Speed - M4 (Daily)", by_station = FALSE)
```

#### M5

```{r}
create_dygraph(get_station_data(buoy_daily, "M5"), "wind_speed", "Wind Speed - M5 (Daily)", by_station = FALSE)
```

#### M6

```{r}
create_dygraph(get_station_data(buoy_daily, "M6"), "wind_speed", "Wind Speed - M6 (Daily)", by_station = FALSE)
```

## Row {height=35%}

### Column {.tabset}

#### Statistics

```{r}
wind_stats <- buoy_data |>
  group_by(station_id) |>
  summarise(
    Records = n(),
    `Mean Wind (kn)` = round(mean(wind_speed, na.rm = TRUE), 1),
    `Max Wind (kn)` = round(max(wind_speed, na.rm = TRUE), 1),
    `Max Gust (kn)` = round(max(gust, na.rm = TRUE), 1),
    .groups = "drop"
  )
create_dt(wind_stats, "Wind Statistics by Station")
```

#### Data Table

```{r}
wind_data <- buoy_data |>
  select(time, station_id, wind_speed, gust, atmospheric_pressure) |>
  arrange(desc(time)) |>
  head(5000) |>
  mutate(
    time = format(time, "%Y-%m-%d %H:%M"),
    wind_speed = round(wind_speed, 1),
    gust = round(gust, 1),
    atmospheric_pressure = round(atmospheric_pressure, 1)
  )
create_dt(wind_data, paste0("Wind Data (most recent 5000 of ", format(nrow(buoy_data), big.mark=","), " records)"))
```

# Wind vs Wave

## Row

### Column {width=100%}

```{r}
#| fig-height: 18
# Compute regression slopes to order panels
reg_stats <- buoy_data |>
  filter(!is.na(wind_speed), !is.na(wave_height)) |>
  group_by(Station = station_id) |>
  summarise(
    Records = n(),
    Correlation = round(cor(wind_speed, wave_height), 2),
    Intercept = round(coef(lm(wave_height ~ wind_speed))[1], 2),
    Slope = round(coef(lm(wave_height ~ wind_speed))[2], 2),
    .groups = "drop"
  ) |>
  arrange(desc(Slope))

# Order stations by slope (largest first)
station_order <- reg_stats$Station
panel_labels <- paste0(station_order, " (slope=", sprintf("%.2f", reg_stats$Slope), ")")

# Sample data for scatter plot to keep HTML size manageable
set.seed(42)
scatter_sample <- buoy_data |>
  filter(!is.na(wind_speed), !is.na(wave_height))
if (nrow(scatter_sample) > 15000) {
  scatter_sample <- scatter_sample |> slice_sample(n = 15000)
}
plot_data <- scatter_sample |>
  mutate(
    station_label = factor(
      paste0(station_id, " (slope=", sprintf("%.2f", reg_stats$Slope[match(station_id, reg_stats$Station)]), ")"),
      levels = panel_labels
    )
  )

p <- ggplot(plot_data, aes(x = wind_speed, y = wave_height)) +
  geom_point(aes(
    text = paste0("Station: ", station_id, "<br>",
                  "Wind: ", round(wind_speed, 1), " knots<br>",
                  "Signif Wave: ", round(wave_height, 2), " m<br>",
                  "Time: ", format(time, "%Y-%m-%d %H:%M"))
  ), color = "#1f77b4", alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linewidth = 1) +
  facet_wrap(~station_label, ncol = 2) +
  labs(x = "Wind Speed (knots)", y = "Signif Wave (m)") +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.background = element_rect(fill = "grey95", color = NA),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_line(color = "white")
  )

ggplotly(p, tooltip = "text") |>
  layout(
    title = paste0("Wind Speed vs Signif Wave - ordered by slope (", date_caption, ")"),
    hoverlabel = list(bgcolor = "white", font = list(color = "black", size = 12))
  )
```

## Row {height=30%}

### Column {.tabset}

#### Regression Stats

```{r}
create_dt(reg_stats, "Linear Regression by Station (ordered by slope)")
```

#### Data

```{r}
scatter_data <- buoy_data |>
  select(time, station_id, wind_speed, wave_height, hmax) |>
  filter(!is.na(wind_speed), !is.na(wave_height)) |>
  arrange(desc(time)) |>
  head(5000) |>
  mutate(
    time = format(time, "%Y-%m-%d %H:%M"),
    wind_speed = round(wind_speed, 1),
    wave_height = round(wave_height, 2),
    hmax = round(hmax, 2)
  )
create_dt(scatter_data, paste0("Wind-Wave Data (most recent 5000 of ", format(nrow(buoy_data), big.mark=","), " records)"))
```

# Recent Data

## Row

### Column

```{r}
recent_data <- buoy_data |>
  arrange(desc(time)) |>
  head(500) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    Station = station_id,
    `Max Wave (m)` = round(hmax, 2),
    `Signif (m)` = round(wave_height, 2),
    `Wind (kn)` = round(wind_speed, 1),
    `Gust (kn)` = round(gust, 1),
    `Air (C)` = round(air_temperature, 1),
    `Sea (C)` = round(sea_temperature, 1),
    `hPa` = round(atmospheric_pressure, 1)
  ) |>
  select(`Time (UTC)`, Station, `Max Wave (m)`, `Signif (m)`, `Wind (kn)`, `Gust (kn)`, `Air (C)`, `Sea (C)`, `hPa`)

datatable(
  recent_data,
  caption = "Most recent 500 observations",
  filter = 'top',
  extensions = 'Buttons',
  options = list(
    pageLength = 25,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel'),
    order = list(list(0, 'desc')),
    scrollX = TRUE,
    columnDefs = list(list(width = '80px', targets = 0))
  ),
  class = 'cell-border stripe hover compact'
) %>%
  DT::formatStyle(columns = 1, fontSize = '10px', whiteSpace = 'nowrap')
```

# Reference

## Row

### Column {.tabset}

#### Column Names

| Column Variable | Display Name | Units | Definition |
|-----------------|--------------|-------|------------|
| `hmax` | **Max Wave** | m | Largest individual wave during each 17.5-min measurement (hourly) |
| `wave_height` | **Signif Wave** | m | Mean of highest 1/3 of waves during each 17.5-min measurement (hourly) |
| `wind_speed` | Wind Speed | knots | 10-minute average wind speed |
| `gust` | Wind Gust | knots | Maximum 3-second gust during the hour |
| `wave_period` | Wave Period | s | Average wave period via zero-crossing analysis |
| `atmospheric_pressure` | Pressure | hPa | Barometric pressure |
| `air_temperature` | Air Temp | C | Air temperature at buoy (~3m above sea) |
| `sea_temperature` | Sea Temp | C | Sea surface temperature |

#### Aggregated Stats

| Statistic | Display Name | Definition |
|-----------|--------------|------------|
| Max of hourly `hmax` | Max Wave (m) | Largest individual wave recorded from dataset start to end |
| Max of hourly `wave_height` | Max Signif (m) | Highest hourly Signif Wave value over dataset period |
| Mean of hourly `wave_height` | Avg Signif (m) | Mean of hourly Signif Wave values over dataset period |

**Note:** Max Wave column appears FIRST in all tables, followed by Signif Wave columns.

#### Ratios

| Ratio | Typical Range | Interpretation |
|-------|---------------|----------------|
| Max Wave / Signif Wave | 1.5 - 1.9 | Normal sea state |
| Max Wave / Signif Wave | > 2.0 | Potential rogue wave conditions |

This ratio interpretation also appears in captions for plots/tables showing these columns.

#### Measurement Window

Each buoy samples the sea surface every hour. At the top of each hour, a **17.5-minute measurement period** begins. During this window, the buoy measures approximately 300 individual wave heights (crest to preceding trough).

**Important:** The buoy does NOT measure continuously. Only the 17.5-min window is sampled; waves occurring in the remaining 42.5 minutes of each hour are not recorded.

From this 17.5-min sample:

1. Waves are sorted by height, tallest first
2. **Signif Wave**: Mean of the highest 1/3 (~100 waves) = "mean of top 1/3"
3. **Max Wave**: The single tallest wave

Both Signif Wave and Max Wave come from the SAME 17.5-min window.

#### Wind Measurement

| Measurement | Window | Frequency | What it Represents |
|-------------|--------|-----------|-------------------|
| Wind Speed | 10 min | Hourly | Average wind during 10-min window |
| Wind Gust | 3 sec | Hourly | Peak 3-second gust in the hour |

#### Glossary

**Technical Terms (for reference only):**

| Term | Equivalent | Notes |
|------|------------|-------|
| Hs, H1/3, Significant Wave Height | Signif Wave | All mean the same thing |
| Hmax, Maximum Wave Height | Max Wave | Largest individual wave |
| 4σ | Signif Wave | Statistical definition: 4 × standard deviation of sea surface |

These technical terms appear in scientific literature. This dashboard uses "Signif Wave" and "Max Wave" consistently.

## Row

### Column

#### Station Information

```{r}
create_dt(station_info, "Irish Weather Buoy Network Stations")
```

## Row {height=20%}

### Column

**Data Sources:** [Marine Institute Ireland](https://www.marine.ie/) | [ERDDAP Data Portal](https://erddap.marine.ie/erddap/tabledap/IWBNetwork.html) | [irishbuoys R Package](https://github.com/JohnGavin/irishbuoys)

*Dashboard generated `r generated_time` using irishbuoys v`r pkg_version`*
