---
title: "Buoy Data Explorer (Perspective)"
format:
  html:
    code-fold: true
    code-summary: "Show code"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This dashboard uses [perspectiveR](https://github.com/EydlinIlya/perspectiveR)
to render buoy data client-side via WebAssembly. Data is pre-aggregated to daily
resolution (~12K rows vs 300K hourly), keeping the HTML under 3MB.

See [issue #72](https://github.com/JohnGavin/irishbuoys/issues/72) for context.

## Data

Pre-aggregate hourly data to daily summaries. This reduces 300K rows to ~12K
while preserving the key metrics users care about.

```{r load-data}
parquet_path <- file.path("..", "docs", "vignettes", "data", "buoy_data.parquet")
if (!file.exists(parquet_path)) {
  parquet_path <- system.file(
    "extdata", "dashboard_buoy_data.rds",
    package = "irishbuoys"
  )
  buoy_data <- readRDS(parquet_path)
} else {
  buoy_data <- arrow::read_parquet(parquet_path)
}

# Pre-aggregate to daily resolution
buoy_daily <- buoy_data |>
  dplyr::mutate(date = as.Date(time)) |>
  dplyr::summarise(
    mean_wave_height = round(mean(wave_height, na.rm = TRUE), 2),
    max_hmax = round(max(hmax, na.rm = TRUE), 2),
    mean_wind_speed = round(mean(wind_speed, na.rm = TRUE), 1),
    max_gust = round(max(gust, na.rm = TRUE), 1),
    mean_sea_temp = round(mean(sea_temperature, na.rm = TRUE), 2),
    mean_pressure = round(mean(atmospheric_pressure, na.rm = TRUE), 1),
    n_obs = dplyr::n(),
    .by = c(station_id, date)
  )

# Replace Inf/-Inf from empty groups with NA
buoy_daily[buoy_daily == Inf | buoy_daily == -Inf] <- NA

cat(sprintf(
  "Aggregated: %s daily rows from %s hourly rows (%s to %s)\n",
  format(nrow(buoy_daily), big.mark = ","),
  format(nrow(buoy_data), big.mark = ","),
  min(buoy_daily$date, na.rm = TRUE),
  max(buoy_daily$date, na.rm = TRUE)
))
```

## Pivot Table

The Perspective widget runs entirely in the browser via WebAssembly.
You can:

- **Group** rows by station, year, month
- **Split** columns by station
- **Filter** by any column
- **Sort** by clicking column headers
- **Switch views**: datagrid, line chart, bar chart, heatmap

```{r perspective-pivot}
if (requireNamespace("perspectiveR", quietly = TRUE)) {
  perspectiveR::perspective(
    buoy_daily,
    group_by = "station_id",
    columns = c("mean_wave_height", "max_hmax", "mean_wind_speed",
                "max_gust", "mean_sea_temp", "n_obs"),
    aggregates = list(
      mean_wave_height = "avg",
      max_hmax = "max",
      mean_wind_speed = "avg",
      max_gust = "max",
      mean_sea_temp = "avg",
      n_obs = "sum"
    )
  )
} else {
  cat("perspectiveR not installed.")
}
```

## Time Series View

Daily mean wave height by station — drag and group to explore patterns.

```{r perspective-timeseries}
if (requireNamespace("perspectiveR", quietly = TRUE)) {
  perspectiveR::perspective(
    buoy_daily[, c("date", "station_id", "mean_wave_height",
                   "max_hmax", "mean_sea_temp")],
    group_by = "date",
    split_by = "station_id",
    columns = c("mean_wave_height", "max_hmax", "mean_sea_temp"),
    aggregates = list(
      mean_wave_height = "avg",
      max_hmax = "max",
      mean_sea_temp = "avg"
    ),
    plugin = "Y Line"
  )
} else {
  cat("perspectiveR not installed.")
}
```

## Size Comparison

```{r size-comparison, echo=FALSE}
current_size <- file.size("../docs/vignettes/dashboard_static.html")
perspective_size <- file.size("../docs/vignettes/dashboard_perspective.html")
cat(sprintf(
  "| Dashboard | HTML Size | Data Rows | Interactivity |\n|---|---|---|---|\n| Current (dygraphs) | %s MB | %s (hourly) | Zoom only |\n| This (perspectiveR) | %s MB | %s (daily) | Pivot, filter, chart type |\n| **Reduction** | **%.0f%%** | **%.0f%%** | — |\n",
  round(current_size / 1e6, 1),
  format(nrow(buoy_data), big.mark = ","),
  round(perspective_size / 1e6, 1),
  format(nrow(buoy_daily), big.mark = ","),
  100 * (1 - perspective_size / current_size),
  100 * (1 - nrow(buoy_daily) / nrow(buoy_data))
))
```
