---
title: "Irish Buoys Explorer"
format:
  dashboard:
    orientation: rows
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(dygraphs)
library(xts)
library(tidyr)

# Load pre-computed data from targets pipeline
data_dir <- "../inst/extdata"
if (!dir.exists(data_dir)) {
  data_dir <- system.file("extdata", package = "irishbuoys")
}

buoy_data <- readRDS(file.path(data_dir, "dashboard_buoy_data.rds"))
dashboard_stats <- readRDS(file.path(data_dir, "dashboard_stats.rds"))
dashboard_ts <- readRDS(file.path(data_dir, "dashboard_timeseries.rds"))

# Get package version
pkg_version <- tryCatch(
  as.character(packageVersion("irishbuoys")),
  error = function(e) {
    desc_file <- "../DESCRIPTION"
    if (file.exists(desc_file)) {
      desc <- read.dcf(desc_file)
      desc[1, "Version"]
    } else "dev"
  }
)

# Get unique stations
stations <- sort(unique(buoy_data$station_id))

# Use HOURLY data directly - preserve full detail
# Only sample if dataset is extremely large (>200k rows) to keep HTML manageable
set.seed(42)
if (nrow(buoy_data) > 200000) {
  buoy_hourly <- buoy_data |> slice_sample(n = 200000)
} else {
  buoy_hourly <- buoy_data
}

# Human-readable observation count (e.g., "200k")
format_k <- function(n) {
  if (n >= 1000) paste0(round(n / 1000), "k") else as.character(n)
}
n_hourly_display <- format_k(nrow(buoy_hourly))

# Calculate summary stats
valid_ww <- complete.cases(buoy_data$wind_speed, buoy_data$wave_height)
valid_wh <- complete.cases(buoy_data$wave_height, buoy_data$hmax)
cor_wind_wave <- round(cor(buoy_data$wind_speed[valid_ww], buoy_data$wave_height[valid_ww]), 3)
cor_wave_hmax <- round(cor(buoy_data$wave_height[valid_wh], buoy_data$hmax[valid_wh]), 3)
days_span <- round(as.numeric(difftime(max(buoy_data$time), min(buoy_data$time), units = "days")))
generated_time <- format(Sys.time(), "%Y-%m-%d %H:%M UTC")
date_caption <- paste0(format(min(buoy_data$time), "%Y-%m-%d"), " to ", format(max(buoy_data$time), "%Y-%m-%d"))
n_records <- format(nrow(buoy_data), big.mark = ",")

# Helper function to create station-specific data
get_station_data <- function(data, station) {
  if (station != "All Buoys") {
    data <- data |> filter(station_id == station)
  }
  data
}

# Helper function to create dygraph time series - full width
# Maps variable names to display names (removes hmax from legends)
var_display_names <- c(
  "hmax" = "Max Wave (m)",
  "wave_height" = "Signif Wave (m)",
  "wind_speed" = "Wind Speed (kn)",
  "gust" = "Gust (kn)"
)

create_dygraph <- function(data, variable = "wave_height", title = "", by_station = TRUE) {
  display_name <- if (variable %in% names(var_display_names)) var_display_names[[variable]] else variable

  if (by_station && length(unique(data$station_id)) > 1) {
    wide_data <- data |>
      select(time, station_id, all_of(variable)) |>
      tidyr::pivot_wider(names_from = station_id, values_from = all_of(variable))
    ts_data <- xts(wide_data[, -1], order.by = wide_data$time)
  } else {
    ts_data <- xts(data[[variable]], order.by = data$time)
    colnames(ts_data) <- display_name
  }

  # Calculate y-axis range to ensure all data including max values are visible
  y_max <- max(ts_data, na.rm = TRUE) * 1.05  # 5% padding above max
  y_min <- min(0, min(ts_data, na.rm = TRUE))  # Start at 0 or below if negative

  dygraph(ts_data, main = title, width = "100%") |>
    dyOptions(
      labelsUTC = TRUE,
      fillGraph = TRUE,
      fillAlpha = 0.1,
      drawGrid = TRUE,
      drawPoints = FALSE,
      strokeWidth = 1.5,
      colors = RColorBrewer::brewer.pal(max(3, ncol(ts_data)), "Set1")[1:ncol(ts_data)]
    ) |>
    dyAxis("y", valueRange = c(y_min, y_max)) |>
    dyRangeSelector(height = 40) |>
    dyCrosshair(direction = "vertical") |>
    dyHighlight(highlightCircleSize = 5, highlightSeriesBackgroundAlpha = 0.2, hideOnMouseOut = FALSE) |>
    dyLegend(show = "always", width = 400)
}

# Helper function for DT tables with optimized timestamp display and frozen headers
create_dt <- function(data, caption = "", time_width = "140px") {
  time_cols <- which(names(data) %in% c("time", "Time", "timestamp", "Timestamp", "datetime", "Time (UTC)", "First", "Last"))

  dt <- datatable(
    data,
    caption = caption,
    filter = 'top',
    rownames = FALSE,
    extensions = c('Buttons', 'FixedHeader'),
    options = list(
      pageLength = 15,
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel'),
      order = list(list(0, 'desc')),
      scrollX = TRUE,
      fixedHeader = TRUE,
      columnDefs = if (length(time_cols) > 0) {
        list(list(width = time_width, targets = time_cols - 1))
      } else NULL
    ),
    class = 'cell-border stripe hover compact'
  )

  if (length(time_cols) > 0) {
    dt <- dt %>%
      DT::formatStyle(columns = time_cols, fontSize = '10px', whiteSpace = 'nowrap')
  }

  dt
}

# Station information with coordinates and depths
station_info <- data.frame(
  Station = c("M2", "M3", "M4", "M5", "M6"),
  Location = c("Southwest Ireland", "Southwest Ireland", "Southeast Ireland", "West Ireland", "Northwest Ireland"),
  Latitude = c(51.22, 51.22, 51.69, 51.69, 53.07),
  Longitude = c(-9.99, -10.55, -6.70, -10.00, -15.93),
  `Water Depth (m)` = c(155, 155, 62, 70, 3000),
  `Distance from Shore (km)` = c(50, 65, 30, 55, 320),
  check.names = FALSE
)

# Pre-compute max values for captions with dates
max_hmax_row <- buoy_data |> filter(hmax == max(hmax, na.rm = TRUE)) |> slice(1)
max_hmax <- round(max_hmax_row$hmax, 1)
max_hmax_date <- format(max_hmax_row$time, "%Y-%m-%d %H:%M")
max_station_hmax <- max_hmax_row$station_id

max_signif_row <- buoy_data |> filter(wave_height == max(wave_height, na.rm = TRUE)) |> slice(1)
max_signif <- round(max_signif_row$wave_height, 1)
max_signif_date <- format(max_signif_row$time, "%Y-%m-%d %H:%M")
max_station_signif <- max_signif_row$station_id

# Compute max values per station for highlighting
max_per_station_hmax <- buoy_data |>
  group_by(station_id) |>
  filter(hmax == max(hmax, na.rm = TRUE)) |>
  slice(1) |>
  ungroup()

max_per_station_signif <- buoy_data |>
  group_by(station_id) |>
  filter(wave_height == max(wave_height, na.rm = TRUE)) |>
  slice(1) |>
  ungroup()

# Rogue waves and rogue winds for Data tab
rogue_waves_full <- buoy_data |>
  filter(!is.na(hmax), !is.na(wave_height), wave_height > 0) |>
  mutate(wave_ratio = hmax / wave_height) |>
  filter(wave_ratio > 2.0)

rogue_winds_full <- buoy_data |>
  filter(!is.na(gust), !is.na(wind_speed), wind_speed > 0) |>
  mutate(gust_ratio = gust / wind_speed) |>
  filter(gust_ratio > 1.8)
```

# Max Wave (m)

## Time Series

### Station Views {.tabset}

#### All Stations

```{r}
#| title: "Max Wave (m): Tallest Individual Wave per Hour - All 5 Stations"
create_dygraph(buoy_hourly, "hmax", paste0("Max Wave (m) - Hourly Data (", date_caption, ")"))
```

**Caption:** Time series of [Max Wave](#reference) (m) showing the tallest individual wave recorded during each hourly 17.5-min measurement window. **Hourly data (`r n_hourly_display` observations)** - no daily averaging to preserve full detail. The highest Max Wave (**`r max_hmax` m**) was recorded at **`r max_station_hmax`** on **`r max_hmax_date` UTC** (zoom to this date to see it). M6 (deep Atlantic, 3000m depth) typically shows larger waves due to open ocean exposure.

#### M2

```{r}
#| title: "Max Wave (m) at M2 - Southwest Ireland (155m depth, 50km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M2"), "hmax", "Max Wave (m) - M2 Hourly", by_station = FALSE)
```

#### M3

```{r}
#| title: "Max Wave (m) at M3 - Southwest Ireland (155m depth, 65km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M3"), "hmax", "Max Wave (m) - M3 Hourly", by_station = FALSE)
```

#### M4

```{r}
#| title: "Max Wave (m) at M4 - Southeast Ireland (62m depth, 30km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M4"), "hmax", "Max Wave (m) - M4 Hourly", by_station = FALSE)
```

#### M5

```{r}
#| title: "Max Wave (m) at M5 - West Ireland (70m depth, 55km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M5"), "hmax", "Max Wave (m) - M5 Hourly", by_station = FALSE)
```

#### M6

```{r}
#| title: "Max Wave (m) at M6 - Northwest Atlantic (3000m depth, 320km offshore)"
create_dygraph(get_station_data(buoy_hourly, "M6"), "hmax", "Max Wave (m) - M6 Hourly", by_station = FALSE)
```

## Station Comparison {height=35%}

### Max Wave Details {.tabset}

#### Statistics

```{r}
# Get date of max wave per station
max_wave_dates <- buoy_data |>
  group_by(station_id) |>
  filter(hmax == max(hmax, na.rm = TRUE)) |>
  slice(1) |>
  select(station_id, max_wave_time = time) |>
  ungroup()

hmax_stats <- buoy_data |>
  group_by(`Station` = station_id) |>
  summarise(
    `Records (hourly)` = format(n(), big.mark = ","),
    `Max Wave (m)` = round(max(hmax, na.rm = TRUE), 2),
    `Avg Max Wave (m)` = round(mean(hmax, na.rm = TRUE), 2),
    `Max Signif (m)` = round(max(wave_height, na.rm = TRUE), 2),
    `Avg Ratio` = round(mean(hmax / wave_height, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  left_join(max_wave_dates, by = c("Station" = "station_id")) |>
  mutate(`Max Wave Date` = format(max_wave_time, "%Y-%m-%d %H:%M")) |>
  select(Station, `Records (hourly)`, `Max Wave Date`, `Max Wave (m)`, `Avg Max Wave (m)`, `Max Signif (m)`, `Avg Ratio`) |>
  arrange(desc(`Max Wave (m)`))

create_dt(hmax_stats, paste0(
  "Max Wave Statistics by Station (", date_caption, "). ",
  "Max Wave = tallest individual wave in each 17.5-min hourly window. ",
  "Avg Max Wave = mean of all hourly Max Wave values (typical tallest wave per hour). ",
  "Avg Ratio = mean(Max Wave / Signif Wave); typical 1.5-1.9, >2.0 = rogue conditions. ",
  "Conclusion: M6 records highest waves due to deep Atlantic exposure. See [Variable Definitions](#variable-definitions)."
))
```

#### Data Table

```{r}
hmax_data <- buoy_data |>
  select(time, station_id, hmax, wave_height) |>
  mutate(ratio = round(hmax / wave_height, 2)) |>
  arrange(desc(hmax)) |>
  head(5000) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    `Max Wave (m)` = round(hmax, 2),
    `Signif Wave (m)` = round(wave_height, 2),
    `Ratio` = ratio
  ) |>
  select(`Time (UTC)`, Station = station_id, `Max Wave (m)`, `Signif Wave (m)`, `Ratio`)

create_dt(hmax_data, paste0(
  "Max Wave Data ordered by tallest waves (top 5000 of ", n_records, " hourly records). ",
  "Max Wave = single tallest wave. Signif Wave = mean of top 1/3. ",
  "Ratio > 2.0 = potential rogue wave. Conclusion: Extreme waves cluster in winter storms."
))
```

# Signif Wave (m)

## Time Series

### Station Views {.tabset}

#### All Stations

```{r}
#| title: "Signif Wave (m): Mean of Top 1/3 Waves per Hour - All 5 Stations"
create_dygraph(buoy_hourly, "wave_height", paste0("Signif Wave (m) - Hourly Data (", date_caption, ")"))
```

**Caption:** Time series of [Signif Wave](#reference) (m) showing the mean of the highest 1/3 of waves measured during each hourly 17.5-min window. **Hourly data (`r n_hourly_display` observations)** - no daily averaging to preserve full detail. This metric smooths out individual wave variability while capturing sea state severity. Max Signif Wave: **`r max_signif` m** at **`r max_station_signif`** on **`r max_signif_date` UTC** (zoom to this date). Clear seasonal pattern with higher values in winter (Oct-Mar).

#### M2

```{r}
#| title: "Signif Wave (m) at M2 - Southwest Ireland (155m depth)"
create_dygraph(get_station_data(buoy_hourly, "M2"), "wave_height", "Signif Wave (m) - M2 Hourly", by_station = FALSE)
```

#### M3

```{r}
#| title: "Signif Wave (m) at M3 - Southwest Ireland (155m depth)"
create_dygraph(get_station_data(buoy_hourly, "M3"), "wave_height", "Signif Wave (m) - M3 Hourly", by_station = FALSE)
```

#### M4

```{r}
#| title: "Signif Wave (m) at M4 - Southeast Ireland (62m depth)"
create_dygraph(get_station_data(buoy_hourly, "M4"), "wave_height", "Signif Wave (m) - M4 Hourly", by_station = FALSE)
```

#### M5

```{r}
#| title: "Signif Wave (m) at M5 - West Ireland (70m depth)"
create_dygraph(get_station_data(buoy_hourly, "M5"), "wave_height", "Signif Wave (m) - M5 Hourly", by_station = FALSE)
```

#### M6

```{r}
#| title: "Signif Wave (m) at M6 - Northwest Atlantic (3000m depth)"
create_dygraph(get_station_data(buoy_hourly, "M6"), "wave_height", "Signif Wave (m) - M6 Hourly", by_station = FALSE)
```

## Station Comparison {height=35%}

### Signif Wave Details {.tabset}

#### Statistics

```{r}
# Get date of max signif wave per station
max_signif_dates <- buoy_data |>
  group_by(station_id) |>
  filter(wave_height == max(wave_height, na.rm = TRUE)) |>
  slice(1) |>
  select(station_id, max_signif_time = time) |>
  ungroup()

wave_stats <- buoy_data |>
  group_by(`Station` = station_id) |>
  summarise(
    `Records (hourly)` = format(n(), big.mark = ","),
    `Max Wave (m)` = round(max(hmax, na.rm = TRUE), 2),
    `Max Signif (m)` = round(max(wave_height, na.rm = TRUE), 2),
    `Avg Signif (m)` = round(mean(wave_height, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  left_join(max_signif_dates, by = c("Station" = "station_id")) |>
  mutate(`Max Signif Date` = format(max_signif_time, "%Y-%m-%d %H:%M")) |>
  select(Station, `Records (hourly)`, `Max Signif Date`, `Max Wave (m)`, `Max Signif (m)`, `Avg Signif (m)`) |>
  arrange(desc(`Max Signif (m)`))

create_dt(wave_stats, paste0(
  "Signif Wave Statistics by Station (", date_caption, "). ",
  "Signif Wave = mean of highest 1/3 of ~300 waves per 17.5-min window (hourly). ",
  "Max Wave = tallest single wave. ",
  "Conclusion: M6 has highest average Signif Wave due to Atlantic swell exposure. See [Variable Definitions](#variable-definitions)."
))
```

#### Data Table

```{r}
wave_data <- buoy_data |>
  select(time, station_id, wave_height, hmax, wave_period) |>
  arrange(desc(wave_height)) |>
  head(5000) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    `Signif Wave (m)` = round(wave_height, 2),
    `Max Wave (m)` = round(hmax, 2),
    `Wave Period (s)` = round(wave_period, 1)
  ) |>
  select(`Time (UTC)`, Station = station_id, `Signif Wave (m)`, `Max Wave (m)`, `Wave Period (s)`)

create_dt(wave_data, paste0(
  "Signif Wave Data ordered by largest values (top 5000 of ", n_records, " hourly records). ",
  "Signif Wave = mean of top 1/3. Wave Period = average time between wave crests (s). ",
  "Conclusion: Highest Signif Waves typically have longer periods (>10s), indicating Atlantic swell."
))
```

# Wind Speed (kn)

## Time Series

### Station Views {.tabset}

#### All Stations

```{r}
#| title: "Wind Speed (knots): 10-min Average per Hour - All 5 Stations"
create_dygraph(buoy_hourly, "wind_speed", paste0("Wind Speed (knots) - Hourly Data (", date_caption, ")"))
```

**Caption:** Time series of [Wind Speed](#reference) (knots) showing the 10-minute average wind speed recorded each hour. **Hourly data (`r n_hourly_display` observations)** - no daily averaging to preserve full detail. Wind measurements are taken at ~3m above sea level. Gust speeds (not shown here) can be 40-60% higher. Strong correlation with [Signif Wave](#reference) (r = `r cor_wind_wave`).

#### M2

```{r}
#| title: "Wind Speed (kn) at M2 - Southwest Ireland"
create_dygraph(get_station_data(buoy_hourly, "M2"), "wind_speed", "Wind Speed (kn) - M2 Hourly", by_station = FALSE)
```

#### M3

```{r}
#| title: "Wind Speed (kn) at M3 - Southwest Ireland"
create_dygraph(get_station_data(buoy_hourly, "M3"), "wind_speed", "Wind Speed (kn) - M3 Hourly", by_station = FALSE)
```

#### M4

```{r}
#| title: "Wind Speed (kn) at M4 - Southeast Ireland"
create_dygraph(get_station_data(buoy_hourly, "M4"), "wind_speed", "Wind Speed (kn) - M4 Hourly", by_station = FALSE)
```

#### M5

```{r}
#| title: "Wind Speed (kn) at M5 - West Ireland"
create_dygraph(get_station_data(buoy_hourly, "M5"), "wind_speed", "Wind Speed (kn) - M5 Hourly", by_station = FALSE)
```

#### M6

```{r}
#| title: "Wind Speed (kn) at M6 - Northwest Atlantic"
create_dygraph(get_station_data(buoy_hourly, "M6"), "wind_speed", "Wind Speed (kn) - M6 Hourly", by_station = FALSE)
```

## Station Comparison {height=35%}

### Wind Details {.tabset}

#### Statistics

```{r}
wind_stats <- buoy_data |>
  group_by(`Station` = station_id) |>
  summarise(
    `Records (hourly)` = n(),
    `Avg Wind (kn)` = round(mean(wind_speed, na.rm = TRUE), 1),
    `Max Wind (kn)` = round(max(wind_speed, na.rm = TRUE), 1),
    `Max Gust (kn)` = round(max(gust, na.rm = TRUE), 1),
    `Gust Factor` = round(max(gust, na.rm = TRUE) / max(wind_speed, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(desc(`Max Gust (kn)`))

create_dt(wind_stats, paste0(
  "Wind Statistics by Station (", date_caption, "). ",
  "Wind Speed = 10-min average (hourly). Gust = peak 3-second wind (hourly). ",
  "Gust Factor = Gust/Wind ratio. ",
  "Conclusion: M6 experiences strongest sustained winds due to open Atlantic exposure. See [Variable Definitions](#variable-definitions)."
))
```

#### Data Table

```{r}
wind_data <- buoy_data |>
  select(time, station_id, wind_speed, gust, atmospheric_pressure) |>
  arrange(desc(gust)) |>
  head(5000) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    `Wind Speed (kn)` = round(wind_speed, 1),
    `Gust (kn)` = round(gust, 1),
    `Pressure (hPa)` = round(atmospheric_pressure, 1)
  ) |>
  select(`Time (UTC)`, Station = station_id, `Wind Speed (kn)`, `Gust (kn)`, `Pressure (hPa)`)

create_dt(wind_data, paste0(
  "Wind Data ordered by strongest gusts (top 5000 of ", n_records, " hourly records). ",
  "Gust = peak 3-second wind. Low pressure often accompanies high winds. ",
  "Conclusion: Strongest gusts cluster in winter storm events with pressure <990 hPa."
))
```

# Wind vs Wave

## Scatter Analysis

### Regression Plots {width=100%}

```{r}
#| fig-height: 36
# Compute regression slopes to order panels (6x1 layout needs more height)
reg_stats <- buoy_data |>
  filter(!is.na(wind_speed), !is.na(wave_height)) |>
  group_by(Station = station_id) |>
  summarise(
    Records = n(),
    Correlation = round(cor(wind_speed, wave_height), 2),
    Intercept = round(coef(lm(wave_height ~ wind_speed))[1], 2),
    Slope = round(coef(lm(wave_height ~ wind_speed))[2], 2),
    .groups = "drop"
  ) |>
  arrange(desc(Slope))

# Order stations by slope (largest first)
station_order <- reg_stats$Station
panel_labels <- paste0(station_order, " (slope=", sprintf("%.2f", reg_stats$Slope), ")")

# Filter for wind_speed > 20 knots and sample for scatter plot
set.seed(42)
scatter_sample <- buoy_data |>
  filter(!is.na(wind_speed), !is.na(wave_height), wind_speed > 20)
n_strong_wind <- nrow(scatter_sample)
if (nrow(scatter_sample) > 15000) {
  scatter_sample <- scatter_sample |> slice_sample(n = 15000)
}
plot_data <- scatter_sample |>
  mutate(
    station_label = factor(
      paste0(station_id, " (slope=", sprintf("%.2f", reg_stats$Slope[match(station_id, reg_stats$Station)]), ")"),
      levels = panel_labels
    )
  )

p <- ggplot(plot_data, aes(x = wind_speed, y = wave_height)) +
  geom_point(aes(
    text = paste0("Station: ", station_id, "<br>",
                  "Wind Speed: ", round(wind_speed, 1), " knots<br>",
                  "Signif Wave: ", round(wave_height, 2), " m<br>",
                  "Time: ", format(time, "%Y-%m-%d %H:%M"))
  ), color = "#00bfff", alpha = 0.7, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "#ff4444", linewidth = 1) +
  facet_wrap(~station_label, ncol = 1) +
  labs(
    x = "Wind Speed (knots) - 10-min average",
    y = "Signif Wave (m) - mean of top 1/3 waves"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 14, color = "#1a1a1a"),
    strip.background = element_rect(fill = "#999999", color = NA),
    panel.background = element_rect(fill = "#B3B3B3", color = NA),
    plot.background = element_rect(fill = "#B3B3B3", color = NA),
    panel.grid.major = element_line(color = "#666666", linewidth = 0.3),
    panel.grid.minor = element_line(color = "#808080", linewidth = 0.1),
    axis.text = element_text(color = "#1a1a1a"),
    axis.title = element_text(color = "#1a1a1a")
  )

ggplotly(p, tooltip = "text") |>
  layout(
    title = list(
      text = paste0("Wind Speed vs Signif Wave (wind >20 kn, n=", format_k(n_strong_wind), ", ", date_caption, ")"),
      font = list(size = 14, color = "#1a1a1a"),
      y = 0.98
    ),
    paper_bgcolor = "#B3B3B3",
    plot_bgcolor = "#B3B3B3",
    margin = list(t = 80),
    hoverlabel = list(bgcolor = "white", font = list(color = "black", size = 12))
  )
```

**Caption:** Wind Speed vs [Signif Wave](#variable-definitions) relationship for observations with **wind >20 knots** (strong wind conditions). Panels ordered by regression slope. The slope indicates wave height increase (m) per knot of wind. M6 (deep Atlantic) shows steepest slope due to unlimited fetch. Filtering for high winds focuses on storm conditions where wind-wave coupling is strongest.

## Summary Statistics {height=25%}

### Wind-Wave Details {.tabset}

#### Regression Statistics

```{r}
create_dt(reg_stats, paste0(
  "Linear Regression by Station ordered by slope (", date_caption, "). ",
  "Slope = Signif Wave increase (m) per knot of wind. ",
  "Higher correlation = wind explains more wave variability. ",
  "Conclusion: Deep-water stations (M6) show strongest wind-wave coupling."
))
```

#### Data Sample

```{r}
scatter_data <- buoy_data |>
  select(time, station_id, wind_speed, wave_height, hmax) |>
  filter(!is.na(wind_speed), !is.na(wave_height)) |>
  arrange(desc(wave_height)) |>
  head(5000) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    `Wind Speed (kn)` = round(wind_speed, 1),
    `Signif Wave (m)` = round(wave_height, 2),
    `Max Wave (m)` = round(hmax, 2)
  ) |>
  select(`Time (UTC)`, Station = station_id, `Wind Speed (kn)`, `Signif Wave (m)`, `Max Wave (m)`)

create_dt(scatter_data, paste0(
  "Wind-Wave Data ordered by Signif Wave (top 5000 of ", n_records, " hourly records). ",
  "Shows hourly measurements used in regression analysis. ",
  "Conclusion: High Signif Waves generally coincide with high winds, but swell can arrive without local wind."
))
```

# Forecasts

```{r forecasts-setup}
return_levels <- tryCatch(
  readRDS(file.path(data_dir, "dashboard_return_levels.rds")),
  error = function(e) NULL
)
```

## Forecast Tables

### Return Level Forecasts {.tabset}

#### 1-Year Return Level

```{r forecast-1yr}
if (!is.null(return_levels)) {
  rl_1yr <- return_levels |>
    filter(return_period == 1, !is.na(return_level)) |>
    mutate(return_level = round(return_level, 2)) |>
    select(station, variable_label, return_level) |>
    tidyr::pivot_wider(names_from = variable_label, values_from = return_level)

  create_dt(rl_1yr, "1-Year Return Levels by Station (GPD, 95th percentile threshold)")
} else {
  htmltools::div(class = "alert alert-info", "Return level data not yet computed. Run tar_make() to generate.")
}
```

#### 5-Year Return Level

```{r forecast-5yr}
if (!is.null(return_levels)) {
  rl_5yr <- return_levels |>
    filter(return_period == 5, !is.na(return_level)) |>
    mutate(return_level = round(return_level, 2)) |>
    select(station, variable_label, return_level) |>
    tidyr::pivot_wider(names_from = variable_label, values_from = return_level)

  create_dt(rl_5yr, "5-Year Return Levels by Station (GPD, 95th percentile threshold)")
} else {
  htmltools::div(class = "alert alert-info", "Return level data not yet computed. Run tar_make() to generate.")
}
```

#### 10-Year Return Level

```{r forecast-10yr}
if (!is.null(return_levels)) {
  rl_10yr <- return_levels |>
    filter(return_period == 10, !is.na(return_level)) |>
    mutate(return_level = round(return_level, 2)) |>
    select(station, variable_label, return_level) |>
    tidyr::pivot_wider(names_from = variable_label, values_from = return_level)

  create_dt(rl_10yr, "10-Year Return Levels by Station (GPD, 95th percentile threshold)")
} else {
  htmltools::div(class = "alert alert-info", "Return level data not yet computed. Run tar_make() to generate.")
}
```

#### All Return Levels

```{r forecast-all}
if (!is.null(return_levels)) {
  rl_wide <- return_levels |>
    filter(!is.na(return_level)) |>
    mutate(
      col_name = paste0(variable_label, " ", return_period, "yr"),
      return_level = round(return_level, 2)
    ) |>
    select(station, col_name, return_level) |>
    tidyr::pivot_wider(names_from = col_name, values_from = return_level)

  create_dt(rl_wide, "All Return Levels by Station (GPD, 95th percentile threshold)")
} else {
  htmltools::div(class = "alert alert-info", "Return level data not yet computed. Run tar_make() to generate.")
}
```

#### Method

**GPD Return Level Methodology:**

- **Method:** Generalized Pareto Distribution (GPD) via `mev::fit.gpd()`
- **Threshold:** 95th percentile of each variable per station
- **Variables:** Avg Wave (m), Rogue Wave (m), Avg Wind (kn), Wind Gust (kn)
- **Return periods:** 1, 5, and 10 years
- **Confidence intervals:** 95% via delta method

For detailed methodology and threshold sensitivity analysis, see the [Wave Analysis](wave_analysis.html) vignette.

# Data

## Data Table

### Rogue Events {.tabset}

#### Rogue Events (Waves + Winds)

```{r}
# Merge rogue waves (ratio > 2.0) with rogue winds (gust factor > 1.8)
# These are the extreme events of interest

# Filter: wind_speed > 20 knots AND wave_height > 2 m (significant storm conditions)
rogue_merged <- buoy_data |>
  filter(!is.na(hmax), !is.na(wave_height), wave_height > 2,
         !is.na(gust), !is.na(wind_speed), wind_speed > 20) |>
  mutate(
    wave_ratio = hmax / wave_height,
    gust_ratio = gust / wind_speed
  ) |>
  filter(wave_ratio > 2.0 | gust_ratio > 1.8) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    Station = station_id,
    `Max Wave (m)` = round(hmax, 2),
    `Signif Wave (m)` = round(wave_height, 2),
    `Wave Ratio` = round(wave_ratio, 2),
    `Rogue Wave?` = ifelse(wave_ratio > 2.0, "Yes", "No"),
    `Gust (kn)` = round(gust, 1),
    `Wind (kn)` = round(wind_speed, 1),
    `Gust Ratio` = round(gust_ratio, 2),
    `Rogue Wind?` = ifelse(gust_ratio > 1.8, "Yes", "No")
  ) |>
  arrange(desc(wave_ratio), desc(gust_ratio)) |>
  select(`Time (UTC)`, Station, `Max Wave (m)`, `Signif Wave (m)`, `Wave Ratio`, `Rogue Wave?`,
         `Gust (kn)`, `Wind (kn)`, `Gust Ratio`, `Rogue Wind?`)

n_rogue_merged <- nrow(rogue_merged)

datatable(
  rogue_merged,
  caption = htmltools::HTML(paste0(
    "<b>Rogue Events in Storm Conditions</b> (", format(n_rogue_merged, big.mark = ","), " events). ",
    "Filtered: <b>wind >20 kn</b> AND <b>Signif Wave >2 m</b>. ",
    "Shows <b>Rogue Waves</b> (Wave Ratio > 2.0) and/or <b>Rogue Winds</b> (Gust Ratio > 1.8). ",
    "Focusing on significant storm events where extreme waves/gusts are most dangerous. ",
    "See <a href='#variable-definitions'>Variable Definitions</a>."
  )),
  filter = 'top',
  rownames = FALSE,
  extensions = c('Buttons', 'FixedHeader'),
  options = list(
    pageLength = 25,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel'),
    order = list(list(4, 'desc')),
    scrollX = TRUE,
    fixedHeader = TRUE,
    columnDefs = list(list(width = '100px', targets = 0))
  ),
  class = 'cell-border stripe hover compact'
) %>%
  DT::formatStyle(columns = 1, fontSize = '10px', whiteSpace = 'nowrap')
```

#### Rogue Waves (Ratio > 2.0)

```{r}
rogue_waves <- buoy_data |>
  filter(!is.na(hmax), !is.na(wave_height), wave_height > 0) |>
  mutate(ratio = hmax / wave_height) |>
  filter(ratio > 2.0) |>
  arrange(desc(ratio)) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    Station = station_id,
    `Max Wave (m)` = round(hmax, 2),
    `Signif Wave (m)` = round(wave_height, 2),
    `Ratio` = round(ratio, 2),
    `Wind (kn)` = round(wind_speed, 1),
    `Gust (kn)` = round(gust, 1)
  ) |>
  select(`Time (UTC)`, Station, `Max Wave (m)`, `Signif Wave (m)`, `Ratio`, `Wind (kn)`, `Gust (kn)`)

n_rogue <- nrow(rogue_waves)

datatable(
  rogue_waves,
  caption = htmltools::HTML(paste0(
    "<b>Rogue Wave Events</b> (", format(n_rogue, big.mark = ","), " events where Ratio > 2.0). ",
    "A rogue wave is defined as [Max Wave](#variable-definitions) > 2× [Signif Wave](#variable-definitions). ",
    "These are potentially dangerous individual waves that exceed normal sea state. ",
    "Sorted by Ratio (most extreme first)."
  )),
  rownames = FALSE,
  filter = 'top',
  extensions = c('Buttons', 'FixedHeader'),
  options = list(
    pageLength = 25,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel'),
    order = list(list(4, 'desc')),
    scrollX = TRUE,
    fixedHeader = TRUE,
    columnDefs = list(list(width = '100px', targets = 0))
  ),
  class = 'cell-border stripe hover compact'
) %>%
  DT::formatStyle(columns = 1, fontSize = '10px', whiteSpace = 'nowrap')
```

#### Extreme Gusts (Top 1%)

```{r}
gust_threshold <- quantile(buoy_data$gust, 0.99, na.rm = TRUE)
extreme_gusts <- buoy_data |>
  filter(!is.na(gust), gust >= gust_threshold) |>
  arrange(desc(gust)) |>
  mutate(
    `Time (UTC)` = format(time, "%Y-%m-%d %H:%M"),
    Station = station_id,
    `Gust (kn)` = round(gust, 1),
    `Wind (kn)` = round(wind_speed, 1),
    `Gust Factor` = round(gust / wind_speed, 2),
    `Max Wave (m)` = round(hmax, 2),
    `Pressure (hPa)` = round(atmospheric_pressure, 1)
  ) |>
  select(`Time (UTC)`, Station, `Gust (kn)`, `Wind (kn)`, `Gust Factor`, `Max Wave (m)`, `Pressure (hPa)`)

n_gusts <- nrow(extreme_gusts)

datatable(
  extreme_gusts,
  caption = htmltools::HTML(paste0(
    "<b>Extreme Gust Events</b> (", format(n_gusts, big.mark = ","), " events in top 1%, ≥", round(gust_threshold, 1), " kn). ",
    "Gust Factor = Gust / Wind Speed; typical 1.3-1.6, >1.8 indicates gusty conditions. ",
    "Sorted by strongest gusts. Low pressure often accompanies extreme wind events."
  )),
  filter = 'top',
  extensions = c('Buttons', 'FixedHeader'),
  options = list(
    pageLength = 25,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel'),
    order = list(list(2, 'desc')),
    scrollX = TRUE,
    fixedHeader = TRUE,
    columnDefs = list(list(width = '100px', targets = 0))
  ),
  class = 'cell-border stripe hover compact'
) %>%
  DT::formatStyle(columns = 1, fontSize = '10px', whiteSpace = 'nowrap')
```

# Reference

## Technical Reference

### Reference Details {.tabset}

#### Variable Definitions

##### Variable Categories {.tabset}

###### Column Names

**Wave Measurements:**

| Variable | Display Name | Units | Definition |
|----------|--------------|-------|------------|
| `hmax` | **Max Wave** | m | Tallest individual wave (crest-to-trough) during each [17.5-min hourly measurement window](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html) |
| `wave_height` | **Signif Wave** | m | Mean height of the highest 1/3 of waves (~35 of ~105 waves) during each 17.5-min hourly window. [NOAA definition](https://www.weather.gov/key/marine_sigwave) |
| `wave_period` | **Wave Period** | s | Average time between successive wave crests via [zero-crossing analysis](https://www.nortekgroup.com/knowledge-center/wiki/new-to-waves) |

**Key distinction:** Max Wave is a single wave; Signif Wave is an average of the top ~35 waves (highest 1/3). Both come from the same 17.5-min measurement window each hour.

**Wind Measurements:**

| Variable | Display Name | Units | Definition |
|----------|--------------|-------|------------|
| `wind_speed` | **Wind Speed** | knots | 10-minute average wind speed measured each hour at ~3m above sea level |
| `gust` | **Gust** | knots | Peak 3-second wind speed recorded during the hour |

**Atmospheric:**

| Variable | Display Name | Units | Definition |
|----------|--------------|-------|------------|
| `atmospheric_pressure` | **Pressure** | hPa | Barometric pressure at sea level. Low pressure (<1000 hPa) often indicates storm systems |
| `air_temperature` | **Air Temp** | C | Air temperature at buoy (~3m above sea) |
| `sea_temperature` | **Sea Temp** | C | Sea surface temperature |

###### Aggregated Stats

**Computed Statistics (aggregated over all hourly records):**

| Statistic | Source Column | Definition |
|-----------|---------------|------------|
| **Max Wave (m)** | `hmax` | Highest `hmax` across all hours (largest single wave ever recorded in dataset) |
| **Max Signif (m)** | `wave_height` | Highest `wave_height` across all hours (highest hourly "average of top 1/3" value) |
| **Avg Signif (m)** | `wave_height` | Mean of all hourly `wave_height` values (typical sea state) |
| **Avg Max Wave (m)** | `hmax` | Mean of all hourly `hmax` values (typical tallest wave per hour) |
| **Avg Ratio** | `hmax / wave_height` | Mean ratio of Max Wave to Signif Wave. Typical: 1.5-1.9. >2.0 = rogue conditions |

**Why different?** Max Wave (m) and Max Signif (m) come from DIFFERENT columns:
- Max Wave = largest `hmax` = one extreme wave
- Max Signif = largest `wave_height` = one extreme hourly average

**Example:** A station might record Max Wave = 15.2m (one extreme individual wave) and Max Signif = 10.8m (one extreme hourly average of top 1/3 waves). These come from different measurement windows.

###### Rogue

**Rogue Waves:**

A **rogue wave** is an individual wave with height > 2× the Signif Wave (i.e., Wave Ratio > 2.0).

| Wave Ratio (Max Wave / Signif Wave) | Interpretation |
|-------------------------------------|----------------|
| 1.5 - 1.9 | Normal sea state |
| > 2.0 | Rogue wave - individual wave significantly exceeds average |

Rogue waves are rare but dangerous. They can appear without warning in otherwise moderate seas. The ratio helps identify when individual waves significantly exceed the average sea state.

**Important:** The raw ERDDAP data does NOT contain a rogue wave indicator. We CALCULATE the ratio ourselves: `rogue_ratio = hmax / wave_height`. If ratio > 2.0, we classify it as a rogue wave event.

**Rogue Winds:**

A **rogue wind** (or extreme gust) is a gust > 1.8× the mean wind speed (i.e., Gust Ratio > 1.8).

| Gust Ratio (Gust / Wind Speed) | Interpretation |
|--------------------------------|----------------|
| 1.3 - 1.6 | Normal conditions |
| > 1.8 | Rogue wind - gust significantly exceeds average |

High gust ratios indicate atmospheric instability, often from convective activity, frontal passages, or turbulent flow. Combined with rogue waves, these events represent the most dangerous marine conditions.

**Important:** Similarly, rogue winds are CALCULATED: `gust_ratio = gust / wind_speed`. If ratio > 1.8, we classify it as a rogue wind event.

###### Technical Terms

**Scientific notation used in literature (for reference only):**

| Scientific Term | Display Name | Definition |
|-----------------|--------------|------------|
| Hs, H1/3, Significant Wave Height | **Signif Wave** | Mean height of highest 1/3 of waves in a measurement window. Equivalent to 4× standard deviation (4σ) of sea surface elevation for narrow-band spectra. |
| Hmax, Maximum Wave Height | **Max Wave** | Tallest single wave (crest-to-trough) in measurement window |

This dashboard uses "Signif Wave" and "Max Wave" consistently instead of technical symbols (Hs, Hmax).

#### Methodology

##### Methodology Details {.tabset}

###### Measurement Window

Each buoy samples the sea surface **once per hour**. At the top of each hour, a [**17.5-minute measurement period**](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html) begins. During this window, the buoy measures approximately 105 individual wave heights (assuming ~10s wave period).

**Sampling limitation:** The buoy does NOT measure continuously. Only the 17.5-min window is sampled; waves occurring in the remaining 42.5 minutes of each hour are not recorded. Source: [Marine Institute ERDDAP metadata](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html) states *"The sensor measures this for 17.5 minutes and then takes an average."*

From this 17.5-min sample:

1. Waves are sorted by height, tallest first
2. **Max Wave**: The single tallest wave = `hmax`
3. **Signif Wave**: Mean of the highest 1/3 (~35 waves) = `wave_height` ([NOAA definition](https://www.weather.gov/key/marine_sigwave))

Both Max Wave and Signif Wave come from the SAME 17.5-min window, reported at the same timestamp.

###### Wind Measurement

**Wind Speed vs Gust:**

Wind measurements are taken at approximately 3 meters above sea level. Two distinct metrics are reported:

| Measurement | Averaging Period | Definition |
|-------------|-----------------|------------|
| **Wind Speed** | 10-minute average | Mean wind speed over a 10-minute period, measured once per hour |
| **Gust** | 3-second peak | Maximum 3-second average wind speed during the hour |

**Gust Factor:** The ratio of Gust to Wind Speed indicates atmospheric turbulence and stability:

| Gust Factor | Interpretation |
|-------------|----------------|
| 1.3 - 1.6 | Normal conditions |
| > 1.8 | Gusty/unstable conditions |

High gust factors often occur with convective activity, frontal passages, or flow over rough terrain/seas. Low pressure systems (<1000 hPa) frequently accompany extreme wind events.

###### Station Info

```{r}
create_dt(station_info, paste0(
  "Irish Weather Buoy Network Stations. ",
  "Water depth affects wave behavior: shallow (M4) vs deep (M6) stations show different wave characteristics. ",
  "Distance offshore affects wind exposure and fetch length. ",
  "M6 is in deep Atlantic waters (3000m depth, 320km offshore), experiencing open ocean conditions. ",
  "Other stations are on the continental shelf (62-155m depth, 30-65km offshore)."
))
```

#### Data Sources

**Primary Data:**

- [Marine Institute Ireland](https://www.marine.ie/) - Irish Weather Buoy Network operator
- [ERDDAP Data Portal](https://erddap.marine.ie/erddap/tabledap/IWBNetwork.html) - Access raw buoy data
- [ERDDAP Metadata](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html) - Dataset documentation

**Reference Documentation:**

- [NOAA Significant Wave Height Definition](https://www.weather.gov/key/marine_sigwave)
- [Nortek Wave Measurement Guide](https://www.nortekgroup.com/knowledge-center/wiki/new-to-waves) - Zero-crossing analysis explanation

**Code & Analysis:**

- [irishbuoys R Package](https://github.com/JohnGavin/irishbuoys) - Data processing and dashboard code

---

*Dashboard generated `r generated_time` using irishbuoys v`r pkg_version`*

*Data coverage: `r date_caption` | Total records: `r n_records` hourly observations*
