---
title: "Irish Weather Buoy Explorer"
subtitle: "Interactive Dashboard with Sortable Tables"
format:
  html:
    page-layout: full
    toc: true
    toc-location: left
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(jsonlite)
library(DT)

# Load the buoy data
data_path <- "data/buoy_data.json"
buoy_data <- fromJSON(data_path)
buoy_data$time <- as.POSIXct(buoy_data$time, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")

# Get unique stations
stations <- sort(unique(buoy_data$station_id))

# Variable definitions
VAR_CHOICES <- c(
  "Wave Height (m)" = "wave_height",
  "Max Wave Height (m)" = "hmax",
  "Wave Period (s)" = "wave_period",
  "Wind Speed (knots)" = "wind_speed",
  "Wind Gust (knots)" = "gust",
  "Air Temperature (C)" = "air_temperature",
  "Sea Temperature (C)" = "sea_temperature",
  "Pressure (hPa)" = "atmospheric_pressure"
)

# Helper function to create station-specific data
get_station_data <- function(data, station, days = 30) {
  cutoff <- max(data$time) - (days * 24 * 3600)
  if (station == "All Buoys") {
    data |> filter(time >= cutoff)
  } else {
    data |> filter(station_id == station, time >= cutoff)
  }
}

# Helper function to create a plot
create_time_series_plot <- function(data, variable = "wave_height", title_suffix = "") {
  var_label <- names(VAR_CHOICES)[VAR_CHOICES == variable]

  ggplot(data, aes(x = time, y = .data[[variable]], color = station_id)) +
    geom_line(alpha = 0.7) +
    geom_point(size = 0.5, alpha = 0.5) +
    labs(
      title = paste(var_label, title_suffix),
      x = "Time (UTC)",
      y = var_label,
      color = "Station"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Helper function for DT tables
create_dt <- function(data, caption = "") {
  datatable(
    data,
    caption = caption,
    filter = 'top',
    options = list(
      pageLength = 15,
      dom = 'Bfrtip',
      order = list(list(0, 'desc')),
      scrollX = TRUE
    ),
    class = 'cell-border stripe hover'
  )
}
```

## Data Overview

This dashboard displays data from the **Irish Weather Buoy Network**, operated by the Marine Institute of Ireland. Data sourced from the [Marine Institute ERDDAP Server](https://erddap.marine.ie/erddap/index.html).

```{r data-summary}
# Calculate data validation metrics
valid_ww <- complete.cases(buoy_data$wind_speed, buoy_data$wave_height)
valid_wh <- complete.cases(buoy_data$wave_height, buoy_data$hmax)
cor_wind_wave <- round(cor(buoy_data$wind_speed[valid_ww], buoy_data$wave_height[valid_ww]), 3)
cor_wave_hmax <- round(cor(buoy_data$wave_height[valid_wh], buoy_data$hmax[valid_wh]), 3)

days_span <- as.numeric(difftime(max(buoy_data$time), min(buoy_data$time), units = "days"))
```

| Metric | Value |
|--------|-------|
| Total Records | `r format(nrow(buoy_data), big.mark = ",")` |
| Date Range | `r format(min(buoy_data$time), "%Y-%m-%d")` to `r format(max(buoy_data$time), "%Y-%m-%d")` |
| Time Span | `r round(days_span, 1)` days |
| Stations | `r paste(stations, collapse = ", ")` |
| Wind-Wave Correlation | `r cor_wind_wave` |
| Wave-Hmax Correlation | `r cor_wave_hmax` |

---

## Wave Height by Station {#wave-height}

::: {.panel-tabset}

### All Buoys (Plot)

```{r}
#| fig-height: 5
stn_data <- get_station_data(buoy_data, "All Buoys", 30)
create_time_series_plot(stn_data, "wave_height", " - Last 30 Days")
```

### All Buoys (Data)

```{r}
stn_data <- get_station_data(buoy_data, "All Buoys", 30) |>
  select(time, station_id, wave_height, hmax, wind_speed, gust) |>
  mutate(
    time = format(time, "%Y-%m-%d %H:%M"),
    wave_height = round(wave_height, 2),
    hmax = round(hmax, 2),
    wind_speed = round(wind_speed, 1),
    gust = round(gust, 1)
  )
create_dt(stn_data, "All Stations - Last 30 Days")
```

### M2 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M2", 30)
create_time_series_plot(stn_data, "wave_height", " - M2")
```

### M2 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M2", 30) |>
  select(time, wave_height, hmax, wind_speed, gust, air_temperature) |>
  mutate(
    time = format(time, "%Y-%m-%d %H:%M"),
    wave_height = round(wave_height, 2),
    hmax = round(hmax, 2),
    wind_speed = round(wind_speed, 1),
    gust = round(gust, 1)
  )
create_dt(stn_data, "M2 - Last 30 Days")
```

### M3 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M3", 30)
create_time_series_plot(stn_data, "wave_height", " - M3")
```

### M3 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M3", 30) |>
  select(time, wave_height, hmax, wind_speed, gust, air_temperature) |>
  mutate(time = format(time, "%Y-%m-%d %H:%M"), across(c(wave_height, hmax), ~round(., 2)), across(c(wind_speed, gust), ~round(., 1)))
create_dt(stn_data, "M3 - Last 30 Days")
```

### M4 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M4", 30)
create_time_series_plot(stn_data, "wave_height", " - M4")
```

### M4 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M4", 30) |>
  select(time, wave_height, hmax, wind_speed, gust, air_temperature) |>
  mutate(time = format(time, "%Y-%m-%d %H:%M"), across(c(wave_height, hmax), ~round(., 2)), across(c(wind_speed, gust), ~round(., 1)))
create_dt(stn_data, "M4 - Last 30 Days")
```

### M5 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M5", 30)
create_time_series_plot(stn_data, "wave_height", " - M5")
```

### M5 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M5", 30) |>
  select(time, wave_height, hmax, wind_speed, gust, air_temperature) |>
  mutate(time = format(time, "%Y-%m-%d %H:%M"), across(c(wave_height, hmax), ~round(., 2)), across(c(wind_speed, gust), ~round(., 1)))
create_dt(stn_data, "M5 - Last 30 Days")
```

### M6 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M6", 30)
create_time_series_plot(stn_data, "wave_height", " - M6")
```

### M6 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M6", 30) |>
  select(time, wave_height, hmax, wind_speed, gust, air_temperature) |>
  mutate(time = format(time, "%Y-%m-%d %H:%M"), across(c(wave_height, hmax), ~round(., 2)), across(c(wind_speed, gust), ~round(., 1)))
create_dt(stn_data, "M6 - Last 30 Days")
```

:::

---

## Wind Speed by Station {#wind-speed}

::: {.panel-tabset}

### All Buoys (Plot)

```{r}
#| fig-height: 5
stn_data <- get_station_data(buoy_data, "All Buoys", 30)
create_time_series_plot(stn_data, "wind_speed", " - Last 30 Days")
```

### All Buoys (Data)

```{r}
stn_data <- get_station_data(buoy_data, "All Buoys", 30) |>
  select(time, station_id, wind_speed, gust, wave_height, atmospheric_pressure) |>
  mutate(
    time = format(time, "%Y-%m-%d %H:%M"),
    wind_speed = round(wind_speed, 1),
    gust = round(gust, 1),
    wave_height = round(wave_height, 2),
    atmospheric_pressure = round(atmospheric_pressure, 1)
  )
create_dt(stn_data, "Wind Speed - All Stations")
```

### M2 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M2", 30)
create_time_series_plot(stn_data, "wind_speed", " - M2")
```

### M2 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M2", 30) |>
  select(time, wind_speed, gust, wave_height, atmospheric_pressure) |>
  mutate(time = format(time, "%Y-%m-%d %H:%M"), across(where(is.numeric), ~round(., 1)))
create_dt(stn_data, "Wind Speed - M2")
```

### M3 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M3", 30)
create_time_series_plot(stn_data, "wind_speed", " - M3")
```

### M3 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M3", 30) |>
  select(time, wind_speed, gust, wave_height, atmospheric_pressure) |>
  mutate(time = format(time, "%Y-%m-%d %H:%M"), across(where(is.numeric), ~round(., 1)))
create_dt(stn_data, "Wind Speed - M3")
```

### M4 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M4", 30)
create_time_series_plot(stn_data, "wind_speed", " - M4")
```

### M4 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M4", 30) |>
  select(time, wind_speed, gust, wave_height, atmospheric_pressure) |>
  mutate(time = format(time, "%Y-%m-%d %H:%M"), across(where(is.numeric), ~round(., 1)))
create_dt(stn_data, "Wind Speed - M4")
```

### M5 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M5", 30)
create_time_series_plot(stn_data, "wind_speed", " - M5")
```

### M5 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M5", 30) |>
  select(time, wind_speed, gust, wave_height, atmospheric_pressure) |>
  mutate(time = format(time, "%Y-%m-%d %H:%M"), across(where(is.numeric), ~round(., 1)))
create_dt(stn_data, "Wind Speed - M5")
```

### M6 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M6", 30)
create_time_series_plot(stn_data, "wind_speed", " - M6")
```

### M6 (Data)

```{r}
stn_data <- get_station_data(buoy_data, "M6", 30) |>
  select(time, wind_speed, gust, wave_height, atmospheric_pressure) |>
  mutate(time = format(time, "%Y-%m-%d %H:%M"), across(where(is.numeric), ~round(., 1)))
create_dt(stn_data, "Wind Speed - M6")
```

:::

---

## Wind vs Wave Scatter {#scatter}

::: {.panel-tabset}

### All Buoys (Plot)

```{r}
#| fig-height: 5
stn_data <- get_station_data(buoy_data, "All Buoys", 30)
ggplot(stn_data, aes(x = wind_speed, y = wave_height, color = station_id)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
  labs(
    title = "Wind Speed vs Wave Height - All Stations",
    x = "Wind Speed (knots)",
    y = "Wave Height (m)",
    color = "Station"
  ) +
  theme_minimal()
```

### All Buoys (Data)

```{r}
stn_data <- get_station_data(buoy_data, "All Buoys", 30) |>
  select(time, station_id, wind_speed, wave_height, hmax) |>
  mutate(
    time = format(time, "%Y-%m-%d %H:%M"),
    wind_speed = round(wind_speed, 1),
    wave_height = round(wave_height, 2),
    hmax = round(hmax, 2)
  )
create_dt(stn_data, "Wind vs Wave Data")
```

### M2 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M2", 30)
ggplot(stn_data, aes(x = wind_speed, y = wave_height)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Wind vs Wave - M2", x = "Wind Speed (knots)", y = "Wave Height (m)") +
  theme_minimal()
```

### M3 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M3", 30)
ggplot(stn_data, aes(x = wind_speed, y = wave_height)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Wind vs Wave - M3", x = "Wind Speed (knots)", y = "Wave Height (m)") +
  theme_minimal()
```

### M4 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M4", 30)
ggplot(stn_data, aes(x = wind_speed, y = wave_height)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Wind vs Wave - M4", x = "Wind Speed (knots)", y = "Wave Height (m)") +
  theme_minimal()
```

### M5 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M5", 30)
ggplot(stn_data, aes(x = wind_speed, y = wave_height)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Wind vs Wave - M5", x = "Wind Speed (knots)", y = "Wave Height (m)") +
  theme_minimal()
```

### M6 (Plot)

```{r}
#| fig-height: 4
stn_data <- get_station_data(buoy_data, "M6", 30)
ggplot(stn_data, aes(x = wind_speed, y = wave_height)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Wind vs Wave - M6", x = "Wind Speed (knots)", y = "Wave Height (m)") +
  theme_minimal()
```

:::

---

## Recent Observations {#recent-observations}

```{r recent-obs}
recent_data <- buoy_data |>
  arrange(desc(time)) |>
  head(100) |>
  mutate(
    time = format(time, "%Y-%m-%d %H:%M"),
    wave_height = round(wave_height, 2),
    hmax = round(hmax, 2),
    wind_speed = round(wind_speed, 1),
    gust = round(gust, 1),
    air_temperature = round(air_temperature, 1)
  ) |>
  select(time, station_id, wave_height, hmax, wind_speed, gust, air_temperature)

datatable(
  recent_data,
  caption = "Most recent observations - click column headers to sort, use filters above each column",
  filter = 'top',
  colnames = c("Time (UTC)", "Station", "Wave Ht (m)", "Hmax (m)", "Wind (kn)", "Gust (kn)", "Air Temp (C)"),
  options = list(
    pageLength = 20,
    order = list(list(0, 'desc')),
    scrollX = TRUE
  ),
  class = 'cell-border stripe hover'
)
```

---

## Variable Definitions

| Variable | Units | Description |
|----------|-------|-------------|
| wave_height | meters | Significant wave height (Hs). 4x RMS of wave motion over 17.5 min |
| hmax | meters | Maximum individual wave height in measurement period |
| wave_period | seconds | Average wave period (zero-crossing analysis) |
| wind_speed | knots | 10-minute average wind speed. 1 knot = 0.514 m/s |
| gust | knots | Maximum 3-second wind speed during the hour |
| atmospheric_pressure | hPa | Barometric pressure (970-1030 typical) |
| air_temperature | degrees C | Air temperature at buoy (~3m above sea) |
| sea_temperature | degrees C | Sea surface temperature |

---

## Data Sources

- [Marine Institute Ireland](https://www.marine.ie/)
- [ERDDAP Data Portal](https://erddap.marine.ie/)
- [irishbuoys R Package](https://github.com/JohnGavin/irishbuoys)

---

*Dashboard generated `r Sys.time()` UTC*
