---
title: "Buoy Data Explorer (Perspective)"
format:
  html:
    code-fold: true
    code-summary: "Show code"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This dashboard uses [perspectiveR](https://github.com/EydlinIlya/perspectiveR)
to render buoy data client-side via WebAssembly. The HTML page is small (~2MB
for the Perspective WASM engine) and data is loaded from a Parquet file on
demand — no 34MB inline JSON.

See [issue #72](https://github.com/JohnGavin/irishbuoys/issues/72) for context.

## Data

Load the pre-computed buoy data from the static API Parquet file. This is the
same data that powers the main dashboard, but rendered via Perspective instead
of dygraphs.

```{r load-data}
parquet_path <- file.path("..", "docs", "vignettes", "data", "buoy_data.parquet")
if (!file.exists(parquet_path)) {
  parquet_path <- system.file(
    "extdata", "dashboard_buoy_data.rds",
    package = "irishbuoys"
  )
  buoy_data <- readRDS(parquet_path)
} else {
  buoy_data <- arrow::read_parquet(parquet_path)
}

# Subset to key columns for the pivot table
buoy_subset <- buoy_data[, c(
  "time", "station_id",
  "wave_height", "hmax", "wind_speed", "gust",
  "sea_temperature", "atmospheric_pressure"
)]

cat(sprintf(
  "Loaded %s rows x %s columns (%s to %s)\n",
  format(nrow(buoy_subset), big.mark = ","),
  ncol(buoy_subset),
  min(buoy_subset$time, na.rm = TRUE),
  max(buoy_subset$time, na.rm = TRUE)
))
```

## Pivot Table

The Perspective widget below runs entirely in the browser via WebAssembly.
You can:

- **Group** rows by station, year, month
- **Split** columns by station
- **Filter** by any column
- **Sort** by clicking column headers
- **Switch views**: datagrid, line chart, bar chart, heatmap

```{r perspective-pivot}
if (requireNamespace("perspectiveR", quietly = TRUE)) {
  perspectiveR::perspective(
    buoy_subset,
    group_by = "station_id",
    columns = c("wave_height", "hmax", "wind_speed", "gust",
                "sea_temperature", "atmospheric_pressure"),
    aggregates = list(
      wave_height = "avg",
      hmax = "max",
      wind_speed = "avg",
      gust = "max",
      sea_temperature = "avg",
      atmospheric_pressure = "avg"
    )
  )
} else {
  cat("perspectiveR not installed. Install with: install.packages('perspectiveR')")
}
```

## Time Series View

Group by time period to see trends.

```{r perspective-timeseries}
if (requireNamespace("perspectiveR", quietly = TRUE)) {
  # Add date column for daily aggregation
  buoy_daily <- buoy_subset
  buoy_daily$date <- as.Date(buoy_daily$time)

  perspectiveR::perspective(
    buoy_daily[, c("date", "station_id", "wave_height", "hmax",
                   "sea_temperature")],
    group_by = "date",
    split_by = "station_id",
    columns = c("wave_height", "hmax", "sea_temperature"),
    aggregates = list(
      wave_height = "avg",
      hmax = "max",
      sea_temperature = "avg"
    ),
    plugin = "Y Line"
  )
} else {
  cat("perspectiveR not installed.")
}
```

## Size Comparison

| Dashboard | HTML Size | Load Time | Interactivity |
|-----------|-----------|-----------|---------------|
| Current (dygraphs) | 34 MB | 10-30s | Zoom only |
| This (perspectiveR) | ~2 MB + data on demand | 2-5s | Pivot, filter, chart type |

## Next Steps

1. Compare render quality and load time with current dashboard
2. If successful, replace `dashboard_static.qmd` with this approach
3. Wire to HuggingFace Parquet via `ib_hf_url()` (#71) for zero-copy data access
