---
title: "Debug"
format:
  dashboard:
    orientation: rows
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# Debug page only needs targets data, not full package
# Avoid pkgload::load_all() which can cause segfaults in some environments
library(dplyr)
library(DT)
library(targets)

# Set targets store to project root
targets::tar_config_set(store = "../_targets")

# Get package version
pkg_version <- tryCatch(
  as.character(packageVersion("irishbuoys")),
  error = function(e) {
    desc_file <- "../DESCRIPTION"
    if (file.exists(desc_file)) {
      desc <- read.dcf(desc_file)
      desc[1, "Version"]
    } else "dev"
  }
)
generated_time <- format(Sys.time(), "%Y-%m-%d %H:%M UTC")

# Load validation report paths if available
validation_paths <- tryCatch(
  targets::tar_read(validation_reports),
  error = function(e) NULL
)
```

# Validation

Data quality validation using [pointblank](https://rstudio.github.io/pointblank/). Reports are generated automatically during `tar_make()`.

## Row

### Column {.tabset}

#### Analysis Data

Validates the main `analysis_data` target with checks for:

- Minimum row count (100+ rows)
- Required columns (`station_id`, `time`, `wave_height`)
- Valid station IDs (M1-M6, FS1)
- Physical measurement bounds (wave height 0-30m, wind 0-100 m/s, etc.)

```{r validation-analysis}
#| echo: false
if (!is.null(validation_paths) && file.exists(validation_paths$analysis_data)) {
  # Embed the pointblank report
  htmltools::tags$iframe(
    src = basename(validation_paths$analysis_data),
    width = "100%",
    height = "800px",
    style = "border: 1px solid #ccc; border-radius: 4px;"
  )
} else {
  htmltools::p(
    "Validation report not available. Run ",
    htmltools::code("tar_make()"),
    " to generate."
  )
}
```

#### Rogue Wave Events

Validates the `rogue_wave_events` target with checks for:

- At least 1 rogue wave event detected
- Required columns present
- `rogue_ratio > 2.0` (by definition of rogue wave)
- `hmax > wave_height` (max wave must exceed significant wave)

```{r validation-rogue}
#| echo: false
if (!is.null(validation_paths) && file.exists(validation_paths$rogue_events)) {
  # Embed the pointblank report
  htmltools::tags$iframe(
    src = basename(validation_paths$rogue_events),
    width = "100%",
    height = "800px",
    style = "border: 1px solid #ccc; border-radius: 4px;"
  )
} else {
  htmltools::p(
    "Validation report not available. Run ",
    htmltools::code("tar_make()"),
    " to generate."
  )
}
```

#### Validation Summary

```{r validation-summary}
#| echo: false
if (!is.null(validation_paths)) {
  summary_df <- tibble::tibble(
    Target = c("analysis_data", "rogue_wave_events"),
    Report = c(
      if (file.exists(validation_paths$analysis_data)) "Generated" else "Missing",
      if (file.exists(validation_paths$rogue_events)) "Generated" else "Missing"
    ),
    `Last Updated` = format(validation_paths$generated_time, "%Y-%m-%d %H:%M UTC")
  )
  DT::datatable(summary_df,
    extensions = "Buttons",
    options = list(dom = "Bfrtip", buttons = c("csv", "excel", "print"),
                   pageLength = 10, scrollX = TRUE),
    rownames = FALSE, class = "display compact"
  )
} else {
  htmltools::div(class = "alert alert-warning", "Validation reports not yet generated. Run `tar_make()` to create them.")
}
```

# Pipeline

Information about the targets pipeline status.

## Row

### Column {.tabset}

#### Target Status

```{r pipeline-status}
#| echo: false
tryCatch({
  # Get pipeline manifest
  manifest <- targets::tar_manifest()

  # Get progress
  progress <- targets::tar_progress()

  if (nrow(progress) > 0) {
    progress |>
      dplyr::select(name, progress) |>
      dplyr::arrange(name) |>
      DT::datatable(caption = "Target execution status",
        extensions = "Buttons",
        options = list(dom = "Bfrtip", buttons = c("csv", "excel", "print"),
                       pageLength = 10, scrollX = TRUE),
        rownames = FALSE, class = "display compact"
      )
  } else {
    htmltools::div(class = "alert alert-warning", "No targets have been run yet. Execute `tar_make()` to build the pipeline.")
  }
}, error = function(e) {
  htmltools::div(class = "alert alert-warning", paste("Could not read pipeline status:", e$message))
})
```

#### Outdated Targets

```{r pipeline-outdated}
#| echo: false
tryCatch({
  outdated <- targets::tar_outdated()

  if (length(outdated) > 0) {
    tibble::tibble(
      `Outdated Target` = outdated
    ) |>
      DT::datatable(caption = paste(length(outdated), "targets need rebuilding"),
        extensions = "Buttons",
        options = list(dom = "Bfrtip", buttons = c("csv", "excel", "print"),
                       pageLength = 10, scrollX = TRUE),
        rownames = FALSE, class = "display compact"
      )
  } else {
    htmltools::div(class = "alert alert-warning", "All targets are up to date.")
  }
}, error = function(e) {
  htmltools::div(class = "alert alert-warning", paste("Could not check outdated targets:", e$message))
})
```

#### Network Graph

The pipeline dependency graph (requires `tar_visnetwork()` to be run interactively).

```{r pipeline-graph}
#| echo: false
cat('To view the interactive pipeline graph, run in R:\n\n')
cat('\`\`\`r\n')
cat('targets::tar_visnetwork()\n')
cat('\`\`\`\n\n')
cat('This will open an interactive visualization showing all targets and their dependencies.\n')
```

This will open an interactive visualization showing all targets and their dependencies.
")
```

# Session Info

## Row

### Column

#### R Session

```{r session-r}
#| echo: false
sessionInfo()
```

### Column

#### Package Versions

```{r session-packages}
#| echo: false
core_pkgs <- c("irishbuoys", "pointblank", "targets", "dplyr", "duckdb", "arrow")

pkg_versions <- sapply(core_pkgs, function(pkg) {
  tryCatch(
    as.character(packageVersion(pkg)),
    error = function(e) "not installed"
  )
})

tibble::tibble(
  Package = names(pkg_versions),
  Version = pkg_versions
) |>
  DT::datatable(caption = "Core package versions",
    extensions = "Buttons",
    options = list(dom = "Bfrtip", buttons = c("csv", "excel", "print"),
                   pageLength = 10, scrollX = TRUE),
    rownames = FALSE, class = "display compact"
  )
```

---

*Debug page generated `r generated_time` using irishbuoys v`r pkg_version`*
