---
title: "Storm Éowyn — October 27-28, 2020"
format:
  closeread-html:
    code-fold: true
    code-summary: "Show code"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
library(plotly)
```

```{r load-data}
storm_path <- system.file("extdata", "storm_oct2020.rds", package = "irishbuoys")
if (!nzchar(storm_path)) storm_path <- "../inst/extdata/storm_oct2020.rds"
storm <- readRDS(storm_path)

pkgload::load_all("../", quiet = TRUE)
stations <- get_station_info()

# Peak stats per station
peaks <- storm |>
  filter(!is.na(hmax)) |>
  group_by(station_id) |>
  summarise(
    peak_hmax = round(max(hmax, na.rm = TRUE), 1),
    peak_time = time[which.max(hmax)],
    peak_gust = round(max(gust, na.rm = TRUE), 1),
    mean_wave = round(mean(wave_height, na.rm = TRUE), 1),
    n_obs = n(),
    .groups = "drop"
  ) |>
  arrange(desc(peak_hmax))
```

:::{.cr-section}

## The Calm Before {focus-on="cr-map"}

On October 26, 2020, five weather buoys stood watch off Ireland's Atlantic
coast. Wave heights were modest — 2-3 metres at the offshore stations M6 and
M4, barely a metre near the coast.

Within 48 hours, the network would record the highest wave in its
history.

@cr-map

:::

:::{#cr-map}
```{r map-buoys}
plot_ly() |>
  add_markers(
    data = stations,
    x = ~lon, y = ~lat,
    text = ~station_id,
    marker = list(size = 14, color = "#377EB8"),
    hovertemplate = "%{text}<br>%{y:.2f}°N, %{x:.2f}°W<extra></extra>"
  ) |>
  layout(
    title = "Irish Weather Buoy Network",
    xaxis = list(title = "Longitude"),
    yaxis = list(title = "Latitude"),
    paper_bgcolor = "#1a1a1a",
    plot_bgcolor = "#1a1a1a",
    font = list(color = "white")
  )
```
:::

:::{.cr-section}

## The Storm Builds {focus-on="cr-wave-plot"}

By October 27, an Atlantic low-pressure system was driving swells across
the network. M6, the westernmost buoy 320 km offshore, was the first to
feel it. Wave heights climbed through 5, then 8, then 12 metres.

The pressure at M6 dropped from 1010 to 975 hPa in 18 hours — a rapid
deepening characteristic of a "weather bomb."

@cr-wave-plot

:::

:::{#cr-wave-plot}
```{r wave-timeseries}
p <- storm |>
  filter(!is.na(hmax)) |>
  plot_ly(
    x = ~time, y = ~hmax,
    color = ~station_id,
    colors = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00"),
    type = "scatter", mode = "lines",
    hovertemplate = "%{x}<br>Hmax: %{y:.1f} m<extra>%{data.name}</extra>"
  ) |>
  layout(
    title = "Maximum Wave Height (Hmax) — Oct 26-29, 2020",
    xaxis = list(title = "", rangeslider = list(visible = TRUE)),
    yaxis = list(title = "Hmax (m)"),
    paper_bgcolor = "#1a1a1a",
    plot_bgcolor = "#1a1a1a",
    font = list(color = "white"),
    legend = list(
      orientation = "h", y = -0.3,
      font = list(color = "white"),
      bgcolor = "rgba(0,0,0,0.5)"
    )
  )
p
```
:::

:::{.cr-section}

## Peak: 29.9 Metres {focus-on="cr-peak-table"}

On October 28 at approximately 06:00 UTC, station M6 recorded a maximum
individual wave height of `r peaks$peak_hmax[peaks$station_id == "M6"]` metres —
the highest in the network's history.

The wave was roughly the height of a 10-storey building.

Stations M3 and M4 recorded their peaks 6-12 hours later as the storm
tracked eastward, consistent with the spatial propagation pattern
identified by the H3 rogue wave analysis.

@cr-peak-table

:::

:::{#cr-peak-table}
```{r peak-table}
peaks |>
  mutate(
    peak_time = format(peak_time, "%b %d %H:%M UTC")
  ) |>
  DT::datatable(
    colnames = c("Station", "Peak Hmax (m)", "Peak Time", "Peak Gust (kn)",
                 "Mean Wave (m)", "Observations"),
    options = list(dom = "t", pageLength = 10),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; color: white;",
      "Peak conditions per station during Oct 26-29, 2020 storm."
    ),
    rownames = FALSE
  )
```
:::

:::{.cr-section}

## The Aftermath {focus-on="cr-pressure"}

By October 29, the system had passed. Wave heights returned to 3-4 metres,
pressure climbed back above 1000 hPa. The storm had affected all reporting
stations across `r length(unique(storm$station_id))` buoys over
`r as.integer(difftime(max(storm$time), min(storm$time), units = "hours"))` hours.

The event demonstrates why Ireland's offshore buoy network is critical:
M6's early detection gave `r round(as.numeric(difftime(peaks$peak_time[peaks$station_id == "M3"], peaks$peak_time[peaks$station_id == "M6"], units = "hours")), 0)` hours
of lead time before the storm reached the coastal stations.

@cr-pressure

:::

:::{#cr-pressure}
```{r pressure-plot}
storm |>
  filter(!is.na(atmospheric_pressure)) |>
  plot_ly(
    x = ~time, y = ~atmospheric_pressure,
    color = ~station_id,
    colors = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00"),
    type = "scatter", mode = "lines",
    hovertemplate = "%{x}<br>%{y:.0f} hPa<extra>%{data.name}</extra>"
  ) |>
  layout(
    title = "Atmospheric Pressure — Oct 26-29, 2020",
    xaxis = list(title = ""),
    yaxis = list(title = "Pressure (hPa)"),
    paper_bgcolor = "#1a1a1a",
    plot_bgcolor = "#1a1a1a",
    font = list(color = "white"),
    legend = list(
      orientation = "h", y = -0.2,
      font = list(color = "white"),
      bgcolor = "rgba(0,0,0,0.5)"
    )
  )
```
:::
