---
title: "Telemetry"
format:
  dashboard:
    orientation: rows
    theme: cosmo
    scrolling: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# This vignette performs NO computation.
# All data is loaded from pre-computed targets via tar_load().
# MANDATORY: Uses FULL historical data (2019-2026), never sampled.
library(targets)

# Set targets store relative to vignette location
targets::tar_config_set(store = "../_targets")

# Helper: Create informative "not available" message
not_available_msg <- function(target_name) {
  htmltools::div(
    class = "alert alert-warning",
    htmltools::h4("Data Not Available"),
    htmltools::p(
      "The ", htmltools::code(target_name), " target has not been built."
    ),
    htmltools::h5("How to fix:"),
    htmltools::tags$ol(
      htmltools::tags$li("Enter Nix shell: ", htmltools::code("./default.sh")),
      htmltools::tags$li("Run pipeline: ", htmltools::code("targets::tar_make()")),
      htmltools::tags$li("Refresh this page")
    ),
    htmltools::h5("Common issues:"),
    htmltools::tags$ul(
      htmltools::tags$li("Database not initialized: Run ", htmltools::code("initialize_database(start_date = '2019-01-01')")),
      htmltools::tags$li("DuckDB lock: Close other R sessions accessing the database"),
      htmltools::tags$li("Missing packages: Re-enter Nix shell")
    )
  )
}

# Load all telemetry targets with error handling
tryCatch({
  tar_load(telemetry_summary)
  tar_load(telemetry_station_coverage)
  tar_load(telemetry_db_info)
  tar_load(telemetry_db_file_size)
  tar_load(telemetry_data_freshness)
  tar_load(telemetry_pipeline_status)
  tar_load(telemetry_qc_trends)
  tar_load(telemetry_rogue_frequency)
  tar_load(telemetry_git_commits)
  tar_load(telemetry_commit_frequency)
  tar_load(telemetry_file_types)

  # Tables
  tar_load(table_telemetry_station_coverage)
  tar_load(table_telemetry_db_info)
  tar_load(table_telemetry_freshness)
  tar_load(table_telemetry_commits)
  tar_load(table_telemetry_file_types)
  tar_load(telemetry_target_metrics)
  tar_load(table_telemetry_target_metrics)

  # Test statistics
  tar_load(telemetry_test_stats)
  tar_load(telemetry_test_summary)
  tar_load(table_telemetry_test_stats)
  tar_load(table_telemetry_test_summary)

  # LLM Usage & Costs
  tar_load(telemetry_llm_sessions)
  tar_load(telemetry_llm_daily)
  tar_load(telemetry_llm_summary)
  tar_load(telemetry_llm_by_model)
  tar_load(plot_telemetry_llm_cost_trend)
  tar_load(plot_telemetry_llm_tokens)
  tar_load(plot_telemetry_llm_by_model)
  tar_load(table_telemetry_llm_summary)
  tar_load(table_telemetry_llm_daily)
  tar_load(table_telemetry_llm_by_model)

  # Plots
  tar_load(plot_telemetry_coverage_timeline)
  tar_load(plot_telemetry_qc_trends)
  tar_load(plot_telemetry_rogue_frequency)
  tar_load(plot_telemetry_commit_frequency)

  targets_loaded <- TRUE
}, error = function(e) {
  targets_loaded <<- FALSE
})
```

# Overview {#overview}

Project health and data pipeline metrics for the irishbuoys package.

## Key Metrics {height=20%}

### Total Observations {#total-obs}

::: {.valuebox title="Total Observations" icon="database"}
```{r}
#| output: asis
if (exists("telemetry_summary") && !is.null(telemetry_summary)) {
  cat(format(telemetry_summary$total_observations, big.mark = ","))
} else {
  cat("--")
}
```
:::

### Database Size {#db-size}

::: {.valuebox title="Database Size" icon="hdd"}
```{r}
#| output: asis
if (exists("telemetry_summary") && !is.null(telemetry_summary$db_size_mb)) {
  cat(paste0(telemetry_summary$db_size_mb, " MB"))
} else {
  cat("--")
}
```
:::

### Active Stations {#active-stations}

::: {.valuebox title="Active Stations" icon="broadcast"}
```{r}
#| output: asis
if (exists("telemetry_summary") && !is.null(telemetry_summary)) {
  cat(telemetry_summary$stations_active)
} else {
  cat("--")
}
```
:::

### Recent Commits {#recent-commits-count}

::: {.valuebox title="Commits (30d)" icon="git-branch"}
```{r}
#| output: asis
if (exists("telemetry_summary") && !is.null(telemetry_summary)) {
  cat(telemetry_summary$n_commits_30d)
} else {
  cat("--")
}
```
:::

## Status Summary {#status-summary}

### Last Updated {#last-updated}

::: {.card title="Pipeline Status"}
```{r}
#| output: asis
if (exists("telemetry_summary") && !is.null(telemetry_summary)) {
  cat("**Last updated:** ", format(telemetry_summary$generated_at, "%Y-%m-%d %H:%M UTC"), "\n\n")
  cat("**Date range:** 2019-01-01 to ", format(Sys.Date()), " (full historical)\n\n")
  cat("**Data source:** Irish Weather Buoy Network via ERDDAP")
} else {
  cat("Targets pipeline has not been run.\n\n")
  cat("**To initialize:**\n")
  cat("1. `./default.sh` (enter Nix shell)\n")
  cat("2. `initialize_database(start_date = '2019-01-01')` (download full history)\n")
  cat("3. `targets::tar_make()` (build all targets)\n")
}
```
:::

# Data Coverage {#data-coverage}

Station data availability and freshness metrics across the full historical period (2019-present).

## Coverage Analysis {#coverage-analysis}

### Coverage Details {.tabset}

#### Timeline {#coverage-timeline}

```{r}
#| fig-height: 6
if (exists("plot_telemetry_coverage_timeline") && !is.null(plot_telemetry_coverage_timeline)) {
  plot_telemetry_coverage_timeline
} else {
  not_available_msg("plot_telemetry_coverage_timeline")
}
```

#### By Station {#coverage-by-station}

```{r}
if (exists("table_telemetry_station_coverage")) {
  table_telemetry_station_coverage
} else {
  not_available_msg("table_telemetry_station_coverage")
}
```

#### Freshness {#coverage-freshness}

```{r}
if (exists("table_telemetry_freshness")) {
  table_telemetry_freshness
} else {
  not_available_msg("table_telemetry_freshness")
}
```

# Data Quality {#data-quality}

QC flag trends and rogue wave detection metrics for the full historical period.

## Quality Metrics {#quality-metrics}

### QC Distribution {#qc-distribution width=50%}

::: {.card title="QC Flag Trends (Monthly, 2019-Present)"}
QC flags: 0=unknown, 1=good, 9=missing. Chart shows monthly distribution across full historical data, colored by season.

```{r}
#| fig-height: 5
if (exists("plot_telemetry_qc_trends") && !is.null(plot_telemetry_qc_trends)) {
  plot_telemetry_qc_trends
} else {
  not_available_msg("plot_telemetry_qc_trends")
}
```
:::

### Rogue Wave Events {#rogue-events width=50%}

::: {.card title="Rogue Wave Frequency (Monthly)"}
Rogue waves: Hmax > 2 × significant wave height. Shows monthly event counts across full historical period.

```{r}
#| fig-height: 5
if (exists("plot_telemetry_rogue_frequency") && !is.null(plot_telemetry_rogue_frequency)) {
  plot_telemetry_rogue_frequency
} else {
  not_available_msg("plot_telemetry_rogue_frequency")
}
```
:::

# Database {#database}

Storage metrics and table information for the DuckDB database.

## Storage Details {#storage-details}

### Database Tables {.tabset}

#### Table Summary {#table-summary}

```{r}
if (exists("table_telemetry_db_info")) {
  table_telemetry_db_info
} else {
  not_available_msg("table_telemetry_db_info")
}
```

#### Storage Size {#storage-size}

```{r}
if (exists("telemetry_db_file_size") && nrow(telemetry_db_file_size) > 0) {
  df <- telemetry_db_file_size
  htmltools::div(
    htmltools::h4("Database File: ", df$path),
    htmltools::tags$ul(
      htmltools::tags$li("Size: ", df$size_mb, " MB"),
      htmltools::tags$li("Last modified: ", format(df$last_modified, "%Y-%m-%d %H:%M"))
    )
  )
} else {
  not_available_msg("telemetry_db_file_size")
}
```

# Pipeline {#pipeline}

Targets pipeline status, timing metrics, and execution history.

## Pipeline Status {#pipeline-status}

### Status Summary {#status-summary-pipeline width=40%}

::: {.card title="Target Execution Status"}
```{r}
if (exists("telemetry_pipeline_status") && !is.null(telemetry_pipeline_status$status_summary)) {
  knitr::kable(telemetry_pipeline_status$status_summary, format = "html")
} else {
  not_available_msg("telemetry_pipeline_status")
}
```
:::

### Timing Metrics {#timing-metrics width=60%}

::: {.card title="Pipeline Performance"}
```{r}
if (exists("telemetry_pipeline_status") && !is.null(telemetry_pipeline_status$timing)) {
  timing <- telemetry_pipeline_status$timing
  htmltools::tags$ul(
    htmltools::tags$li(paste("Total runtime:", round(timing$total_runtime_sec, 1), "seconds")),
    htmltools::tags$li(paste("Average target:", round(timing$avg_runtime_sec, 2), "seconds")),
    htmltools::tags$li(paste("Slowest target:", round(timing$max_runtime_sec, 1), "seconds")),
    htmltools::tags$li(paste("Last run:", format(timing$last_run, "%Y-%m-%d %H:%M")))
  )
} else {
  not_available_msg("telemetry_pipeline_status")
}
```
:::

## Targets by Plan {#targets-by-plan}

### Plan Metrics {#plan-metrics}

::: {.card title="Target Counts, Size, and Timing by Pipeline Plan"}
Each row shows a tar_plan (from R/tar_plans/plan_*.R) with aggregate metrics.

```{r}
if (exists("table_telemetry_target_metrics")) {
  table_telemetry_target_metrics
} else {
  not_available_msg("table_telemetry_target_metrics")
}
```
:::

# Git Activity {#git-activity}

Repository commit history and development activity.

## Development Metrics {#dev-metrics}

### Commit Frequency {#commit-frequency width=40%}

::: {.card title="Commits per Day (Last 30 Days)"}
```{r}
#| fig-height: 4
if (exists("plot_telemetry_commit_frequency") && !is.null(plot_telemetry_commit_frequency)) {
  plot_telemetry_commit_frequency
} else {
  not_available_msg("plot_telemetry_commit_frequency")
}
```
:::

### Commit History {#commit-history width=60%}

::: {.card title="Recent Commits"}
```{r}
if (exists("table_telemetry_commits")) {
  table_telemetry_commits
} else {
  not_available_msg("table_telemetry_commits")
}
```
:::

# Project Files {#project-files}

Project structure and file statistics.

## File Statistics {#file-stats}

### Files by Type {#files-by-type}

::: {.card title="File Type Distribution"}
```{r}
if (exists("table_telemetry_file_types")) {
  table_telemetry_file_types
} else {
  not_available_msg("table_telemetry_file_types")
}
```
:::

# LLM Usage & Costs {#llm-usage}

Claude Code usage statistics for the irishbuoys project.

## Usage Summary {#usage-summary}

### Current Status {#usage-status}

::: {.card title="LLM Usage Summary"}
```{r}
if (exists("table_telemetry_llm_summary") && !is.null(table_telemetry_llm_summary)) {
  table_telemetry_llm_summary
} else {
  not_available_msg("table_telemetry_llm_summary")
}
```
:::

## Usage Details {#usage-details}

### Usage Analytics {.tabset}

#### Cost Trends {#cost-trends}

::: {.card title="Daily LLM Costs"}
Estimated daily costs based on Claude API pricing (Opus: $15/$75 per M tokens, Sonnet: $3/$15 per M tokens).

```{r}
#| fig-height: 5
if (exists("plot_telemetry_llm_cost_trend") && !is.null(plot_telemetry_llm_cost_trend)) {
  plot_telemetry_llm_cost_trend
} else {
  not_available_msg("plot_telemetry_llm_cost_trend")
}
```
:::

#### Token Usage {#token-usage}

::: {.card title="Token Usage by Type"}
Breakdown of input, output, and cache tokens. Cache tokens (creation and read) help reduce costs.

```{r}
#| fig-height: 5
if (exists("plot_telemetry_llm_tokens") && !is.null(plot_telemetry_llm_tokens)) {
  plot_telemetry_llm_tokens
} else {
  not_available_msg("plot_telemetry_llm_tokens")
}
```
:::

#### By Model {#by-model}

::: {.card title="Cost by Model"}
```{r}
#| fig-height: 4
if (exists("plot_telemetry_llm_by_model") && !is.null(plot_telemetry_llm_by_model)) {
  plot_telemetry_llm_by_model
} else {
  not_available_msg("plot_telemetry_llm_by_model")
}
```

```{r}
if (exists("table_telemetry_llm_by_model") && !is.null(table_telemetry_llm_by_model)) {
  table_telemetry_llm_by_model
} else {
  not_available_msg("table_telemetry_llm_by_model")
}
```
:::

#### Daily Log {#daily-log}

::: {.card title="Daily Usage Log"}
```{r}
if (exists("table_telemetry_llm_daily") && !is.null(table_telemetry_llm_daily)) {
  table_telemetry_llm_daily
} else {
  not_available_msg("table_telemetry_llm_daily")
}
```
:::

#### About Costs {#about-costs}

::: {.card title="Cost Estimation Notes"}

**How costs are calculated:**

Costs are estimated using Claude API pricing (as of 2025):

| Model | Input | Output | Cache Create | Cache Read |
|-------|-------|--------|--------------|------------|
| Opus 4.5 | $15/M | $75/M | $18.75/M | $1.875/M |
| Sonnet 4 | $3/M | $15/M | $3.75/M | $0.375/M |
| Haiku 3.5 | $0.80/M | $4/M | $1/M | $0.08/M |

**Token types:**

- **Input tokens**: Tokens in your prompts
- **Output tokens**: Tokens in Claude's responses
- **Cache creation**: First-time caching of context
- **Cache read**: Reusing cached context (much cheaper)

**Data source:**

Usage data is extracted from Claude Code session files in `~/.claude/projects/`.
:::

# Tests {#tests}

Unit test and quality assurance metrics for the irishbuoys package.

## Test Summary {#test-summary}

### Test Statistics {.tabset}

#### By Category {#test-by-category}

::: {.card title="Test Summary by Type"}
Aggregated counts of tests, expectations, and snapshot tests grouped by test category.

```{r}
if (exists("table_telemetry_test_summary")) {
  table_telemetry_test_summary
} else {
  not_available_msg("table_telemetry_test_summary")
}
```
:::

#### By File {#test-by-file}

::: {.card title="Individual Test Files"}
```{r}
if (exists("table_telemetry_test_stats")) {
  table_telemetry_test_stats
} else {
  not_available_msg("table_telemetry_test_stats")
}
```
:::

#### Test Types {#test-types}

::: {.card title="About Test Categories"}

**Test categories used:**

- **Unit Tests**: Standard test_that() blocks testing individual functions
- **Adversarial/Edge Cases**: Tests for NULL, NA, empty inputs, boundary conditions
- **Defensive/Null Guards**: Tests verifying cli::cli_abort() on invalid inputs
- **Contains Snapshots**: Files with expect_snapshot() calls
- **Snapshot Tests**: Files primarily focused on snapshot testing
- **Integration Tests**: End-to-end tests spanning multiple components

**What we test:**

1. **Input validation**: All exported functions reject NULL/invalid inputs with clear errors
2. **Physical bounds**: Wave heights 0-30m, wind 0-100 m/s, pressure 900-1100 hPa
3. **Data integrity**: QC flags, station IDs, time ranges
4. **Rogue wave detection**: Hmax > 2 × Hs criterion
5. **Database operations**: Connection, query, incremental updates
:::

# Validation: Analysis Data {#validation-analysis}

Pointblank validation report for the main analysis dataset.

## Analysis Data Validation {#analysis-validation}

### Analysis Validation Details {.tabset}

#### Report {#analysis-report}

```{r}
report_path <- "../docs/articles/validation_analysis_data.html"
if (file.exists(report_path)) {
  htmltools::tags$iframe(
    src = "validation_analysis_data.html",
    width = "100%",
    height = "800px",
    style = "border: 1px solid #ccc; border-radius: 4px;"
  )
} else {
  not_available_msg("validation_analysis_data.html")
}
```

#### Notes {#analysis-notes}

::: {.card title="About This Validation"}

**What is pointblank?**

[pointblank](https://rstudio.github.io/pointblank/) is an R package for data validation. It creates agents that interrogate data frames and report on validation failures.

**Validation checks performed:**

- **Row count**: Minimum 100 observations required
- **Required columns**: `station_id`, `time`, `wave_height` must exist
- **Station IDs**: Must be valid (M1-M6, FS1)
- **Wave height**: Physical bounds 0-30 meters
- **Wind speed**: Physical bounds 0-100 m/s
- **Atmospheric pressure**: Physical bounds 900-1100 hPa
- **Time**: Not in the future

**How to regenerate:**

```r
targets::tar_make(names = "validation_reports")
```
:::

# Validation: Rogue Waves {#validation-rogue}

Pointblank validation report for detected rogue wave events.

## Rogue Wave Validation {#rogue-validation}

### Rogue Validation Details {.tabset}

#### Report {#rogue-report}

```{r}
report_path <- "../docs/articles/validation_rogue_events.html"
if (file.exists(report_path)) {
  htmltools::tags$iframe(
    src = "validation_rogue_events.html",
    width = "100%",
    height = "800px",
    style = "border: 1px solid #ccc; border-radius: 4px;"
  )
} else {
  not_available_msg("validation_rogue_events.html")
}
```

#### Notes {#rogue-notes}

::: {.card title="About Rogue Wave Validation"}

**What are rogue waves?**

Rogue waves are unusually large waves defined by `Hmax > 2 × Hs` where:
- **Hmax**: Maximum individual wave height
- **Hs**: Significant wave height (mean of highest 1/3 of waves)

**Validation checks performed:**

- **Row count**: At least 1 rogue event expected
- **Required columns**: `station_id`, `time`, `wave_height`, `hmax`, `rogue_ratio`
- **Rogue ratio**: Must be > 2.0 (by definition)
- **Wave relationship**: `hmax > wave_height` must hold

**Reference:**

See `?detect_rogue_waves` for the detection algorithm.
:::

---

```{r}
#| echo: false
cat("*Generated:", format(Sys.time(), "%Y-%m-%d %H:%M UTC"), "*\n")
cat("*All data loaded from pre-computed targets via tar_load()*\n")
cat("*Data range: Full historical (2019-present), NO sampling*")
```
