---
title: "Telemetry"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# All data loaded from pre-computed targets via tar_load().
# MANDATORY: Uses FULL historical data (2019-2026), never sampled.
library(targets)

targets::tar_config_set(store = "../_targets")

not_available_msg <- function(target_name) {
  htmltools::div(
    class = "alert alert-warning",
    htmltools::h4("Data Not Available"),
    htmltools::p(
      "The ", htmltools::code(target_name), " target has not been built."
    ),
    htmltools::p("Run ", htmltools::code("targets::tar_make()"), " to generate.")
  )
}

tryCatch({
  tar_load(telemetry_summary)
  tar_load(telemetry_station_coverage)
  tar_load(telemetry_db_info)
  tar_load(telemetry_db_file_size)
  tar_load(telemetry_data_freshness)
  tar_load(telemetry_pipeline_status)
  tar_load(telemetry_qc_trends)
  tar_load(telemetry_rogue_frequency)
  tar_load(telemetry_git_commits)
  tar_load(telemetry_commit_frequency)
  tar_load(telemetry_file_types)

  tar_load(table_telemetry_station_coverage)
  tar_load(table_telemetry_db_info)
  tar_load(table_telemetry_freshness)
  tar_load(table_telemetry_commits)
  tar_load(table_telemetry_file_types)
  tar_load(telemetry_target_metrics)
  tar_load(table_telemetry_target_metrics)

  tar_load(telemetry_test_stats)
  tar_load(telemetry_test_summary)
  tar_load(table_telemetry_test_stats)
  tar_load(table_telemetry_test_summary)

  tar_load(telemetry_llm_sessions)
  tar_load(telemetry_llm_daily)
  tar_load(telemetry_llm_summary)
  tar_load(telemetry_llm_by_model)
  tar_load(plot_telemetry_llm_cost_trend)
  tar_load(plot_telemetry_llm_tokens)
  tar_load(plot_telemetry_llm_by_model)
  tar_load(table_telemetry_llm_summary)
  tar_load(table_telemetry_llm_daily)
  tar_load(table_telemetry_llm_by_model)

  tar_load(plot_telemetry_coverage_timeline)
  tar_load(plot_telemetry_qc_trends)
  tar_load(plot_telemetry_rogue_frequency)
  tar_load(plot_telemetry_commit_frequency)

  targets_loaded <- TRUE
}, error = function(e) {
  targets_loaded <<- FALSE
})
```

## Overview {#overview}

Project health and data pipeline metrics for the irishbuoys package.

### Key Metrics {#key-metrics}

```{r}
if (exists("telemetry_summary") && !is.null(telemetry_summary)) {
  metrics <- data.frame(
    Metric = c("Total Observations", "Database Size", "Active Stations",
               "Commits (30d)", "Last Updated", "Date Range"),
    Value = c(
      format(telemetry_summary$total_observations, big.mark = ","),
      paste0(telemetry_summary$db_size_mb, " MB"),
      as.character(telemetry_summary$stations_active),
      as.character(telemetry_summary$n_commits_30d),
      format(telemetry_summary$generated_at, "%Y-%m-%d %H:%M UTC"),
      paste("2019-01-01 to", format(Sys.Date()))
    ),
    stringsAsFactors = FALSE
  )
  knitr::kable(metrics, align = c("l", "r"))
} else {
  not_available_msg("telemetry_summary")
}
```

## Data Coverage {#data-coverage}

Station data availability and freshness across the full historical period (2019-present).

### Coverage Timeline {#coverage-timeline}

```{r}
#| fig-height: 6
if (exists("plot_telemetry_coverage_timeline") && !is.null(plot_telemetry_coverage_timeline)) {
  plot_telemetry_coverage_timeline
} else {
  not_available_msg("plot_telemetry_coverage_timeline")
}
```

### Coverage by Station {#coverage-by-station}

```{r}
if (exists("table_telemetry_station_coverage")) {
  table_telemetry_station_coverage
} else {
  not_available_msg("table_telemetry_station_coverage")
}
```

### Data Freshness {#coverage-freshness}

```{r}
if (exists("table_telemetry_freshness")) {
  table_telemetry_freshness
} else {
  not_available_msg("table_telemetry_freshness")
}
```

## Data Quality {#data-quality}

QC flag trends and rogue wave detection metrics for the full historical period.

### QC Flag Distribution {#qc-distribution}

QC flags: 0=unknown, 1=good, 9=missing. Monthly distribution across full historical data, colored by season.

```{r}
#| fig-height: 5
if (exists("plot_telemetry_qc_trends") && !is.null(plot_telemetry_qc_trends)) {
  plot_telemetry_qc_trends
} else {
  not_available_msg("plot_telemetry_qc_trends")
}
```

### Rogue Wave Frequency {#rogue-events}

Rogue waves: Hmax > 2 x significant wave height. Monthly event counts across full historical period.

```{r}
#| fig-height: 5
if (exists("plot_telemetry_rogue_frequency") && !is.null(plot_telemetry_rogue_frequency)) {
  plot_telemetry_rogue_frequency
} else {
  not_available_msg("plot_telemetry_rogue_frequency")
}
```

## Database {#database}

Storage metrics and table information for the DuckDB database.

### Table Summary {#table-summary}

```{r}
if (exists("table_telemetry_db_info")) {
  table_telemetry_db_info
} else {
  not_available_msg("table_telemetry_db_info")
}
```

### Storage Size {#storage-size}

```{r}
if (exists("telemetry_db_file_size") && nrow(telemetry_db_file_size) > 0) {
  df <- telemetry_db_file_size
  storage <- data.frame(
    Metric = c("File", "Size", "Last Modified"),
    Value = c(df$path, paste0(df$size_mb, " MB"),
              format(df$last_modified, "%Y-%m-%d %H:%M")),
    stringsAsFactors = FALSE
  )
  knitr::kable(storage, align = c("l", "r"))
} else {
  not_available_msg("telemetry_db_file_size")
}
```

## Pipeline {#pipeline}

Targets pipeline status, timing metrics, and execution history.

### Target Execution Status {#target-status}

```{r}
if (exists("telemetry_pipeline_status") && !is.null(telemetry_pipeline_status$status_summary)) {
  DT::datatable(telemetry_pipeline_status$status_summary,
    extensions = "Buttons",
    options = list(dom = "Bfrtip", buttons = c("csv", "excel", "print"),
                   pageLength = 10, scrollX = TRUE),
    rownames = FALSE, class = "display compact"
  )
} else {
  not_available_msg("telemetry_pipeline_status")
}
```

### Pipeline Performance {#timing-metrics}

```{r}
if (exists("telemetry_pipeline_status") && !is.null(telemetry_pipeline_status$timing)) {
  timing <- telemetry_pipeline_status$timing
  perf <- data.frame(
    Metric = c("Total runtime", "Average per target", "Slowest target", "Last run"),
    Value = c(
      paste(round(timing$total_runtime_sec, 1), "seconds"),
      paste(round(timing$avg_runtime_sec, 2), "seconds"),
      paste(round(timing$max_runtime_sec, 1), "seconds"),
      format(timing$last_run, "%Y-%m-%d %H:%M")
    ),
    stringsAsFactors = FALSE
  )
  knitr::kable(perf, align = c("l", "r"))
} else {
  not_available_msg("telemetry_pipeline_status")
}
```

### Targets by Plan {#targets-by-plan}

Target counts, size, and timing by pipeline plan (from `R/tar_plans/plan_*.R`).

```{r}
if (exists("table_telemetry_target_metrics")) {
  table_telemetry_target_metrics
} else {
  not_available_msg("table_telemetry_target_metrics")
}
```

## Tests {#tests}

Unit test and quality assurance metrics for the irishbuoys package.

::: {.panel-tabset}

### By Category {#test-by-category}

Aggregated counts of tests, expectations, and snapshot tests grouped by test category.

```{r}
if (exists("table_telemetry_test_summary")) {
  table_telemetry_test_summary
} else {
  not_available_msg("table_telemetry_test_summary")
}
```

### By File {#test-by-file}

```{r}
if (exists("table_telemetry_test_stats")) {
  table_telemetry_test_stats
} else {
  not_available_msg("table_telemetry_test_stats")
}
```

### Test Types {#test-types}

| Category | Description |
|----------|-------------|
| Unit Tests | Standard `test_that()` blocks testing individual functions |
| Adversarial/Edge Cases | Tests for NULL, NA, empty inputs, boundary conditions |
| Defensive/Null Guards | Tests verifying `cli::cli_abort()` on invalid inputs |
| Contains Snapshots | Files with `expect_snapshot()` calls |
| Integration Tests | End-to-end tests spanning multiple components |

:::

## Git Activity {#git-activity}

Repository commit history and development activity.

### Commit Frequency {#commit-frequency}

```{r}
#| fig-height: 4
if (exists("plot_telemetry_commit_frequency") && !is.null(plot_telemetry_commit_frequency)) {
  plot_telemetry_commit_frequency
} else {
  not_available_msg("plot_telemetry_commit_frequency")
}
```

### Recent Commits {#commit-history}

```{r}
if (exists("table_telemetry_commits")) {
  table_telemetry_commits
} else {
  not_available_msg("table_telemetry_commits")
}
```

## Project Files {#project-files}

### File Type Distribution {#files-by-type}

```{r}
if (exists("table_telemetry_file_types")) {
  table_telemetry_file_types
} else {
  not_available_msg("table_telemetry_file_types")
}
```

## LLM Usage {#llm-usage}

Claude Code usage statistics for the irishbuoys project.

### Usage Summary {#usage-summary}

```{r}
if (exists("table_telemetry_llm_summary") && !is.null(table_telemetry_llm_summary)) {
  table_telemetry_llm_summary
} else {
  not_available_msg("table_telemetry_llm_summary")
}
```

::: {.panel-tabset}

### Cost Trends {#cost-trends}

Estimated daily costs based on Claude API pricing.

```{r}
#| fig-height: 5
if (exists("plot_telemetry_llm_cost_trend") && !is.null(plot_telemetry_llm_cost_trend)) {
  plot_telemetry_llm_cost_trend
} else {
  not_available_msg("plot_telemetry_llm_cost_trend")
}
```

### Token Usage {#token-usage}

Breakdown of input, output, and cache tokens. Cache tokens reduce costs.

```{r}
#| fig-height: 5
if (exists("plot_telemetry_llm_tokens") && !is.null(plot_telemetry_llm_tokens)) {
  plot_telemetry_llm_tokens
} else {
  not_available_msg("plot_telemetry_llm_tokens")
}
```

### By Model {#by-model}

```{r}
#| fig-height: 4
if (exists("plot_telemetry_llm_by_model") && !is.null(plot_telemetry_llm_by_model)) {
  plot_telemetry_llm_by_model
} else {
  not_available_msg("plot_telemetry_llm_by_model")
}
```

```{r}
if (exists("table_telemetry_llm_by_model") && !is.null(table_telemetry_llm_by_model)) {
  table_telemetry_llm_by_model
} else {
  not_available_msg("table_telemetry_llm_by_model")
}
```

### Daily Log {#daily-log}

```{r}
if (exists("table_telemetry_llm_daily") && !is.null(table_telemetry_llm_daily)) {
  table_telemetry_llm_daily
} else {
  not_available_msg("table_telemetry_llm_daily")
}
```

### Pricing Reference {#pricing-reference}

| Model | Input | Output | Cache Create | Cache Read |
|-------|-------|--------|--------------|------------|
| Opus 4.5 | $15/M | $75/M | $18.75/M | $1.875/M |
| Sonnet 4 | $3/M | $15/M | $3.75/M | $0.375/M |
| Haiku 3.5 | $0.80/M | $4/M | $1/M | $0.08/M |

Data source: Claude Code session files in `~/.claude/projects/`.

:::

## Validation Reports {#validation}

Pointblank validation reports embedded below. See also: [analysis_data Validation](validation_analysis_data.html) | [rogue_wave_events Validation](validation_rogue_events.html)

### analysis_data {#validation-analysis}

```{r}
report_path <- "../docs/articles/validation_analysis_data.html"
if (file.exists(report_path)) {
  htmltools::tags$iframe(
    src = "validation_analysis_data.html",
    width = "100%",
    height = "600px",
    style = "border: 1px solid #ccc; border-radius: 4px;"
  )
} else {
  not_available_msg("validation_analysis_data.html")
}
```

### rogue_wave_events {#validation-rogue}

```{r}
report_path <- "../docs/articles/validation_rogue_events.html"
if (file.exists(report_path)) {
  htmltools::tags$iframe(
    src = "validation_rogue_events.html",
    width = "100%",
    height = "600px",
    style = "border: 1px solid #ccc; border-radius: 4px;"
  )
} else {
  not_available_msg("validation_rogue_events.html")
}
```

---

```{r}
#| output: asis
cat("*Generated:", format(Sys.time(), "%Y-%m-%d %H:%M UTC"), "*\n")
cat("*All data loaded from pre-computed targets via tar_load()*\n")
cat("*Data range: Full historical (2019-present), NO sampling*")
```

---

::: {.callout-note appearance="minimal"}
**Disclaimer:** `irishbuoys` is an independent open-source project ([website](https://johngavin.github.io/irishbuoys/), [GitHub](https://github.com/JohnGavin/irishbuoys)). It is not affiliated with, endorsed by, or part of the [Marine Institute](https://www.marine.ie/) or the Irish Weather Buoy Network. For official marine weather warnings, consult [Met Eireann](https://www.met.ie/warnings).
:::
