---
title: "Telemetry"
format:
  dashboard:
    orientation: rows
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# This vignette performs NO computation.
# All data is loaded from pre-computed targets via tar_load().
library(targets)

# Set targets store relative to vignette location
targets::tar_config_set(store = "../_targets")

# Load all telemetry targets
# These are computed by plan_telemetry.R during tar_make()
tryCatch({
  tar_load(telemetry_summary)
  tar_load(telemetry_station_coverage)
  tar_load(telemetry_db_info)
  tar_load(telemetry_db_file_size)
  tar_load(telemetry_data_freshness)
  tar_load(telemetry_pipeline_status)
  tar_load(telemetry_qc_trends)
  tar_load(telemetry_rogue_frequency)
  tar_load(telemetry_git_commits)
  tar_load(telemetry_commit_frequency)
  tar_load(telemetry_file_types)

  # Tables
  tar_load(table_telemetry_station_coverage)
  tar_load(table_telemetry_db_info)
  tar_load(table_telemetry_freshness)
  tar_load(table_telemetry_commits)
  tar_load(table_telemetry_file_types)
  tar_load(telemetry_target_metrics)
  tar_load(table_telemetry_target_metrics)

  # Plots
  tar_load(plot_telemetry_coverage_timeline)
  tar_load(plot_telemetry_qc_trends)
  tar_load(plot_telemetry_commit_frequency)
  tar_load(plot_telemetry_rogue_frequency)

  targets_loaded <- TRUE
}, error = function(e) {
  targets_loaded <<- FALSE
})
```

# Overview

Project health and data pipeline metrics for the irishbuoys package.

## Row {height=20%}

### Column

::: {.valuebox title="Total Observations" icon="database"}
```{r}
if (exists("telemetry_summary") && !is.null(telemetry_summary)) {
  format(telemetry_summary$total_observations, big.mark = ",")
} else {
  "Run tar_make()"
}
```
:::

### Column

::: {.valuebox title="Database Size" icon="hdd"}
```{r}
if (exists("telemetry_summary") && !is.null(telemetry_summary$db_size_mb)) {
  paste0(telemetry_summary$db_size_mb, " MB")
} else {
  "N/A"
}
```
:::

### Column

::: {.valuebox title="Active Stations" icon="broadcast"}
```{r}
if (exists("telemetry_summary") && !is.null(telemetry_summary)) {
  telemetry_summary$stations_active
} else {
  "N/A"
}
```
:::

### Column

::: {.valuebox title="Commits (30d)" icon="git-branch"}
```{r}
if (exists("telemetry_summary") && !is.null(telemetry_summary)) {
  telemetry_summary$n_commits_30d
} else {
  "N/A"
}
```
:::

## Row

### Column

::: {.card title="Last Updated"}
```{r}
if (exists("telemetry_summary") && !is.null(telemetry_summary)) {
  format(telemetry_summary$generated_at, "%Y-%m-%d %H:%M UTC")
} else {
  "Targets not built. Run tar_make() to generate telemetry data."
}
```
:::

# Data Coverage

Station data availability and freshness metrics.

## Row

### Column {.tabset}

#### Coverage Timeline

```{r}
#| fig-height: 6
if (exists("plot_telemetry_coverage_timeline") && !is.null(plot_telemetry_coverage_timeline)) {
  plot_telemetry_coverage_timeline
} else {
  htmltools::p("Plot not available. Run tar_make() to generate.")
}
```

#### Station Details

```{r}
if (exists("table_telemetry_station_coverage")) {
  table_telemetry_station_coverage
} else {
  htmltools::p("Table not available. Run tar_make() to generate.")
}
```

#### Data Freshness

```{r}
if (exists("table_telemetry_freshness")) {
  table_telemetry_freshness
} else {
  htmltools::p("Table not available. Run tar_make() to generate.")
}
```

# Data Quality

QC flag trends and rogue wave detection metrics.

## Row

### Column {width=50%}

::: {.card title="QC Flag Distribution Over Time"}
```{r}
#| fig-height: 5
if (exists("plot_telemetry_qc_trends") && !is.null(plot_telemetry_qc_trends)) {
  plot_telemetry_qc_trends
} else {
  htmltools::p("Plot not available. Run tar_make() to generate.")
}
```
:::

### Column {width=50%}

::: {.card title="Rogue Wave Event Frequency"}
```{r}
#| fig-height: 5
if (exists("plot_telemetry_rogue_frequency") && !is.null(plot_telemetry_rogue_frequency)) {
  plot_telemetry_rogue_frequency
} else {
  htmltools::p("Plot not available. Run tar_make() to generate.")
}
```
:::

# Database

Storage metrics and table information.

## Row

### Column {.tabset}

#### Tables

```{r}
if (exists("table_telemetry_db_info")) {
  table_telemetry_db_info
} else {
  htmltools::p("Table not available. Run tar_make() to generate.")
}
```
#### File Size

```{r}
if (exists("telemetry_db_file_size") && nrow(telemetry_db_file_size) > 0) {
  knitr::kable(telemetry_db_file_size, format = "html")
} else {
  htmltools::p("Database file info not available.")
}
```

# Pipeline

Targets pipeline status and timing metrics.

## Row

### Column {width=40%}

::: {.card title="Target Status Summary"}
```{r}
if (exists("telemetry_pipeline_status") && !is.null(telemetry_pipeline_status$status_summary)) {
  knitr::kable(telemetry_pipeline_status$status_summary, format = "html")
} else {
  htmltools::p("Pipeline status not available.")
}
```
:::

### Column {width=60%}

::: {.card title="Pipeline Timing"}
```{r}
if (exists("telemetry_pipeline_status") && !is.null(telemetry_pipeline_status$timing)) {
  timing <- telemetry_pipeline_status$timing
  htmltools::tags$ul(
    htmltools::tags$li(paste("Total runtime:", round(timing$total_runtime_sec, 1), "seconds")),
    htmltools::tags$li(paste("Average target:", round(timing$avg_runtime_sec, 2), "seconds")),
    htmltools::tags$li(paste("Slowest target:", round(timing$max_runtime_sec, 1), "seconds")),
    htmltools::tags$li(paste("Last run:", format(timing$last_run, "%Y-%m-%d %H:%M")))
  )
} else {
  htmltools::p("Timing info not available.")
}
```
:::

## Row

### Column

::: {.card title="Target Metrics by Plan"}
Count, size (MB), and timing (seconds) for each target, grouped by tar_plan.
TOTAL row shows overall totals.

```{r}
if (exists("table_telemetry_target_metrics")) {
  table_telemetry_target_metrics
} else {
  htmltools::p("Target metrics not available. Run tar_make() to generate.")
}
```
:::

# Git Activity

Repository commit history and activity.

## Row

### Column {width=40%}

::: {.card title="Commit Frequency (Last 30 Days)"}
```{r}
#| fig-height: 4
if (exists("plot_telemetry_commit_frequency") && !is.null(plot_telemetry_commit_frequency)) {
  plot_telemetry_commit_frequency
} else {
  htmltools::p("Plot not available. Run tar_make() to generate.")
}
```
:::

### Column {width=60%}

::: {.card title="Recent Commits"}
```{r}
if (exists("table_telemetry_commits")) {
  table_telemetry_commits
} else {
  htmltools::p("Table not available. Run tar_make() to generate.")
}
```
:::

# Project

Project structure and file statistics.

## Row

### Column

::: {.card title="File Types"}
```{r}
if (exists("table_telemetry_file_types")) {
  table_telemetry_file_types
} else {
  htmltools::p("Table not available. Run tar_make() to generate.")
}
```
:::

---

```{r}
#| echo: false
cat("*Generated:", format(Sys.time(), "%Y-%m-%d %H:%M UTC"), "*\n")
cat("*All data loaded from pre-computed targets via tar_load()*")
```
