---
title: "Irish Buoys Analysis"
format:
  dashboard:
    orientation: rows
    theme: cosmo
    include-in-header:
      - text: |
          <style>
          /* Click-to-zoom: plots and tables expand to fullscreen on click */
          .fullscreen-overlay {
            position: fixed !important;
            top: 0 !important;
            left: 0 !important;
            width: 100vw !important;
            height: 100vh !important;
            z-index: 9999 !important;
            background: rgba(0,0,0,0.95) !important;
            padding: 20px !important;
            overflow: auto !important;
            cursor: zoom-out !important;
            box-sizing: border-box !important;
          }
          .fullscreen-overlay .js-plotly-plot,
          .fullscreen-overlay .plotly {
            width: 100% !important;
            height: calc(100vh - 40px) !important;
          }
          .fullscreen-overlay .datatables {
            max-height: calc(100vh - 40px) !important;
            overflow: auto !important;
          }
          /* Hint cursor on zoomable elements */
          .cell-output-display:has(.js-plotly-plot),
          .cell-output-display:has(.datatables) {
            cursor: zoom-in;
          }
          /* Tab content: show full height, page scrolls instead of card body */
          .card-body, .tab-content, .tab-pane {
            overflow: visible !important;
            max-height: none !important;
          }
          .tab-pane .cell-output-display {
            overflow: visible !important;
            max-height: none !important;
          }
          /* DT tables inside tabs: no internal scroll, show all rows */
          .tab-pane .datatables {
            overflow: visible !important;
            max-height: none !important;
          }
          /* Increase default plotly plot height */
          .js-plotly-plot, .plotly {
            min-height: 600px !important;
          }
          /* DT tables: remove internal scroll, show more content */
          .dataTables_scrollBody {
            max-height: none !important;
            overflow: visible !important;
          }
          /* Cards expand to fit content */
          .card, .card-body {
            min-height: auto !important;
            overflow: visible !important;
            max-height: none !important;
          }
          /* White captions on dark theme */
          caption, .table caption, table caption {
            color: white !important;
          }
          /* Abbr tags: keep default cursor, remove underline */
          abbr {
            cursor: default;
            text-decoration: none;
          }
          </style>
          <script>
          document.addEventListener('click', function(e) {
            // Close fullscreen on click
            var overlay = e.target.closest('.fullscreen-overlay');
            if (overlay) {
              overlay.classList.remove('fullscreen-overlay');
              // Trigger plotly resize after exiting fullscreen
              var plots = overlay.querySelectorAll('.js-plotly-plot');
              plots.forEach(function(p) { if (window.Plotly) Plotly.Plots.resize(p); });
              return;
            }
            // Open fullscreen: walk to outermost .cell-output-display (includes caption)
            var cell = e.target.closest('.cell-output-display');
            if (!cell) return;
            // Walk up to outermost container that holds plot or table
            var parent = cell.parentElement ? cell.parentElement.closest('.cell-output-display') : null;
            while (parent) {
              if (parent.querySelector('.js-plotly-plot') || parent.querySelector('.datatables')) {
                cell = parent;
              }
              parent = parent.parentElement ? parent.parentElement.closest('.cell-output-display') : null;
            }
            var hasPlot = cell.querySelector('.js-plotly-plot');
            var hasTable = cell.querySelector('.datatables');
            if (!hasPlot && !hasTable) return;
            // Don't zoom if clicking plotly modebar buttons
            if (e.target.closest('.modebar')) return;
            cell.classList.add('fullscreen-overlay');
            // Trigger plotly resize to fill fullscreen
            if (hasPlot && window.Plotly) {
              setTimeout(function() { Plotly.Plots.resize(hasPlot); }, 100);
            }
          });
          // Y-axis auto-scale: recompute y-range when rangeslider changes x-range
          (function() {
            function attachAutoScale(plot) {
              // Only attach to plots that have a rangeslider
              if (!plot.layout || !plot.layout.xaxis || !plot.layout.xaxis.rangeslider) return;
              plot.on('plotly_relayout', function(ed) {
                if (plot._yAutoScaling) return;
                var xRange;
                if (ed['xaxis.range[0]'] != null) {
                  xRange = [ed['xaxis.range[0]'], ed['xaxis.range[1]']];
                } else if (ed['xaxis.range']) {
                  xRange = ed['xaxis.range'];
                } else if (ed['xaxis.autorange']) {
                  plot._yAutoScaling = true;
                  Plotly.relayout(plot, {'yaxis.autorange': true}).then(function() {
                    plot._yAutoScaling = false;
                  });
                  return;
                } else {
                  return;
                }
                var xMin = new Date(xRange[0]).getTime();
                var xMax = new Date(xRange[1]).getTime();
                if (isNaN(xMin) || isNaN(xMax)) return;
                var yMin = Infinity, yMax = -Infinity;
                plot.data.forEach(function(trace) {
                  if (!trace.x || !trace.y) return;
                  if (trace.x.length <= 3 && trace.mode === 'lines') return;
                  for (var i = 0; i < trace.x.length; i++) {
                    var t = new Date(trace.x[i]).getTime();
                    if (t >= xMin && t <= xMax && trace.y[i] != null && isFinite(trace.y[i])) {
                      if (trace.y[i] < yMin) yMin = trace.y[i];
                      if (trace.y[i] > yMax) yMax = trace.y[i];
                    }
                  }
                });
                if (yMin !== Infinity && yMax !== -Infinity) {
                  var pad = (yMax - yMin) * 0.05 || 0.1;
                  plot._yAutoScaling = true;
                  Plotly.relayout(plot, {'yaxis.range': [yMin - pad, yMax + pad]}).then(function() {
                    plot._yAutoScaling = false;
                  });
                }
              });
            }
            // Attach to all plotly plots once they render
            setTimeout(function() {
              document.querySelectorAll('.js-plotly-plot').forEach(function(p) {
                if (p.on && !p._yAutoScaleAttached) {
                  p._yAutoScaleAttached = true;
                  attachAutoScale(p);
                }
              });
            }, 3000);
          })();
          // Auto-activate parent tabs on hash navigation (fixes #extreme-value-methods etc.)
          function activateHashTarget() {
            if (!location.hash) return;
            var target = document.getElementById(location.hash.slice(1));
            if (!target) return;
            var pane = target.closest('.tab-pane');
            while (pane) {
              var tabId = pane.getAttribute('id');
              var tabTrigger = document.querySelector('[data-bs-target="#' + tabId + '"], [href="#' + tabId + '"]');
              if (tabTrigger) {
                var tab = new bootstrap.Tab(tabTrigger);
                tab.show();
              }
              pane = pane.parentElement ? pane.parentElement.closest('.tab-pane') : null;
            }
            setTimeout(function() { target.scrollIntoView({behavior: 'smooth'}); }, 200);
          }
          window.addEventListener('hashchange', activateHashTarget);
          document.addEventListener('DOMContentLoaded', activateHashTarget);
          </script>
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

library(targets)
library(ggplot2)
library(dplyr)
library(ggplot2)
library(plotly)
library(DT)
library(dygraphs)
library(xts)
library(tidyr)

# Load package functions for vignette rendering
# This allows create_plot_* functions to work
if (requireNamespace("devtools", quietly = TRUE)) {
  # Get the project root from the _targets store config
  proj_root <- sub("/vignettes$", "", getwd())
  if (!file.exists(file.path(proj_root, "DESCRIPTION"))) {
    proj_root <- dirname(dirname(getwd()))
  }
  devtools::load_all(proj_root, export_all = FALSE, quiet = TRUE)
}

theme_set(theme_minimal(base_size = 12))

# Set targets store - works from project root or vignettes/
store_path <- if (dir.exists("_targets")) "_targets" else "../_targets"
targets::tar_config_set(store = store_path)

# Load targets data only - generate plots inline to keep file size small
# NOTE: Pre-computed plotly objects embed all data as JSON (20MB+)
#       Generating inline from summary data keeps file ~1-2MB
targets::tar_load(c(
  missing_data_grid,
  rogue_wave_events,
  rogue_wave_conditions,
  rogue_wave_statistics,
  analysis_summary,
  analysis_data,
  seasonal_means_wave,
  seasonal_means_wind,
  wave_height_seasonal,
  annual_trends_wave,
  annual_trends_wind,
  gpd_wave_per_station,
  gpd_wind_per_station,
  gpd_hmax_per_station,
  gev_wave_pooled,
  gev_wind_pooled,
  gev_hmax_pooled,
  return_levels_wave_pooled,
  return_levels_wind_pooled,
  return_levels_hmax_pooled,
  gust_factor_analysis,
  return_level_curves_wave,
  return_level_curves_wind
))

# Load new per-station targets (may not exist in older pipeline runs)
gpd_gust_per_station <- tryCatch(
  targets::tar_read(gpd_gust_per_station),
  error = function(e) NULL
)
return_levels_per_station <- tryCatch(
  targets::tar_read(return_levels_per_station),
  error = function(e) NULL
)
ci_comparison_per_station <- tryCatch(
  targets::tar_read(ci_comparison_per_station),
  error = function(e) NULL
)

# Load pre-computed vignette display targets
tryCatch({
  targets::tar_load(c(
    wave_vignette_missing_summary,
    wave_vignette_rogue_table,
    wave_vignette_rl_wide,
    wave_vignette_gpd_params,
    wave_vignette_threshold_sens,
    wave_vignette_gev_pooled_params
  ))
}, error = function(e) {
  # Fallback: targets may not exist in older pipeline runs
  wave_vignette_missing_summary <<- NULL
  wave_vignette_rogue_table <<- NULL
  wave_vignette_rl_wide <<- NULL
  wave_vignette_gpd_params <<- NULL
  wave_vignette_threshold_sens <<- NULL
  wave_vignette_gev_pooled_params <<- NULL
})
wave_vignette_ci_comparison <- tryCatch(
  targets::tar_read(wave_vignette_ci_comparison),
  error = function(e) NULL
)

# Alias for compatibility
rogue_events <- rogue_wave_events
rogue_conditions <- rogue_wave_conditions
wave_stl <- wave_height_seasonal
gust_analysis <- gust_factor_analysis

# For return levels, use pooled GEV (illustrative)
return_levels_wave <- return_levels_wave_pooled
return_levels_wind <- return_levels_wind_pooled
return_levels_hmax <- return_levels_hmax_pooled

has_data <- !is.null(rogue_events) && nrow(rogue_events) > 0

# Get dataset date range for captions
if (!is.null(analysis_summary) && "data_summary" %in% names(analysis_summary)) {
  data_start <- format(analysis_summary$data_summary$date_range[1], "%Y-%m-%d")
  data_end <- format(analysis_summary$data_summary$date_range[2], "%Y-%m-%d")
  date_caption <- paste0(data_start, " to ", data_end)
} else if (has_data) {
  data_start <- format(min(rogue_events$time), "%Y-%m-%d")
  data_end <- format(max(rogue_events$time), "%Y-%m-%d")
  date_caption <- paste0(data_start, " to ", data_end)
} else {
  data_start <- "N/A"
  data_end <- "N/A"
  date_caption <- "N/A"
}

# Get package version
pkg_version <- tryCatch(
  as.character(packageVersion("irishbuoys")),
  error = function(e) {
    desc_file <- "../DESCRIPTION"
    if (file.exists(desc_file)) {
      desc <- read.dcf(desc_file)
      desc[1, "Version"]
    } else "dev"
  }
)
generated_time <- format(Sys.time(), "%Y-%m-%d %H:%M UTC")

# Helper function for formatting numbers with commas
fmt_num <- function(x, digits = 0) {
  if (is.na(x) || is.null(x)) return("N/A")
  format(round(x, digits), big.mark = ",", nsmall = digits)
}

# Helper function for DT tables (caption at top for accessibility)
create_dt <- function(data, caption = "") {
  time_cols <- which(names(data) %in% c("time", "Time", "timestamp", "Timestamp"))

  # Force caption to top position with white text on dark theme
  cap <- if (nzchar(caption)) {
    htmltools::tags$caption(style = "caption-side: top; font-size: 0.9em; color: white;", caption)
  } else {
    NULL
  }

  dt <- datatable(
    data, caption = cap, filter = 'top', rownames = FALSE, extensions = 'Buttons',
    options = list(
      pageLength = 25, dom = 'Bfrtip', buttons = c('copy', 'csv', 'excel'),
      order = list(list(0, 'desc')), scrollX = TRUE,
      columnDefs = if (length(time_cols) > 0) list(list(width = '120px', targets = time_cols - 1)) else NULL
    ),
    class = 'cell-border stripe hover compact'
  )

  # No text wrapping in any column — horizontal scroll preferred
  dt <- dt %>% DT::formatStyle(columns = seq_len(ncol(data)), whiteSpace = 'nowrap')

  if (length(time_cols) > 0) {
    dt <- dt %>% DT::formatStyle(columns = time_cols, fontSize = '10px')
  }
  dt
}

# Beaufort scale classification (wind speed in m/s)
classify_beaufort <- function(ws) {
  cut(ws,
    breaks = c(-Inf, 10.7, 17.1, 24.4, 28.4, 32.6, Inf),
    labels = c("0-5 (Calm-Fresh)", "6-7 (Strong-Near Gale)",
               "8-9 (Gale-Severe)", "10 (Storm)",
               "11 (Violent Storm)", "12 (Hurricane)"),
    right = TRUE
  )
}

beaufort_colors <- c(
  "0-5 (Calm-Fresh)" = "#2166ac",
  "6-7 (Strong-Near Gale)" = "#4393c3",
  "8-9 (Gale-Severe)" = "#d6604d",
  "10 (Storm)" = "#f4a582",
  "11 (Violent Storm)" = "#b2182b",
  "12 (Hurricane)" = "#67001f"
)

# Use irishbuoys_layout() for all plotly plots (dark theme, bottom legend)
# Defined in R/plotly_helpers.R - no local legend_style needed

# Compute additional metrics from analysis_data
if (!is.null(analysis_data) && nrow(analysis_data) > 0) {
  max_mean_wave <- max(analysis_data$wave_height, na.rm = TRUE)
  max_gust <- max(analysis_data$gust, na.rm = TRUE)
  max_mean_wind <- max(analysis_data$wind_speed, na.rm = TRUE)
} else {
  max_mean_wave <- NA
  max_gust <- NA
  max_mean_wind <- NA
}

# Dynamic title based on actual data range
year_start <- format(as.Date(data_start), "%Y")
year_end <- format(as.Date(data_end), "%y")
dynamic_title <- paste0("Irish Buoys Analysis (", year_start, "-", year_end, ")")
```

# Overview

## Key Metrics

### Data Overview

**Wave Analysis Dashboard** for the [Irish Weather Buoy Network](https://www.marine.ie/site-area/infrastructure-facilities/marine-institute-buoy-network){target="_blank"}.

::: {.callout-warning appearance="minimal"}
This website is an independent open-source analysis tool and is **not** part of or affiliated with the Irish Weather Buoy Network or the [Marine Institute](https://www.marine.ie/){target="_blank"}.
:::

#### Data Structure

```{r overview-data-structure}
#| echo: false
#| results: asis
if (!is.null(analysis_summary) && "data_summary" %in% names(analysis_summary)) {
  n_days <- as.numeric(difftime(
    analysis_summary$data_summary$date_range[2],
    analysis_summary$data_summary$date_range[1],
    units = "days"
  ))
  start_date <- format(analysis_summary$data_summary$date_range[1], "%Y-%m-%d")
  end_date <- format(analysis_summary$data_summary$date_range[2], "%Y-%m-%d")
  n_cols <- if (!is.null(analysis_data)) ncol(analysis_data) else 12
  n_raw_cols <- nrow(get_data_dictionary())

  cat(sprintf("- Primary data: %s days\n- %s to %s\n- %d of %d ERDDAP columns selected for analysis",
              fmt_num(round(n_days)), start_date, end_date, n_cols, n_raw_cols))
}
```

### Analysis Summary

#### Key Metrics

```{r key-metrics-table}
#| echo: false
# Build metrics table with hover tooltips linking to definitions
metrics <- data.frame(
  metric = c(
    '<abbr title="Events where Max Wave / Signif Wave > 2.0">Rogue Wave Events</abbr>',
    '<abbr title="Maximum individual wave height (Hmax) recorded across all stations">Max Indiv Wave</abbr>',
    '<abbr title="Highest Max Wave / Signif Wave ratio observed">Max Rogue Ratio</abbr>',
    '<abbr title="Highest hourly mean wave height (Signif Wave = 4σ)">Max Mean Wave</abbr>',
    '<abbr title="Highest peak 3-second wind gust recorded">Max Gust</abbr>',
    '<abbr title="Highest 10-minute mean wind speed recorded">Max Mean Wind</abbr>',
    "Stations Analysed",
    "Number Days",
    "Start",
    "End"
  ),
  value = c(
    if (has_data) fmt_num(nrow(rogue_events)) else "N/A",
    if (has_data) fmt_num(max(rogue_events$hmax), 1) else "N/A",
    if (has_data) fmt_num(max(rogue_events$rogue_ratio), 2) else "N/A",
    fmt_num(max_mean_wave, 1),
    fmt_num(max_gust, 1),
    fmt_num(max_mean_wind, 1),
    if (has_data) fmt_num(length(unique(rogue_events$station_id))) else "N/A",
    if (has_data) fmt_num(round(as.numeric(difftime(max(rogue_events$time), min(rogue_events$time), units = "days")))) else "N/A",
    if (has_data) format(min(rogue_events$time), "%Y-%m-%d") else "N/A",
    if (has_data) format(max(rogue_events$time), "%Y-%m-%d") else "N/A"
  ),
  units = c("count", "m", "ratio", "m", "m/s", "m/s", "count", "days", "date", "date"),
  stringsAsFactors = FALSE
)
# Interactive DT table with caption, sortable columns, hover tooltips
DT::datatable(
  metrics,
  caption = htmltools::tags$caption(
    style = "caption-side: top; font-size: 1.1em; font-weight: bold;",
    "Key Metrics — Rogue wave events, extremes, and analysis period. ",
    "Hover over metric names for definitions. ",
    "See ", htmltools::tags$a(href = "#definitions", "Reference"),
    " for full glossary."
  ),
  escape = FALSE,
  colnames = c("Metric", "Value", "Units"),
  options = list(
    dom = "t",
    pageLength = 10,
    ordering = TRUE,
    columnDefs = list(
      list(className = "dt-right", targets = 1)
    )
  ),
  rownames = FALSE,
  class = "display compact"
) |>
  DT::formatStyle(columns = seq_len(ncol(metrics)), `white-space` = "nowrap")
```

*Analysis generated `r generated_time` using irishbuoys v`r pkg_version`*

## Station Summary

### Data Availability

#### Missing Data by Station

```{r overview-missing-data}
#| echo: false
if (!is.null(wave_vignette_missing_summary)) {
  create_dt(wave_vignette_missing_summary, "Data availability by station (% of days with observations)")
} else {
  htmltools::div(class = "alert alert-warning", "Missing data grid not available. Run targets::tar_make()")
}
```

# Rogue Waves

**Rogue Wave Definition:** Max Wave / Signif Wave > 2.0. See [definitions](#definitions) and [methodology](#methodology) for details.

## All Events

### Rogue Wave Views {.tabset}

#### All Stations

```{r rogue-waves-all}
# Generate plot inline to keep file size small
create_plot_rogue_all(rogue_events)
```

#### M2

```{r rogue-waves-m2}
if (has_data && "M2" %in% rogue_events$station_id) {
  m2_events <- rogue_events |> filter(station_id == "M2")
  if (nrow(m2_events) > 0) {
    plot_ly(m2_events, x = ~time, y = ~rogue_ratio, type = "scatter", mode = "markers",
            marker = list(size = 8, color = "#d62728"),
            hoverinfo = "text",
            text = ~paste0("Time: ", format(time, "%Y-%m-%d %H:%M"),
                          "<br>Ratio: ", round(rogue_ratio, 2),
                          "<br>Max Wave: ", round(hmax, 1), "m",
                          "<br>Signif Wave: ", round(wave_height, 1), "m")) |>
      irishbuoys_layout(title = paste0("M2 Rogue Events (n=", nrow(m2_events), ")"),
             xaxis = list(title = "Time"), yaxis = list(title = "Rogue Ratio"),
             hovermode = "closest")
  } else {
    plotly_empty() |> layout(title = "M2: No rogue events")
  }
} else {
  plotly_empty() |> layout(title = "M2: No data")
}
```

#### M3

```{r rogue-waves-m3}
if (has_data && "M3" %in% rogue_events$station_id) {
  m3_events <- rogue_events |> filter(station_id == "M3")
  if (nrow(m3_events) > 0) {
    plot_ly(m3_events, x = ~time, y = ~rogue_ratio, type = "scatter", mode = "markers",
            marker = list(size = 8, color = "#1f77b4"),
            hoverinfo = "text",
            text = ~paste0("Time: ", format(time, "%Y-%m-%d %H:%M"),
                          "<br>Ratio: ", round(rogue_ratio, 2),
                          "<br>Max Wave: ", round(hmax, 1), "m",
                          "<br>Signif Wave: ", round(wave_height, 1), "m")) |>
      irishbuoys_layout(title = paste0("M3 Rogue Events (n=", nrow(m3_events), ")"),
             xaxis = list(title = "Time"), yaxis = list(title = "Rogue Ratio"),
             hovermode = "closest")
  } else {
    plotly_empty() |> layout(title = "M3: No rogue events")
  }
} else {
  plotly_empty() |> layout(title = "M3: No data")
}
```

#### M4

```{r rogue-waves-m4}
if (has_data && "M4" %in% rogue_events$station_id) {
  m4_events <- rogue_events |> filter(station_id == "M4")
  if (nrow(m4_events) > 0) {
    plot_ly(m4_events, x = ~time, y = ~rogue_ratio, type = "scatter", mode = "markers",
            marker = list(size = 8, color = "#2ca02c"),
            hoverinfo = "text",
            text = ~paste0("Time: ", format(time, "%Y-%m-%d %H:%M"),
                          "<br>Ratio: ", round(rogue_ratio, 2),
                          "<br>Max Wave: ", round(hmax, 1), "m",
                          "<br>Signif Wave: ", round(wave_height, 1), "m")) |>
      irishbuoys_layout(title = paste0("M4 Rogue Events (n=", nrow(m4_events), ")"),
             xaxis = list(title = "Time"), yaxis = list(title = "Rogue Ratio"),
             hovermode = "closest")
  } else {
    plotly_empty() |> layout(title = "M4: No rogue events")
  }
} else {
  plotly_empty() |> layout(title = "M4: No data")
}
```

#### M5

```{r rogue-waves-m5}
if (has_data && "M5" %in% rogue_events$station_id) {
  m5_events <- rogue_events |> filter(station_id == "M5")
  if (nrow(m5_events) > 0) {
    plot_ly(m5_events, x = ~time, y = ~rogue_ratio, type = "scatter", mode = "markers",
            marker = list(size = 8, color = "#ff7f0e"),
            hoverinfo = "text",
            text = ~paste0("Time: ", format(time, "%Y-%m-%d %H:%M"),
                          "<br>Ratio: ", round(rogue_ratio, 2),
                          "<br>Max Wave: ", round(hmax, 1), "m",
                          "<br>Signif Wave: ", round(wave_height, 1), "m")) |>
      irishbuoys_layout(title = paste0("M5 Rogue Events (n=", nrow(m5_events), ")"),
             xaxis = list(title = "Time"), yaxis = list(title = "Rogue Ratio"),
             hovermode = "closest")
  } else {
    plotly_empty() |> layout(title = "M5: No rogue events")
  }
} else {
  plotly_empty() |> layout(title = "M5: No data")
}
```

#### M6

```{r rogue-waves-m6}
if (has_data && "M6" %in% rogue_events$station_id) {
  m6_events <- rogue_events |> filter(station_id == "M6")
  if (nrow(m6_events) > 0) {
    plot_ly(m6_events, x = ~time, y = ~rogue_ratio, type = "scatter", mode = "markers",
            marker = list(size = 8, color = "#9467bd"),
            hoverinfo = "text",
            text = ~paste0("Time: ", format(time, "%Y-%m-%d %H:%M"),
                          "<br>Ratio: ", round(rogue_ratio, 2),
                          "<br>Max Wave: ", round(hmax, 1), "m",
                          "<br>Signif Wave: ", round(wave_height, 1), "m")) |>
      irishbuoys_layout(title = paste0("M6 Rogue Events (n=", nrow(m6_events), ")"),
             xaxis = list(title = "Time"), yaxis = list(title = "Rogue Ratio"),
             hovermode = "closest")
  } else {
    plotly_empty() |> layout(title = "M6: No rogue events")
  }
} else {
  plotly_empty() |> layout(title = "M6: No data")
}
```

#### Faceted View

Use the **range slider** at the bottom to zoom all panels to a specific time period.

```{r rogue-waves-by-station}
#| fig-height: 14
# Generate plot inline to keep file size small
create_plot_rogue_by_station(rogue_events)
```

#### Data Table

**rogue_ratio** = Max Wave / Signif Wave. Values > 2.0 define a rogue wave event.
**gust_ratio** = Gust / Wind. Values > 1.5 indicate a rogue gust.

```{r rogue-waves-table}
if (!is.null(wave_vignette_rogue_table)) {
  create_dt(wave_vignette_rogue_table, paste0(
    "Rogue Wave Events (rogue_ratio > 2.0). ",
    format(min(rogue_events$time), "%Y-%m-%d"), " to ",
    format(max(rogue_events$time), "%Y-%m-%d")
  ))
}
```

# Rogue Gusts

**Rogue Gust Definition:** A rogue gust has **Gust / Wind > 1.5** (ratio of peak gust to [sustained wind](#definitions) speed). See [definitions](#definitions). This parallels the rogue wave definition (Max Wave / Signif Wave > 2.0).

## Gust Analysis

### Rogue Gust Views {.tabset}

#### All Stations

```{r rogue-gusts-all}
# Generate plot inline - load rogue gust events first
targets::tar_load(rogue_gust_events)
if (!is.null(rogue_gust_events) && nrow(rogue_gust_events) > 0) {
  create_plot_rogue_gusts_all(rogue_gust_events)
}
```

#### Events by Station

Use the **range slider** at the bottom to zoom all panels to a specific time period.

```{r rogue-gusts-by-station}
#| fig-height: 14
# Generate plot inline
if (!is.null(rogue_gust_events) && nrow(rogue_gust_events) > 0) {
  create_plot_rogue_gusts_by_station(rogue_gust_events)
}
```

#### Data Table

**gust_ratio** = Gust / Wind. Values > 1.5 define a rogue gust event.

```{r rogue-gusts-table}
targets::tar_load(rogue_gust_events)
if (!is.null(rogue_gust_events) && nrow(rogue_gust_events) > 0) {
  n_total <- nrow(rogue_gust_events)
  rogue_gust_events |>
    arrange(desc(gust_ratio)) |>
    head(500) |>
    mutate(
      time = format(time, "%Y-%m-%d %H:%M"),
      across(where(is.numeric), ~round(., 2))
    ) |>
    create_dt(paste0("Top 500 Rogue Gust Events of ",
                     format(n_total, big.mark = ","),
                     " total (gust_ratio > 1.5). ",
                     format(min(rogue_gust_events$time), "%Y-%m-%d"), " to ",
                     format(max(rogue_gust_events$time), "%Y-%m-%d")))
}
```

#### vs Rogue Waves

Comparing wave rogue events (Max Wave / Signif Wave > 2.0) with gust rogue events (Gust / Wind > 1.5).

```{r rogue-gusts-vs-waves}
# Generate plot inline - need analysis_data for this
targets::tar_load(analysis_data)
create_plot_gusts_vs_waves(analysis_data)
```

# Gust Factor

**[Gust Factor](#definitions)** = Peak Gust / [Sustained Wind](#definitions). Typical value ~1.3. See [methodology](#methodology) for measurement details.
**Rogue Gust** threshold: GF > 1.5 (see [Rogue Gusts](#rogue-gusts) page for detailed analysis).

## Factor Analysis

### Gust Factor Views {.tabset}

#### Overall Summary

```{r gust-overall}
if (!is.null(gust_analysis) && "summary" %in% names(gust_analysis)) {
  summary_df <- gust_analysis$summary |>
    mutate(across(where(is.numeric), ~round(., 2)))
  # Right-justify numeric columns and narrow layout
  numeric_cols <- which(vapply(summary_df, is.numeric, logical(1))) - 1L
  DT::datatable(
    summary_df,
    caption = htmltools::tags$caption(
      style = "caption-side: top; font-size: 0.9em; color: white;",
      "Gust Factor Statistics (All Stations Combined)"
    ),
    rownames = FALSE, class = "cell-border stripe hover compact",
    options = list(
      pageLength = 25, dom = "t",
      columnDefs = list(
        list(className = "dt-right", targets = numeric_cols)
      )
    )
  )
} else {
  htmltools::div(class = "alert alert-warning", "Gust analysis not available. Run targets::tar_make()")
}
```

#### By Station

```{r gust-by-station}
if (!is.null(gust_analysis) && "by_station" %in% names(gust_analysis) && !is.null(gust_analysis$by_station)) {
  gust_analysis$by_station |>
    mutate(across(where(is.numeric), ~round(., 2))) |>
    create_dt("Gust Factor Statistics by Station. Stations with sufficient data for analysis.")
} else {
  htmltools::div(class = "alert alert-warning", "Station data not available")
}
```

#### By Wind Category

```{r gust-by-category}
# Generate plot inline
create_plot_gust_by_category(gust_analysis)
```

#### Wind Category (Data)

```{r gust-category-data}
if (!is.null(gust_analysis) && "by_station_category" %in% names(gust_analysis) && !is.null(gust_analysis$by_station_category)) {
  gust_analysis$by_station_category |>
    mutate(across(where(is.numeric), ~round(., 2))) |>
    create_dt("Gust Factor by Wind Category and Station")
} else if (!is.null(gust_analysis) && "by_category" %in% names(gust_analysis)) {
  gust_analysis$by_category |>
    mutate(across(where(is.numeric), ~round(., 2))) |>
    create_dt("Gust Factor by Wind Category")
}
```

#### Rogue Gusts

```{r gust-rogue}
# Generate plot inline
create_plot_rogue_gusts(gust_analysis)
```

# Hourly Max

Extreme value analysis based on **hourly maximum** observations. Two approaches are compared: Model fitting ([GEV/GPD](#definitions)) and Predictions (return levels). See [extreme value methods](#extreme-value-methods) for statistical details.

## Hourly Patterns

### Extreme Value Views {.tabset}

#### Return Levels: All Stations

```{r eva-return-levels-all}
if (!is.null(wave_vignette_rl_wide)) {
  create_dt(wave_vignette_rl_wide, "Per-Station GPD Return Levels (95th percentile threshold)")
} else {
  htmltools::div(class = "alert alert-warning", "Return levels not available. Run targets::tar_make()")
}
```

#### Return Levels: Avg Wave

```{r eva-return-levels-wave}
create_plot_return_levels_per_station(return_levels_per_station, "avg_wave")
```

#### Return Levels: Rogue Wave

```{r eva-return-levels-rogue}
create_plot_return_levels_per_station(return_levels_per_station, "rogue_wave")
```

#### Return Levels: Avg Wind

```{r eva-return-levels-wind}
create_plot_return_levels_per_station(return_levels_per_station, "avg_wind")
```

#### Return Levels: Wind Gust

```{r eva-return-levels-gust}
create_plot_return_levels_per_station(return_levels_per_station, "wind_gust")
```

#### GPD Parameters

```{r eva-gpd-params}
if (!is.null(wave_vignette_gpd_params) && nrow(wave_vignette_gpd_params) > 0) {
  create_dt(wave_vignette_gpd_params, "GPD Parameters (95% threshold) - All Variables and Stations")
} else {
  htmltools::div(class = "alert alert-warning", "GPD parameters not available")
}
```

#### CI Comparison

Comparison of confidence interval methods for GPD return levels: **delta-method** (asymptotic, from standard errors) vs **block bootstrap** (resampled, 500 replicates, block size 48h). Format: *estimate [lower, upper]*.

```{r eva-ci-comparison}
if (!is.null(wave_vignette_ci_comparison) && nrow(wave_vignette_ci_comparison) > 0) {
  create_dt(wave_vignette_ci_comparison, "CI Comparison: Delta-method vs Bootstrap (95% level)")
} else {
  htmltools::div(class = "alert alert-info", "CI comparison not yet available. Run the pipeline with updated targets.")
}
```

#### Theory and Method

The **Generalized Pareto Distribution** has CDF for exceedances $y = x - u$ above threshold $u$:

$$F(y) = 1 - \left(1 + \frac{\xi y}{\sigma}\right)^{-1/\xi}$$

with two parameters:

| Parameter | Symbol | Interpretation |
|-----------|--------|----------------|
| **Scale** | $\sigma$ | Spread of exceedances |
| **Shape** | $\xi$ | Tail behaviour (same as GEV) |

**Return levels** for T-year event:

$$z_T = u + \frac{\sigma}{\xi}\left[(n_y \lambda T)^\xi - 1\right]$$

where $n_y$ = observations per year, $\lambda$ = exceedance rate.

**Threshold sensitivity:** GPD results depend on threshold choice. Lower thresholds give more data but may violate asymptotic theory. The [mev](https://cran.r-project.org/package=mev){target="_blank"} package provides diagnostics.

```{r eva-threshold-sensitivity}
if (!is.null(wave_vignette_threshold_sens) && nrow(wave_vignette_threshold_sens) > 0) {
  create_dt(wave_vignette_threshold_sens, "Threshold sensitivity for M3 wave height (hourly data)")
}
```

#### GEV Pooled (Illustrative)

**Note: Limited sample size (n~8 years).** The pooled GEV analysis below is illustrative only due to few annual maxima. The per-station GPD approach above is preferred.

```{r eva-gev-pooled-illustrative}
if (!is.null(wave_vignette_gev_pooled_params)) {
  create_dt(wave_vignette_gev_pooled_params, "GEV Parameter Estimates - Pooled (illustrative, n~8)")
} else {
  htmltools::div(class = "alert alert-warning", "GEV fit data not available")
}
```

**Bayesian approaches:** The [mev](https://cran.r-project.org/package=mev){target="_blank"} package supports profile likelihood and Bayesian methods for improved uncertainty quantification.

**Storm propagation:** Cross-correlation lag analysis between stations could enable offshore-to-nearshore prediction with lead times.

# How Signif Wave is Calculated

This page shows a step-by-step worked example of how **Signif Wave** (significant wave height) is calculated from raw buoy measurements during one 17.5-minute sampling window.

## Worked Example

### Step-by-Step Example {width=100%}

```{r worked-example}
#| echo: false
# Load pre-computed example from targets
targets::tar_load(worked_signif_wave_example)

if (!is.null(worked_signif_wave_example)) {
  example <- worked_signif_wave_example

  # Show first and last 5 rows
  waves_display <- rbind(
    head(example$waves_table, 5),
    tibble::tibble(wave_number = NA, wave_height_m = NA, in_top_third = NA),
    tail(example$waves_table, 5)
  )

  DT::datatable(waves_display,
    caption = "Step 1: Buoy records ~100-300 individual waves during 17.5-min window. Shown: First and last 5 waves (sorted by height, tallest first)",
    colnames = c("Rank", "Wave Height (m)", "In Top 1/3?"),
    extensions = "Buttons",
    options = list(dom = "Bfrtip", buttons = c("csv", "excel", "print"),
                   pageLength = 15, scrollX = TRUE),
    rownames = FALSE, class = "display compact"
  )
}
```

### Calculation Summary {width=100%}

**Step 2: Identify the highest 1/3 of waves**

Total waves: 100
Top 1/3 count: `r if(!is.null(worked_signif_wave_example)) worked_signif_wave_example$n_top_third else "N/A"` waves

**Step 3: Calculate Signif Wave = mean of top 1/3**

Signif Wave = mean(top 34 waves) = **`r if(!is.null(worked_signif_wave_example)) worked_signif_wave_example$signif_wave else "N/A"` m**

**Step 4: Max Wave = tallest single wave**

Max Wave = **`r if(!is.null(worked_signif_wave_example)) worked_signif_wave_example$max_wave else "N/A"` m**

**Step 5: Calculate Ratio**

Ratio = Max Wave / Signif Wave = `r if(!is.null(worked_signif_wave_example)) paste0(worked_signif_wave_example$max_wave, " / ", worked_signif_wave_example$signif_wave, " = **", worked_signif_wave_example$ratio, "**") else "N/A"`

`r if(!is.null(worked_signif_wave_example) && worked_signif_wave_example$ratio > 2.0) "**Rogue Wave Detected!** Ratio > 2.0" else if(!is.null(worked_signif_wave_example)) "**Normal Sea State.** Ratio in typical range (1.5-1.9)" else ""`

*`r if(!is.null(worked_signif_wave_example)) worked_signif_wave_example$caption else "Worked example not available. Run targets::tar_make()"`*

# Joint Analysis

**Question:** Does M6 (furthest offshore, 320km) predict wave conditions at other stations hours in advance?

```{r joint-setup}
#| include: false
# Load joint analysis targets
targets::tar_load(c(
  joint_analysis_results,
  station_distances,
  pair_correlations_wave,
  m6_prediction_summary,
  joint_extremes_wave,
  copula_summary
), store = "../_targets")

has_joint_data <- !is.null(joint_analysis_results)
```

## Cross-Station

### Joint Analysis Views {.tabset}

#### Station Distances

**Station positions and distances (km):**

```{r joint-distances}
#| echo: false
if (has_joint_data) {
  station_info <- joint_analysis_results$station_info
  dist_matrix <- joint_analysis_results$station_distances

  # Create a nice display table
  dist_df <- as.data.frame(dist_matrix)
  dist_df$Station <- rownames(dist_df)
  dist_df <- dist_df[, c("Station", colnames(dist_matrix))]

  create_dt(dist_df, "Distances between stations (km). M6 is 320km offshore in deep Atlantic (3000m depth).")
} else {
  htmltools::div(class = "alert alert-warning", "Joint analysis data not available. Run targets::tar_make()")
}
```

**Station Map (conceptual positions):**

```{r joint-map}
#| echo: false
#| fig-height: 5
if (has_joint_data) {
  station_info <- joint_analysis_results$station_info

  p <- ggplot(station_info, aes(x = lon, y = lat)) +
    geom_point(aes(size = depth_m, color = distance_km), alpha = 0.8) +
    geom_text(aes(label = station_id), vjust = -1.5, size = 4, fontface = "bold") +
    scale_size_continuous(name = "Depth (m)", range = c(4, 12)) +
    scale_color_viridis_c(name = "Distance\nOffshore (km)", option = "plasma") +
    labs(
      title = "Irish Weather Buoy Network",
      subtitle = "M6 is 320km offshore in deep Atlantic; others are on continental shelf",
      x = "Longitude", y = "Latitude"
    ) +
    theme_minimal(base_size = 12) +
    coord_fixed(ratio = 1.5)

  irishbuoys_ggplotly(p)
}
```

#### Cross-Correlations

**Cross-correlation between station pairs (wave height):**

Positive optimal lag means the first station leads (predicts) the second.

```{r joint-ccf}
#| echo: false
if (has_joint_data && !is.null(pair_correlations_wave)) {
  ccf_df <- pair_correlations_wave |>
    mutate(
      `Station Pair` = paste(station1, "→", station2),
      `Distance (km)` = round(distance_km, 0),
      `Optimal Lag (h)` = optimal_lag,
      `Max Correlation` = round(max_correlation, 3),
      `Expected Lag (h)` = expected_lag,
      `Observations` = format(n_obs, big.mark = ",")
    ) |>
    select(`Station Pair`, `Distance (km)`, `Optimal Lag (h)`, `Max Correlation`,
           `Expected Lag (h)`, `Observations`) |>
    arrange(desc(`Max Correlation`))

  create_dt(ccf_df, paste0(
    "Cross-correlations between station pairs. ",
    "Optimal Lag = hours at maximum correlation. ",
    "Expected Lag assumes ~30 km/h wave propagation (typical swell speed). ",
    "High correlations between nearby stations (M2-M3) suggest shared storm exposure."
  ))
}
```

#### M6 Predictions

**Can M6 (offshore) predict wave heights at coastal stations?**

M6 is 320km offshore in the deep Atlantic. Waves reaching M6 will propagate toward coastal stations over several hours.

```{r joint-m6-pred}
#| echo: false
if (has_joint_data && !is.null(m6_prediction_summary)) {
  pred_df <- m6_prediction_summary |>
    mutate(
      `Target Station` = target_station,
      `Distance from M6 (km)` = round(distance_km, 0),
      `Optimal Lag (hours)` = lag_hours,
      `R²` = round(r_squared, 3),
      `RMSE (m)` = round(rmse, 2),
      `Observations` = format(n_obs, big.mark = ",")
    ) |>
    select(`Target Station`, `Distance from M6 (km)`, `Optimal Lag (hours)`,
           `R²`, `RMSE (m)`, `Observations`) |>
    arrange(desc(`R²`))

  create_dt(pred_df, paste0(
    "M6 wave height predicting coastal stations. ",
    "Lag = hours before target observation. ",
    "R² indicates prediction skill (1.0 = perfect). ",
    "Higher R² for stations closer to M6's Atlantic exposure."
  ))
}
```

**Interpretation:**

- **Positive lag** means M6 observations *precede* coastal observations
- **R² > 0.5** indicates useful predictive skill
- Stations facing the Atlantic (M5, M2, M3) should have better predictions
- M4 (southeast) may have lower skill due to different storm exposure

#### Joint Extremes

**Do extreme waves occur simultaneously across stations?**

Extreme = top 5% of wave heights at each station.

```{r joint-extremes}
#| echo: false
if (has_joint_data && !is.null(joint_extremes_wave)) {
  # Joint counts matrix
  joint_counts <- joint_extremes_wave$joint_extreme_counts

  # Convert to data frame for display
  counts_df <- as.data.frame(joint_counts)
  counts_df$Station <- rownames(counts_df)
  counts_df <- counts_df[, c("Station", colnames(joint_counts))]

  create_dt(counts_df, paste0(
    "Joint extreme event counts. Diagonal = total extreme events per station. ",
    "Off-diagonal = simultaneous extreme events at both stations. ",
    "High off-diagonal values indicate spatially correlated extremes (storm events)."
  ))
}
```

**Conditional Probabilities:**

P(Station B extreme | Station A extreme) - probability of extreme at B given extreme at A:

```{r joint-cond-prob}
#| echo: false
if (has_joint_data && !is.null(joint_extremes_wave)) {
  cond_probs <- joint_extremes_wave$conditional_probs

  # Convert to data frame
  prob_df <- as.data.frame(cond_probs)
  prob_df$`If Extreme At` <- rownames(prob_df)
  prob_df <- prob_df[, c("If Extreme At", colnames(cond_probs))]

  # Format as percentages
  for (col in colnames(cond_probs)) {
    prob_df[[col]] <- paste0(round(prob_df[[col]] * 100, 1), "%")
  }

  create_dt(prob_df, paste0(
    "Conditional probabilities: P(column station extreme | row station extreme). ",
    "Example: If M6 has extreme wave, what's the probability M2 also has extreme wave? ",
    "High values indicate strong spatial dependence during storms."
  ))
}
```

#### Copula Analysis

**Tail dependence: How correlated are extreme events?**

Copulas model the dependence structure, especially in the tails (extremes).

```{r joint-copula}
#| echo: false
if (has_joint_data && !is.null(copula_summary)) {
  cop_df <- copula_summary |>
    mutate(
      `Station Pair` = pair,
      `Kendall's τ` = round(kendall_tau, 3),
      `Upper Tail Dep (λU)` = round(upper_tail_dep, 3),
      `Lower Tail Dep (λL)` = round(lower_tail_dep, 3),
      `Observations` = format(n_obs, big.mark = ",")
    ) |>
    select(`Station Pair`, `Kendall's τ`, `Upper Tail Dep (λU)`,
           `Lower Tail Dep (λL)`, `Observations`)

  create_dt(cop_df, paste0(
    "Gumbel copula fits for station pairs. ",
    "Kendall's τ = rank correlation (higher = more similar). ",
    "Upper Tail Dep (λU) = probability of joint extremes in the upper tail (high waves). ",
    "λU > 0 means extreme waves tend to occur together. ",
    "Gumbel copula is appropriate for upper tail dependence."
  ))
}
```

**What does tail dependence mean?**

- **λU = 0**: Extremes are independent (one station's extreme doesn't predict the other's)
- **λU > 0**: Extremes are asymptotically dependent (joint extremes more likely than by chance)
- **λU → 1**: Perfect dependence in the tail (extremes always occur together)

For ocean buoys, λU > 0.3 indicates that storm events affect multiple stations simultaneously.

#### Theory

**Why analyze joint distributions?**

1. **Forecasting:** If M6 (offshore) shows large waves, can we predict when they'll reach coastal stations?

2. **Risk assessment:** During storms, are all stations simultaneously at risk, or do waves propagate sequentially?

3. **Rogue wave correlation:** Do rogue wave events at one station predict increased risk at others?

**Wave propagation model:**

Waves travel at the **group velocity**, which depends on wave period:

$$v_g = \frac{gT}{4\pi} \approx 1.56T \text{ m/s}$$

For typical swell with period T = 10s: $v_g \approx 15.6$ m/s $\approx 56$ km/h

For longer period swell (T = 15s): $v_g \approx 23$ m/s $\approx 84$ km/h

**Expected lag from M6:**

| Target | Distance (km) | Lag (T=10s) | Lag (T=15s) |
|--------|--------------|-------------|-------------|
| M5 | 200 | 3.6 h | 2.4 h |
| M2 | 420 | 7.5 h | 5.0 h |
| M3 | 380 | 6.8 h | 4.5 h |
| M4 | 580 | 10.4 h | 6.9 h |

*Actual lags vary with wave period, direction, and bathymetry.*

# Predictions {#predictions}

Random Forest model predicting **significant wave height** from meteorological variables and lagged features. Model uses [`ranger`](https://imbs-hl.github.io/ranger/){target="_blank"} with 500 trees trained on the first 70% of data (time-ordered split); the remaining 30% is held out as a **test set** for the metrics below. The time-ordered split ensures no future data leaks into training. See [methodology](#methodology) for measurement details and the [Model Summary](#model-summary) tab for the fitted formula and [`ranger`](https://imbs-hl.github.io/ranger/){target="_blank"} output.

```{r predictions-setup}
#| include: false
# Load RF targets separately (may not exist if pipeline hasn't run)
wave_rf_model <- wave_rf_eval <- wave_rf_report <- NULL
tryCatch({
  targets::tar_load(c(wave_rf_model, wave_rf_eval, wave_rf_report),
                    store = "../_targets")
}, error = function(e) {
  cli::cli_alert_warning("RF targets not available: {e$message}")
})

has_rf <- !is.null(wave_rf_model) && !is.null(wave_rf_eval)

if (has_rf) {
  rf_metrics <- wave_rf_eval$overall
  rf_r2 <- rf_metrics$value[rf_metrics$metric == "R_squared"]
  rf_rmse <- rf_metrics$value[rf_metrics$metric == "RMSE"]
  rf_mae <- rf_metrics$value[rf_metrics$metric == "MAE"]
  rf_n_test <- rf_metrics$value[rf_metrics$metric == "N_test"]
}
```

## Row

### Column

```{r rf-valuebox-r2}
#| content: valuebox
#| title: "Test R-squared"
list(
  value = if (has_rf) round(rf_r2, 3) else "N/A",
  color = if (has_rf && rf_r2 > 0.9) "success" else if (has_rf && rf_r2 > 0.7) "primary" else "warning"
)
```

### Column

```{r rf-valuebox-rmse}
#| content: valuebox
#| title: "Test RMSE (m)"
list(
  value = if (has_rf) paste0(round(rf_rmse, 3), " m") else "N/A",
  color = if (has_rf && rf_rmse < 0.5) "success" else if (has_rf && rf_rmse < 1.0) "primary" else "warning"
)
```

### Column

```{r rf-valuebox-mae}
#| content: valuebox
#| title: "Test MAE (m)"
list(
  value = if (has_rf) paste0(round(rf_mae, 3), " m") else "N/A",
  color = "info"
)
```

### Column

```{r rf-valuebox-ntest}
#| content: valuebox
#| title: "Test Samples"
list(
  value = if (has_rf) fmt_num(rf_n_test) else "N/A",
  color = "secondary"
)
```

## Row

### Column {.tabset}

#### Actual vs Predicted

```{r rf-actual-vs-predicted}
#| echo: false
if (has_rf) {
  preds <- wave_rf_eval$predictions
  # Sample if too many points for plotly
  if (nrow(preds) > 5000) {
    set.seed(42)
    preds_plot <- preds[sample(nrow(preds), 5000), ]
  } else {
    preds_plot <- preds
  }

  # LOESS fit line for actual model performance
  preds_sorted <- preds_plot[order(preds_plot$actual), ]
  loess_fit <- loess(predicted ~ actual, data = preds_sorted, span = 0.3)
  preds_sorted$fitted <- predict(loess_fit)

  plot_ly(preds_plot, x = ~actual, y = ~predicted, type = "scatter", mode = "markers",
          color = ~category,
          colors = c(Low = "#2196F3", Moderate = "#4CAF50", High = "#FF9800", Extreme = "#F44336"),
          marker = list(size = 6, opacity = 0.7),
          hoverinfo = "text",
          text = ~paste0("Actual: ", round(actual, 2), " m",
                         "<br>Predicted: ", round(predicted, 2), " m",
                         "<br>Residual: ", round(residual, 2), " m",
                         "<br>Category: ", category)) |>
    add_trace(x = c(0, max(preds$actual, na.rm = TRUE)),
              y = c(0, max(preds$actual, na.rm = TRUE)),
              type = "scatter", mode = "lines",
              line = list(dash = "dash", color = "#888888", width = 1),
              name = "y = x (perfect)", showlegend = TRUE,
              hoverinfo = "skip", inherit = FALSE) |>
    add_trace(data = preds_sorted, x = ~actual, y = ~fitted,
              type = "scatter", mode = "lines",
              line = list(color = "#00ff88", width = 2),
              name = "LOESS fit", showlegend = TRUE,
              hoverinfo = "skip", inherit = FALSE) |>
    irishbuoys_layout(
      title = paste0(
        "Actual vs Predicted Signif Wave on test set (n=",
        fmt_num(nrow(preds)), ", R\u00b2=", round(rf_r2, 3), ")<br>",
        "<sup>Dashed line = y=x (perfect prediction, not a fit). ",
        "Points coloured by 2m wave-height bins (Low/Moderate/High/Extreme) to show where the model struggles. ",
        "Click a legend entry to hide that category; the y=x line remains as reference.</sup>"
      ),
      xaxis = list(title = "Actual Signif Wave (m)"),
      yaxis = list(title = "Predicted Signif Wave (m)"),
      hovermode = "closest"
    )
} else {
  plotly_empty() |> layout(title = "RF model not available. Run targets::tar_make()")
}
```

#### Residuals by Category

```{r rf-residuals-by-category}
#| echo: false
if (has_rf) {
  preds <- wave_rf_eval$predictions
  plot_ly(preds, y = ~residual, x = ~category, type = "box",
          color = ~category,
          boxpoints = "outliers",
          hoverinfo = "text",
          text = ~paste0("Category: ", category, "<br>Residual: ", round(residual, 2), " m")) |>
    irishbuoys_layout(
      title = paste0(
        "Residuals by 2m Wave Height Category (test set)<br>",
        "<sup>Categories: Low (0-2m), Moderate (2-4m), High (4-6m), Extreme (>6m). ",
        "2m bins separate calm seas from storm conditions. ",
        "Residual = Actual - Predicted; zero = perfect. Wider boxes = more error.</sup>"
      ),
      xaxis = list(title = "Wave Height Category"),
      yaxis = list(title = "Residual (Actual - Predicted, m)"),
      showlegend = FALSE,
      shapes = list(
        list(type = "line", x0 = -0.5, x1 = 3.5, y0 = 0, y1 = 0,
             line = list(dash = "dash", color = "grey", width = 1))
      )
    )
} else {
  plotly_empty() |> layout(title = "RF model not available")
}
```

## Row

### Column {.tabset}

#### Feature Importance

```{r rf-feature-importance}
#| echo: false
if (has_rf) {
  imp <- head(wave_rf_model$importance, 10)
  imp <- imp[order(imp$importance), ]  # ascending for horizontal bar
  imp$variable <- factor(imp$variable, levels = imp$variable)

  plot_ly(imp, y = ~variable, x = ~importance, type = "bar",
          orientation = "h",
          marker = list(color = "#1f77b4")) |>
    irishbuoys_layout(
      title = paste0(
        "Top 10 Feature Importance (Impurity-based, <a href='https://imbs-hl.github.io/ranger/'>ranger</a>)<br>",
        "<sup>Higher = variable contributes more to reducing prediction error. ",
        "Lagged wave heights dominate because wave conditions persist hour-to-hour.</sup>"
      ),
      xaxis = list(title = "Importance"),
      yaxis = list(title = ""),
      margin = list(l = 150, b = 80)
    )
} else {
  plotly_empty() |> layout(title = "RF model not available")
}
```

#### Performance by Category

```{r rf-category-table}
#| echo: false
if (has_rf) {
  cat_df <- wave_rf_eval$by_category
  cat_df$rmse <- round(cat_df$rmse, 3)
  cat_df$mae <- round(cat_df$mae, 3)
  create_dt(cat_df, paste0(
    "Model performance by 2m wave height category (test set). ",
    "Categories: Low (0-2m calm seas), Moderate (2-4m), High (4-6m storm), Extreme (>6m). ",
    "RMSE = root mean squared error; MAE = mean absolute error (both in metres). ",
    "Higher errors for Extreme waves indicate the model is less accurate for rare storm events."
  ))
} else {
  knitr::asis_output("RF model not available. Run targets::tar_make()")
}
```

#### Model Summary {#model-summary}

```{r rf-model-summary}
#| echo: false
#| results: asis
if (has_rf) {
  # Formula
  cat("**Fitted formula:**\n\n")
  cat("```\n")
  cat(deparse(wave_rf_model$formula, width.cutoff = 120), sep = "\n")
  cat("\n```\n\n")

  # Model parameters table
  cat("**[`ranger`](https://imbs-hl.github.io/ranger/) model parameters:**\n\n")
  model_params <- data.frame(
    Parameter = c("Number of trees", "Predictors used", "Training observations",
                  "Test observations", "Training R\u00b2", "OOB RMSE (m)",
                  "Test R\u00b2", "Test RMSE (m)", "Test MAE (m)", "Test Bias (m)"),
    Value = c(
      "500",
      as.character(length(wave_rf_model$predictors)),
      format(length(wave_rf_model$train_idx), big.mark = ","),
      format(length(wave_rf_model$test_idx), big.mark = ","),
      round(wave_rf_model$train_r_squared, 4),
      round(wave_rf_model$oob_rmse, 4),
      round(rf_r2, 4),
      round(rf_rmse, 4),
      round(rf_mae, 4),
      round(wave_rf_eval$overall$value[wave_rf_eval$overall$metric == "Bias"], 4)
    ),
    stringsAsFactors = FALSE
  )
  knitr::kable(model_params, format = "pipe", align = c("l", "r"),
               caption = paste0(
                 "ranger Random Forest model summary. ",
                 "OOB = out-of-bag error (internal cross-validation within training set). ",
                 "Test metrics use the held-out 30% of data (time-ordered, no leakage)."
               ))
} else {
  cat("[`ranger`](https://imbs-hl.github.io/ranger/) model not available. Run `targets::tar_make()`.\n")
}
```

```{r rf-predictors-table}
#| echo: false
if (has_rf) {
  imp_df <- wave_rf_model$importance
  imp_df$importance <- round(imp_df$importance, 1)
  imp_df$rank <- seq_len(nrow(imp_df))
  imp_df <- imp_df[, c("rank", "variable", "importance")]
  create_dt(imp_df, paste0(
    "All predictor variables ranked by impurity-based importance (ranger). ",
    "Higher importance = larger reduction in prediction error when splitting on this variable. ",
    "Lagged wave heights (wave_height_lag1-3) are strongest because sea state is autocorrelated."
  ))
} else {
  knitr::asis_output("RF model not available. Run targets::tar_make()")
}
```

#### Model Report

```{r rf-report}
#| echo: false
#| results: asis
if (has_rf) {
  cat("```\n", wave_rf_report, "\n```", sep = "")
} else {
  cat("[`ranger`](https://imbs-hl.github.io/ranger/) model not available. Run `targets::tar_make()`.\n")
}
```

# Reference

## Technical Reference

### Reference Views {.tabset}

#### Definitions

##### Definition Categories {.tabset}

###### Key Terms

**Table: Key Oceanographic and Meteorological Terms**

| Term | Definition | Reference |
|------|------------|-----------|
| **Signif Wave = H1/3 = 4σ** | All three mean the same: significant wave height = mean of highest 1/3 of waves = 4 × standard deviation of sea surface. See [worked example](#how-signif-wave-is-calculated). | [Wikipedia](https://en.wikipedia.org/wiki/Significant_wave_height){target="_blank"}, [Longuet-Higgins 1952](https://elischolar.library.yale.edu/journal_of_marine_research/774/){target="_blank"} |
| **Max Wave** | Maximum individual wave height in 17.5-min measurement period | [NOAA wave background](https://www.weather.gov/key/marine_sigwave){target="_blank"}. Typically Max Wave / Signif Wave ~ 1.5-1.9 |
| **Rogue Wave** | Wave where **Max Wave / Signif Wave > 2.0** | [Wikipedia](https://en.wikipedia.org/wiki/Rogue_wave){target="_blank"}, [Dysthe 2008](https://doi.org/10.1146/annurev.fluid.40.111406.102203){target="_blank"} |
| **Rogue Ratio** | = Max Wave / Signif Wave. Values > 2.0 indicate a rogue wave | Used in [Rogue Events](#rogue-events) |
| **Gust Ratio** | = Peak Gust / Sustained Wind. Typical value ~1.3 | Used in [Rogue Gusts](#rogue-gusts) |
| **Rogue Gust** | Observation where **Gust / Wind > 1.5** | Used in [Rogue Gusts](#rogue-gusts) |
| **Sustained Wind** | 10-minute mean wind speed at ~3m height. Denominator in [gust factor](https://glossary.ametsoc.org/wiki/Gust_factor){target="_blank"}. | [WMO Guide](https://library.wmo.int/idurl/4/41650){target="_blank"} |
| **GEV** | Generalized Extreme Value distribution for annual maxima | [Wikipedia](https://en.wikipedia.org/wiki/Generalized_extreme_value_distribution){target="_blank"}. Used in [Hourly Max](#hourly-max) |
| **GPD** | Generalized Pareto Distribution for peaks over threshold | [Wikipedia](https://en.wikipedia.org/wiki/Generalized_Pareto_distribution){target="_blank"}. Used in [Hourly Max](#hourly-max) |
| **Return Level** | Value expected to be exceeded once per T years | Used in [Hourly Max](#hourly-max) |
| **Gust Factor** | Peak Gust / Sustained Wind (~1.3 typical) | [AMS Glossary](https://glossary.ametsoc.org/wiki/Gust_factor){target="_blank"}. Used in [Gust Factor](#gust-factor) |
| **STL** | Seasonal-Trend decomposition (Loess) | [Wikipedia](https://en.wikipedia.org/wiki/Decomposition_of_time_series){target="_blank"} |

*Terminology used throughout this dashboard. Rogue events are identified when ratios exceed defined thresholds.*

###### Rogue Events

**Table: Rogue Event Thresholds**

| Event Type | Ratio Formula | Threshold | Interpretation |
|------------|---------------|-----------|----------------|
| **Rogue Wave** | Max Wave / Signif Wave | > 2.0 | Individual wave significantly exceeds average sea state |
| **Rogue Gust** | Gust / Wind Speed | > 1.5 | Peak gust significantly exceeds sustained wind |

**Normal ranges:**

- Wave Ratio: 1.5–1.9 (normal sea state)
- Gust Ratio: 1.2–1.4 (typical conditions)

*Caption: Rogue events are rare extremes. Wave ratios > 2.0 occur in ~1-2% of observations; gust ratios > 1.5 occur in ~3-5%.*

###### Beaufort Scale

**Table: Beaufort Wind Scale**

| Beaufort | Description | Wind Speed (m/s) | Sea State |
|----------|-------------|------------------|-----------|
| 0-5 | Calm to Fresh Breeze | 0 – 10.7 | Ripples to moderate waves |
| 6-7 | Strong Breeze to Near Gale | 10.8 – 17.1 | Large waves, whitecaps |
| 8-9 | Gale to Severe Gale | 17.2 – 24.4 | High waves, spray |
| 10 | Storm | 24.5 – 28.4 | Very high waves |
| 11 | Violent Storm | 28.5 – 32.6 | Exceptionally high waves |
| 12 | Hurricane Force | > 32.7 | Air filled with foam |

*Caption: Beaufort scale converts wind speed to descriptive categories. Used in [Gust Factor](#gust-factor) analysis.*

#### Methodology

##### Methodology Details {.tabset}

###### Signif Wave Calculation

**Why Signif Wave = H1/3 = 4σ**

These three notations all mean the same thing:

- **Signif Wave** = Significant wave height (common abbreviation)
- **H1/3** = Mean of highest 1/3 of waves (original definition)
- **4σ** = 4 × standard deviation of sea surface elevation

The equivalence arises from the [Rayleigh distribution](https://en.wikipedia.org/wiki/Rayleigh_distribution){target="_blank"} of wave heights. For a narrow-banded wave spectrum, the theoretical relationship is **Signif Wave = 4 × sqrt(m0)** where m0 is the variance of sea elevation. See [Wikipedia: Significant wave height](https://en.wikipedia.org/wiki/Significant_wave_height){target="_blank"} and [NOAA wave background](https://www.weather.gov/key/marine_sigwave){target="_blank"} for further detail.

See [How Signif Wave is Calculated](#how-signif-wave-is-calculated) for a worked example, and the [Methodology](#methodology) tab for measurement details.

###### Measurement Window

**Why 17.5-minute measurement?**

| Requirement | Specification | Rationale |
|-------------|---------------|-----------|
| **Sample size** | ~100+ waves | Statistical validity for H1/3 |
| **Wave periods** | 5–15 seconds | 17.5 min yields ~70–200 waves |
| **WMO standard** | 20–30 minutes | 17.5 is practical lower bound |
| **Trade-off** | Resolution vs. statistics | Shorter = more temporal detail |

*Caption: Measurement window balances statistical requirements with temporal resolution.*

**Data source:** [Marine Institute ERDDAP](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html){target="_blank"}

###### Extreme Value Methods

**Table: Statistical Methods for Extreme Value Analysis**

| Method | Full Name | Input Data | Used In |
|--------|-----------|------------|---------|
| **GEV** | Generalized Extreme Value | Block maxima (hourly max) | [Hourly Max](#hourly-max) |
| **GPD** | Generalized Pareto Distribution | Peaks over threshold | [Hourly Max](#hourly-max) |
| **Return Level** | N/A | GEV/GPD fitted parameters | [Hourly Max](#hourly-max) |

*Caption: GEV fits annual/block maxima; GPD fits all exceedances above a threshold. Both estimate return levels (values exceeded once per T years).*

#### References

##### Reference Sources {.tabset}

###### Academic

**Key Papers:**

| Author | Year | Title | DOI |
|--------|------|-------|-----|
| Longuet-Higgins, M.S. | 1952 | Statistical distribution of sea waves | [Yale archive](https://elischolar.library.yale.edu/journal_of_marine_research/774/){target="_blank"} |
| Dysthe, K. et al. | 2008 | Oceanic rogue waves | [10.1146/annurev.fluid.40.111406.102203](https://doi.org/10.1146/annurev.fluid.40.111406.102203){target="_blank"} |
| Coles, S. | 2001 | *Statistical Modeling of Extreme Values* (Springer) | — |

*Caption: Foundational papers for wave statistics and extreme value theory.*

###### R Packages

**Table: R Packages Used in Analysis**

| Package | Purpose | Link |
|---------|---------|------|
| [`targets`](https://docs.ropensci.org/targets/){target="_blank"} | Pipeline orchestration | [CRAN](https://cran.r-project.org/package=targets){target="_blank"} |
| [`ranger`](https://imbs-hl.github.io/ranger/){target="_blank"} | Fast Random Forest implementation | [CRAN](https://cran.r-project.org/package=ranger){target="_blank"} |
| [`mev`](https://lbelzile.github.io/mev/){target="_blank"} | Extreme value analysis | [CRAN](https://cran.r-project.org/package=mev){target="_blank"} |
| [`extRemes`](https://www.ral.ucar.edu/staff/ericg/extRemes/){target="_blank"} | GEV/GPD fitting | [CRAN](https://cran.r-project.org/package=extRemes){target="_blank"} |
| [`plotly`](https://plotly.com/r/){target="_blank"} | Interactive visualisation | [CRAN](https://cran.r-project.org/package=plotly){target="_blank"} |
| [`DT`](https://rstudio.github.io/DT/){target="_blank"} | Interactive tables | [CRAN](https://cran.r-project.org/package=DT){target="_blank"} |
| [`dygraphs`](https://rstudio.github.io/dygraphs/){target="_blank"} | Time series widgets | [CRAN](https://cran.r-project.org/package=dygraphs){target="_blank"} |
| [`irishbuoys`](https://johngavin.github.io/irishbuoys/){target="_blank"} | This package | [GitHub](https://github.com/JohnGavin/irishbuoys){target="_blank"} |

*Core R packages enabling reproducible analysis and interactive dashboards.*

###### Data Sources

**Primary Data:**

| Source | Description | URL |
|--------|-------------|-----|
| Marine Institute Ireland | Buoy network operator | [marine.ie](https://www.marine.ie/){target="_blank"} |
| ERDDAP Data Portal | Raw buoy data access | [erddap.marine.ie](https://erddap.marine.ie/erddap/tabledap/IWBNetwork.html){target="_blank"} |
| ERDDAP Metadata | Dataset documentation | [info/IWBNetwork](https://erddap.marine.ie/erddap/info/IWBNetwork/index.html){target="_blank"} |

*Caption: All buoy data sourced from [Marine Institute Ireland](https://www.marine.ie/){target="_blank"} via ERDDAP protocol.*

---

*Analysis generated `r generated_time` using irishbuoys v`r pkg_version`*

---

::: {.callout-note appearance="minimal"}
**Disclaimer:** `irishbuoys` is an independent open-source project ([website](https://johngavin.github.io/irishbuoys/){target="_blank"}, [GitHub](https://github.com/JohnGavin/irishbuoys){target="_blank"}). It is not affiliated with, endorsed by, or part of the [Marine Institute](https://www.marine.ie/){target="_blank"} or the Irish Weather Buoy Network. For official marine weather warnings, consult [Met Eireann](https://www.met.ie/warnings){target="_blank"}.
:::
