---
title: "Spatial Extreme Value Analysis"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

library(targets)
library(dplyr)
library(DT)

# Set targets store - works from project root or vignettes/
store_path <- if (dir.exists("_targets")) "_targets" else "../_targets"
targets::tar_config_set(store = store_path)

# Load pre-computed results (ZERO computation in vignette)
targets::tar_load(c(
  spatial_evt_results,
  spatial_dependence_table,
  spatial_rogue_h3_table
))
```

## Overview

This vignette presents the spatial extreme value analysis of significant wave
heights across the Irish Marine Buoy Network. Three hypotheses are tested:

- **H1 (Spatial Coherence):** Do extreme wave events show significant spatial
  dependence between stations?
- **H3 (Rogue Wave Propagation):** Do rogue wave events at one station cluster
  temporally with events at another, consistent with wave propagation?
- **Max-Stable Process:** Can a Brown-Resnick max-stable model capture the
  joint spatial extremal structure? (Illustrative only — 5 stations is below
  the recommended minimum of ~20.)

## H1: Pairwise Extremal Dependence

The upper tail dependence coefficient $\lambda_U$ is estimated for all station
pairs via Gumbel copula fitting. For a Gumbel copula with parameter $\alpha$:

$$\lambda_U = 2 - 2^{1/\alpha}$$

Bootstrap confidence intervals (100 replicates) determine whether $\lambda_U$
is significantly greater than zero.

```{r h1-table}
spatial_dependence_table |>
  mutate(
    across(c(kendall_tau, lambda_upper, lambda_upper_ci_low, lambda_upper_ci_high),
           \(x) round(x, 3)),
    distance_km = round(distance_km, 0),
    h1_verdict = ifelse(h1_significant, "Significant", "Not significant")
  ) |>
  select(
    `Station 1` = station1,
    `Station 2` = station2,
    `Distance (km)` = distance_km,
    `Kendall tau` = kendall_tau,
    `Lambda U` = lambda_upper,
    `CI Low` = lambda_upper_ci_low,
    `CI High` = lambda_upper_ci_high,
    `N obs` = n_concurrent,
    Verdict = h1_verdict
  ) |>
  DT::datatable(
    caption = "Pairwise upper tail dependence (Gumbel copula, 95% bootstrap CI)",
    options = list(pageLength = 15, dom = "t"),
    rownames = FALSE
  )
```

**Key finding:** `r spatial_evt_results$n_h1_significant` of
`r spatial_evt_results$n_pairs` station pairs show significant positive
extremal dependence ($\lambda_U$ CI > 0).

## H3: Rogue Wave Spatial Propagation

Rogue waves are defined as events where $H_{max} > 2.0 \times H_s$ with
$H_s \geq 2.0$ m. For each directed station pair, a permutation test
(500 replicates) assesses whether rogue events at the source station are
followed by rogue events at the receiver station within a propagation
time window (distance / 30 km/h deep-water group velocity).

```{r h3-table}
if (nrow(spatial_rogue_h3_table) > 0) {
  spatial_rogue_h3_table |>
    mutate(
      across(c(co_occurrence_rate, marginal_rate, perm_mean_rate, p_value),
             \(x) round(x, 3)),
      distance_km = round(distance_km, 0),
      theoretical_lag_hrs = round(theoretical_lag_hrs, 1)
    ) |>
    select(
      Source = station1,
      Receiver = station2,
      `Distance (km)` = distance_km,
      `Lag (hrs)` = theoretical_lag_hrs,
      `N rogues (src)` = n_rogue_s1,
      `N rogues (rcv)` = n_rogue_s2,
      `Co-occur rate` = co_occurrence_rate,
      `Marginal rate` = marginal_rate,
      `p-value` = p_value,
      Verdict = h3_verdict
    ) |>
    DT::datatable(
      caption = "Rogue wave spatial clustering permutation test",
      options = list(pageLength = 15, dom = "t"),
      rownames = FALSE
    )
} else {
  cat("No rogue wave events detected — H3 test not applicable.")
}
```

**Result:** `r spatial_evt_results$n_h3_significant` pairs show significant
spatial clustering of rogue wave events (p < 0.05).

## Max-Stable Process

```{r maxstable}
ms <- spatial_evt_results$maxstable
if (ms$fitted) {
  cat(sprintf(
    "A %s max-stable model was fitted successfully.\nParameters: %s\n",
    ms$model_type,
    paste(names(ms$parameters), round(ms$parameters, 3), sep = " = ", collapse = ", ")
  ))
} else {
  cat("Max-stable model fitting failed or was not attempted.\n")
  if (!is.null(ms$reason)) cat("Reason:", ms$reason, "\n")
}
cat("\nLimitation:", ms$limitation)
```
